```{=html}
<!--
Chapter 1 - Introduction
-->
```
```{r setup1, echo=FALSE}
pacman::p_load(tidyverse, thesisdown, knitr, kableExtra)

# Set how wide the R output will go
options(width = 70)
```

# Introduction {#introduction}

## Background {#background}

Energy demand of residential buildings represents one of the largest energy demand points in Germany and worldwide. In Germany specifically, building energy demand of private households accounted for 20.2% of the overall primary energy demand in 2020 [@ageb21]. While households consume energy in their homes for various purposes, 68.1% of the overall residential energy demand in 2020 was used for the generation of space heating [@rwi21]. Space heating therefore represents the by far largest type of energy end-use in residential buildings in Germany, far exceeding other types of end-uses such as hot water generation (16.1%), process heat (6.0%), room cooling, information technology, or lighting (all below 5%) [@rwi21].

From a greenhouse gas (GHG) emissions perspective, it is also the two end-uses of space heating and hot water generation that are the predominant contributors for the occurrence of GHG emissions in the residential buildings sector, as they continue to be predominantly produced from fossil fuels. Again for 2020, 80.1% of the energy used to generate space heating in private households in Germany came from either natural gas (45.4%), oil (24.5%), or district heating (10.1%)[^1] [@rwi21]. Thus, generation of space heating is not only the largest point of energy demand in residential buildings, but also the largest contributor of GHG emissions due to its generation from fossil sources.

Over the past decades, there have been multiple trends in the residential buildings sector that effected energy demand and associated GHG emissions. Looking at a longer period between 1995 and 2020, greenhouse gas emissions in the entire building sector (including commercially used buildings) in Germany have fallen by 34% [@erk20]. When, however, only considering a shorter historical period, it becomes apparent that much of the energy efficiency gains in the buildings sector were already realized in the 1990s and 2000s. For the last decade, on the contrary, several independent sources of data indicate that primary energy demand for residential space heating per square meter in Germany has largely stagnated if effective energy demands are corrected for spatial and temporal variations in climatic conditions [@bmwi21; @stede_etal20; @techem19]. Furthermore, from a sectoral perspective, energy efficiency gains (better insulation of building stock, more efficient heating technologies) and a lower emission intensity from fuel switches are being partly offset by countervailing effects including an increase in the number of apartments (smaller average household size) as well as the increase in average living space [@bmwi21].

[^1]: Which is at present also to almost 90% generated from fossil fuels (natural gas or coal) or waste incineration [@destatis18].

In the face of climate change, there is a urgent need for the decarbonisation of the building sector. In response to the Paris Agreement, which aims to limit global warming to well-below 2 degrees Celsius by 2100, Germany introduced a national Federal Climate Change Act (Bundes-Klimaschutzgesetz, KSG) to govern the transition. The latest 2021 amendment of the KSG stipulates that national GHG emissions must be reduced by 65% until 2030 (compared to the base year 1990) and net-zero emissions must be reached in 2045. Congruently, at the level of the European Union (EU), the Green Deal defines 2050 as the target year to reach net-zero in the whole union [@europeancomission19]. One of the relevant changes in the KSG amendment of 2021 is the introduction of annual sector targets as a concretisation of national emission reduction pathways. One of the six sectors defined in the KSG is the buildings sector, whose emissions are to be reduced from 118 million tonnes of CO2e to 67 million tonnes of CO2e in the period between 2020 and 2030 (-43.2%).[^2] For the residential buildings sector, achieving these ambitious reduction targets requires a comprehensive transformation of energy use, with measures to improve the efficiency of the building stock, demand-side electrification of the remaining energy use, as well as supply-side decarbonisation of energy generation [@hennes_etal21; @levesque_etal21].

[^2]: Federal Climate Change Act of 12 December 2019 (Federal Law Gazette I, p. 2513), as last amended by Article 1 of the Act of 18 August 2021 (Federal Law Gazette I, p. 3905).

## Objective and relevance {#relevance}

Economic theory relies on the rationale that prices represent the key driver for the level of demand for a good. Therefore, carbon pricing instruments are one of the main instruments to drive and support the decarbonisation of the buildings sector. Yet, modelling of the effectiveness of any price-related energy policy instrument strongly relies on the assumed responsiveness of the demand reactions of consumers. Against this background, this thesis pursues the objective of using historical observational data to conduct a statistical analysis of determinants of space heating demand in Germany as the main end-use of energy in buildings and contributor of residential GHG emissions. More specifically, the thesis focuses on estimating short-term own-price elasticities of space heating demand to supplement the existing literature on the responsiveness of the demand response for space heating. The empirical basis for the analysis is a sample of energy bills of multi-apartment residential buildings in Germany in the period between 2007 and 2019.

**Evaluation of price-related energy policies**

Estimating price elasticities of space heating demand in the residential buildings sector is relevant for a number of reasons. First, empirical evidence on price elasticities is generally important for evaluating policies that affect the price of energy. For any such price-related policy one needs to understand the relevant determinants which affect the demand for energy. Based on elasticity estimates, the economic, distributional and environmental impacts of policy measures can be better understood and estimated [@alberini_etal11; @labandeira_etal17; @zhu_etal18]. This is especially applicable with regard to instruments that aim to contribute to the attainment of ambitious climate targets.

One of the most important measures to induce GHG emission reductions in the building sector in Germany is the introduction of a carbon price for heating fuels, which was rolled out in 2021 on the basis of the Fuel Emissions Trading Act (BEHG).[^4] In general, the introduction of a carbon pricing aims to correct the market failure of the previously unrecognized external effects of greenhouse gas emissions [@aldy_etal10; @stiglitz19]. It is expected that the consideration of external effects will have a steering effect on demand through an increased price level. The recently introduced Fuel Emissions Trading Act specifies a fixed price path for the period until 2025. The path rises from an initial price of 25 euros per ton of carbon dioxide (CO~2~) in 2021 to 55 euros per ton of CO~2~ in 2025.

[^4]: Fuel Emissions Allowance Trading Act (BEHG) of 12 December 2019 (Federal Law Gazette I p. 2728) as amended by Article 1 of the Act of 3 November 2020 (Federal Law Gazette I p. 2291). Emissions from the building sector have not been subject to carbon pricing in Germany in the past, as they do not fall within the scope of the EU Emissions Trading Scheme (ETS). Besides emissions in the building sector, the BEHG also covers emissions from the transport sector. From 2026 onward the fixed pricing will be superseded by an emission trading system (ETS) with a price corridor.

Figure \@ref(fig:behg) exemplifies the effect of carbon prices for the buildings sector through the BEHG. For gas (Panel A) and oil (Panel B) used for residential space heating, the respective left bar (green) represents the average end consumer prices for the five-year period 2016-2020 from national statistics [@bmwi21].[^5] The two middle bars (yellow) show the absolute price effects that occur based on the carbon intensity of the energy carrier in 2021 (25 euro per ton CO~2~) and in 2025 (55 euro per ton CO~2~). The yellow bars are larger for oil due to its higher emission intensity.[^6] Assuming otherwise constant prices, the initial CO~2~ price in 2021 would thus lead to a price increase of +0.50 cents (+7.4%) per kWh gas and +0.66 cents (+13.4%) per kWh oil. In 2025, the price effects rise to +1.10 cents (+16.3%) per kWh gas and +1.46 cents (+29.6%) per kWh oil.

[^5]: The 5-year time period (2016-2020) was chosen due to the relatively strong volatility of oil prices in the past years. An exclusive consideration of the price for 2020 would therefore have led to a distortion of the relative price effects. In addition, district heating is not considered here, as most district heating power plants were already subject to carbon pricing under the EU ETS. Only plants with a total combustion capacity of less than 20 megawatts were not previously subject to the EU ETS and will now also be subject to BEHG pricing.

[^6]: Emission factors for gas (0.201 kg CO~2~e per kWh) and oil (0.266 kg CO~2~e per kWh) were taken from [@uba16].

```{r behg, echo = FALSE, out.width = "100%", fig.align = "center", fig.cap = "BEHG carbon price effects on gas and oil until 2025", fig.scap="BEHG carbon price effects on gas and oil until 2025"}
knitr::include_graphics(path = "figure/price_effect_behg.pdf")
```

These price figures clearly indicate that relevant steering effects induced by the BEHG are likely to occur already in the next years. Higher price levels driven by the structural price component of carbon pricing can be expected to have two impact channels. The first impact channel are short-term price reactions. Consumers adjust their consumption levels for heating fuels due to higher or lower prices [@alberini_etal11]. Short-term demand responses are therefore of direct relevance to forecasting building sector energy demand and emission levels in the coming years. The expected effect size of these short-term reactions are reflected by short-term price elasticities of heating demand which will be estimated based on empirical data in this thesis.

The second impact channel, on the other hand, is that higher prices for fossil fuels can be expected to induce long-term steering effects for investments in low-carbon heating technologies and energy efficiency. In the literature, long-term investment decisions driven by mitigation costs potentials are modeled using integrated assessment models (IAMs) for the buildings sector [e.g., @burger_etal19; @levesque_etal21]. Depending on the model structure, these IAM sector modules are often informed by assumptions on the price elasticities of heating demand as one of their input factor besides other parameters such as the cost of capital or substitution elasticities.[^7] Therefore, empirical estimates of price elasticities are also important for further analyses in the literature on the second long-term impact channel.

[^7]: For a more detailed example of how own price elasticities of demand are taken into account in IAMs, see the model structure description of the ETSAP TIAM model in @loulou_labriet08. Since IAMs vary in their specific structure and relations other models may consider price elasticities differently.

Additionally, the empirical assessment of the demand effect may be particularly important in the current phase of strong energy price increases following from the economic imbalances after the external shock of the COVID-19 pandemic and Russia's war of aggression on Ukraine. While the current price spikes can lead to stronger short-term demand responses, only a fraction of the response is due to the structural component of CO~2~ pricing and -- if the effect does not last long enough to trigger significant long-term investments -- would disappear when oil and gas prices might return to lower levels. In order not to overestimate the effect of the structural pricing component of the CO~2~ price, empirical estimates from past price reactions could be used to separate the effect into a CO~2~ and a commodity price-related element.

**Billing data as a complementary source of evidence**

A second line of argument for the relevance of this study is that the existing empirical evidence on the price elasticity of heating demand is based on different types of data sources, but few studies have been conducted on actual billing data.

Studies on the price elasticities of heating demand rely on different types of data sources. For the case of Germany, ex-post statistical analyses for the residential buildings sector use data from social panel surveys [e.g., @rehdanz07; @schmitz_madlener20; @schulte_heindl17]. In these studies, household expenditure on heating reported in the survey is usually examined as the relevant variable, but the actual level of energy consumption and energy prices are not observed. This means that, for instance, different energy contract conditions between households may obscure real consumption levels and thus affect the validity of the estimated results. The data available for this thesis is different since the energy bills do provides actual consumption and price data. Other studies using data from energy bills have previously been conducted in the United States (US) [e.g., @auffhammer_rubin18].

In this context it should be noted that the data sample used in this thesis also has its weaknesses. The data does not observe demand responses at the level of individual households but at the level of multi-household buildings. Yet despite this aggregation of multiple households on the building level, I believe the informative value of estimates to be complementary to estimates from other types of data sources such as social surveys.

**Research questions**

Against the background of the tightened emissions reduction targets, the recently introduced carbon pricing, and the sharp price increases for heating fuels, the question of the price elasticity of heating demand is once again moving to the center of public attention. The aim of this work is to investigate the price elasticity of heating demand in multi-family houses in Germany based on a large-scale sample of energy bills. The guiding research questions for the analysis are:

-   How does a change in energy price affect the level of demand for space heating?
-   What other variables might affect demand for space heating and need to be taken into account so that their effects are not falsely attributed to energy prices?
-   And, are there potential factors for heterogeneity in the sample that may be obscured when only estimating an overall price elasticity of demand?

By using energy bills as a data source, the study aims to complement existing studies based on different sources. Furthermore, the period studied is more recent and can therefore complement older studies. Also, the analysis aims not only to estimate an overall elasticity, but also to investigate potential factors for heterogeneity. The statistical analysis aims to use several different modelling approaches, both Frequentist and Bayesian. As far as the author is aware, the Bayesian approach has not been used in previous research in this field. Furthermore, the elasticity estimates from this work may also be relevant for other countries and regions that have comparable conditions to Germany in terms of building stock and energy demand and also face the challenge of decarbonising the building sector in the coming decades. Here, my results would serve as an empirical guide to the short-term demand effects for any kind of price-related policy instrument.

The thesis is structured in six Chapters. While the first Chapter was used to provide an introduction, Chapter \@ref(literature) gives an overview on the existing literature on price elasticities of space heating demand. In the following Chapter \@ref(methods) the data and methods used for the analysis are laid out. Chapter \@ref(results) provides the results of the analysis, which are then discussed in Chapter \@ref(discussion). Chapter \@ref(conclusion) provides concluding remarks.
