```{=html}
<!--
Chapter 1 - Introduction
-->
```
```{r setup1, echo=FALSE}
pacman::p_load(tidyverse, thesisdown, knitr, kableExtra)

# Set how wide the R output will go
options(width = 70)
```

# Introduction {#introduction}

## Background {#background}

The energy demand of residential buildings represents one of the largest energy demand points in Germany and worldwide. In Germany specifically, the building energy demand of private households accounted for 20.2% of the overall primary energy demand in 2020 [@ageb21]. While households consume energy in their homes for various purposes, 68.1% of the overall residential energy demand in 2020 was used to generate space heating [@rwi21]. Space heating, therefore, represents the most significant type of energy end-use in residential buildings in Germany. It far exceeds the energy demand for other types of end-uses such as hot water generation (16.1%), process heat (6.0%), room cooling, information technology, or lighting (all below 5%) [@rwi21]. Relating the generation of residential space heating to the overall primary energy demand in Germany, it represents 13.6% of the national primary energy demand.

Also, from an emissions perspective, the two end-uses of space heating and hot water generation are the predominant contributors to GHG emissions in the residential buildings sector. That is because they continue to be produced predominantly from fossil fuels. In 2020, 80.1% of the energy used to generate space heating in private households in Germany came from either natural gas (45.4%), oil (24.5%), or district heating (10.1%)[^1] [@rwi21]. Thus, the generation of space heating is not only the most significant energy demand in residential buildings but also the largest contributor to GHG emissions due to its generation from fossil sources.

[^1]: District heating is currently also generated to almost 90% from fossil fuels (natural gas or coal) or through waste incineration [@destatis18].

Over the past decades, multiple trends in the residential buildings sector have affected energy demand and associated GHG emissions. Looking at a more extended period between 1995 and 2020, greenhouse gas emissions in the entire building sector (including commercially used buildings) in Germany have fallen by 34% [@erk20]. When, however, only considering a shorter historical period, it becomes apparent that much of the energy efficiency gains in the buildings sector were already realised in the 1990s and 2000s. For the last decade, on the contrary, several independent data sources indicate that the primary energy demand for residential space heating per square meter in Germany has stagnated if effective energy demands are corrected for spatial and temporal variations in climatic conditions [@bmwi21; @stede_etal20; @techem19]. Furthermore, from a sectoral perspective, energy efficiency gains (better insulation of building stock, more efficient heating technologies) and a lower emission intensity from fuel switches are being partly offset by countervailing effects, including an increase in the number of apartments (smaller average household size) as well as the rise in per capita living space [@bmwi21].

In the face of climate change, there is an urgent need to decarbonise the buildings sector [@levesque_etal21]. In response to the Paris Agreement, which aims to limit global warming to well below 2°C by 2100, Germany introduced a national Federal Climate Change Act (Bundes-Klimaschutzgesetz, KSG) to govern the low-carbon transformation. The latest 2021 amendment of the KSG stipulates that national GHG emissions must be reduced by 65% by 2030 (compared to the base year 1990), and net-zero emissions must be reached in 2045. Congruently, the Green Deal of the European Union (EU) defines 2050 as the target year to achieve net-zero in the whole union [@europeancomission19]. One of the most relevant changes in the KSG amendment of 2021 is the introduction of annual sector targets as a concretisation of the national emission reduction target. One of the six sectors defined in the KSG is the buildings sector, whose emissions must be reduced from 118 million tonnes of carbon dioxide equivalents (CO~2~e) in 2020 to 67 million tonnes of CO~2~e in 2030 (-43.2%).[^2] To achieve these ambitious reduction targets, a comprehensive transformation of energy use in the buildings sector is required. This includes measures to improve the efficiency of the building stock, the electrification of the remaining energy use on the demand side and the decarbonisation of energy generation on the supply side [@hennes_etal21; @levesque_etal21].

[^2]: Federal Climate Change Act of 12 December 2019 (Federal Law Gazette I, p. 2513), as last amended by Article 1 of the Act of 18 August 2021 (Federal Law Gazette I, p. 3905).

## Objective and Relevance {#relevance}

Economic theory relies on the rationale that prices represent the key driver of the level of demand for a good [@pindyck_rubinfeld18; @stigler87]. Therefore, carbon pricing instruments are one of the key instruments to support and drive the decarbonisation of the buildings sector while being supplemented by additional policies [@braungardt_etal21]. Ultimately, however, modelling the effectiveness and impact of any price-related energy and climate policy instrument strongly depends on the assumed responsiveness of consumer demand [@alberini_etal11; @zhu_etal18].

Against this background, this thesis aims to use historical observational data on the space heating demand of private households in Germany to analyse the determinants of demand levels statistically. More specifically, the thesis focuses on estimating short-term own-price elasticities of space heating demand and thereby supplements the existing literature on the responsiveness of the demand response for space heating. The focus is on space heating, as it is the most relevant type of energy end-use in residential buildings and the primary driver of GHG emissions. The empirical basis for the analysis is a sample of energy bills of multi-apartment residential buildings in Germany between 2007 and 2019.

Russia's war of aggression against Ukraine launched in February 2022, and the associated concern about a gas shortage in Europe has given the discussion about reducing energy demand a whole new dimension in recent months. While empirical price elasticities of space heating demand can, in principle, indicate a price-induced reduction in demand that follows from an energy price increase, the current situation must be seen as structurally different from the period studied in this thesis. While the price developments between 2007 and 2019 were relatively gradual, the current situation, in contrast, represents an extreme event with an unprecedented rise in energy prices coupled with media attention and political appeals to save energy. The presence of these structural differences should be considered if the price elasticities presented in this thesis were to be used as input parameters for an analysis of the current crisis.[^8285]

[^8285]: First studies that focus explicitly on the period of the recent gas crisis already exist. See for example @Ruhnau2022demand.

**Evaluation of price-related energy policies**

Estimating price elasticities of space heating demand in the residential buildings sector is relevant for several reasons.

First, empirical evidence on price elasticities is generally important for evaluating policies that affect the price of energy. For any such price-related policy, one needs to understand the relevant determinants which affect the energy demand. Based on elasticity estimates, the economic, distributional, and environmental impacts of policy measures can be better understood and estimated [@alberini_etal11; @labandeira_etal17; @zhu_etal18]. This also applies to price-related policy instruments that aim to contribute to the attainment of ambitious climate targets.

**BEHG and its short- and long-term effects as example**

One of the essential policy measures to induce GHG emission reductions in the buildings sector in Germany is the recent introduction of a carbon price for heating fuels, which was rolled out in 2021 based on the Fuel Emissions Trading Act (BEHG).[^4] In general, the introduction of carbon pricing aims to correct the market failure of the previously unrecognised external effects of GHG emissions [@aldy_etal10; @stiglitz19]. It is expected that the consideration of external effects will have a steering effect on demand for fossil fuels through an increased price level. In practice, the recently introduced Fuel Emissions Trading Act specifies a fixed price path until 2025. The price path rises from an initial price of 25 Euros per ton of CO~2~e in 2021 to 55 Euros per ton of CO~2~e in 2025.

[^4]: Fuel Emissions Allowance Trading Act (BEHG) of 12 December 2019 (Federal Law Gazette I p. 2728) as amended by Article 1 of the Act of 3 November 2020 (Federal Law Gazette I p. 2291). Emissions from the buildings sector have not been subject to carbon pricing in the past in Germany, as they do not fall within the scope of the EU Emissions Trading Scheme (ETS). Besides emissions in the buildings sector, the BEHG also covers emissions from the transport sector. From 2026, the fixed price path is planned to be replaced by a national ETS with a defined price corridor.

Figure \@ref(fig:behg) illustrates the price effect of the carbon pricing for the buildings sector through the BEHG. For gas (Panel A) and oil (Panel B), the respective left bar (green) represents the average end consumer prices for the five years 2016-2020 from national statistics [@bmwi21].[^5] The two middle bars (yellow) show the absolute price effects that occur based on the carbon intensity of the energy carrier in 2021 (25 Euros per ton CO~2~e) and until 2025 (55 Euros per ton CO~2~e). The yellow bars are larger for oil as an energy carrier due to its higher emission intensity.[^6] Assuming otherwise constant prices, the initial CO~2~e price in 2021 would thus lead to a price increase of +0.50 Cents (+7.4%) per kWh gas and +0.66 Cents (+13.4%) per kWh oil. In 2025, the price effects rise to +1.10 Cents (+16.3%) per kWh gas and +1.46 Cents (+29.6%) per kWh oil.

[^5]: The five-year period (2016-2020) was chosen due to the volatility of oil prices in recent years. An exclusive consideration of the 2020 average price would have distorted the relative price effects. Furthermore, it should be noted that district heating is not considered in the figure because most district heating power plants were already subject to carbon pricing under the EU ETS. Only smaller plants with a total combustion capacity of less than 20 megawatts (MW) were not already subject to the EU ETS and are now subject to BEHG carbon pricing.

[^6]: The emission factors for gas (0.201 kg CO~2~e per kWh) and oil (0.266 kg CO~2~e per kWh) were taken from @uba16.

```{r behg, echo = FALSE, out.width = "100%", fig.align = "center", fig.cap = "Price effects on gas and oil induced by the Fuel Emissions Trading Act (BEHG) until 2025", fig.scap="BEHG carbon price effects on gas and oil until 2025"}
knitr::include_graphics(path = "figure/price_effect_behg.pdf")
```

The relative price increases clearly show that considerable steering effects can already be expected from the BEHG in the next few years. Higher price levels driven by the structural price component of carbon pricing can be expected to have two impact channels. The first impact channel is short-term price reactions. Consumers adjust their demand levels for heating fuels due to higher prices [@alberini_etal11]. Short-term demand responses are of direct relevance when making predictions on the fuel demand of the buildings sector. The expected magnitude of the effect depends on the demand response, which is reflected in the short-term price elasticity of space heating demand.

The second impact channel is that structurally higher fossil fuel prices will incentivise long-term investment in low-carbon heating technologies and energy efficiency. In the literature, long-term investment decisions, driven by mitigation costs potentials, are modelled using integrated assessment models (IAMs) for the buildings sector [e.g., @burger_etal19; @levesque_etal21]. Depending on the specific model structure, these IAM sector modules are often informed by assumptions on the price elasticities of heating demand as one of their input parameters.[^7] Thus, empirical estimates of the short-term price elasticity of space heating demand are also necessary for further analyses of the long-term steering effect concerning investments in energy efficiency and low-carbon technologies in the buildings sector.

[^7]: For a more detailed example of how own-price elasticities of demand are taken into account in IAMs, see the model structure description of the ETSAP TIAM model in @loulou_labriet08. Since IAMs vary in structure, other models may consider price elasticities differently. Additional highly relevant parameters include the cost of capital and substitution elasticities.

**Use of billing data as a complementary source of evidence**

Another line of argument for the relevance of this study is that the existing empirical evidence on the price elasticity of space heating demand is based on varying types of data sources. But to date, few studies have been conducted based on energy billing data. For the case of Germany, ex-post statistical analyses for the residential buildings sector predominantly rely on panel data from social surveys [e.g., @rehdanz07; @schmitz_madlener20; @schulte_heindl17]. The studies are built around the survey item *household expenditure for heating*. But actual levels of energy demand and prices are not observed. This means that, for instance, different energy contract conditions between households may obscure accurate demand levels and thus affect the validity of the estimated results. 

The data available for this thesis is different since the energy bills provide both energy demand and associated costs on the level of the buildings. To the best of the authors' knowledge, previous studies based on energy billing data have only been conducted in the United States [see @auffhammer_rubin18].

While using energy billing data offers unique characteristics that social survey data cannot mirror, it should be pointed out that its use also has its weaknesses. Due to the data structure, the billing data does not observe demand responses at an individual household level but at the level of multi-apartment buildings. This aggregation may obscure the demand response of a single household. Furthermore, the use of billing data makes the consideration of socio-economic determinants difficult (see Chapter \@ref(conceptual-model)). In summary, one can thus argue that the information value of estimates derived from billing data can be seen as complementary to the existing estimates derived from social survey data.

Furthermore, the period under investigation in this thesis (2007-2019) is more recent than the period considered in other studies, which further contributes to the relevance of this study.

**Research questions**

Given the recently tightened climate change mitigation targets for GHG emissions and the newly introduced carbon pricing for the buildings sector, additional evidence on the price elasticities of space heating demand is relevant. Therefore, this work aims to produce estimates of short-term own-price elasticities of space heating demand based on a large-scale energy billing sample in Germany. The guiding research questions for the analysis are:

-   How does a change in energy price affect the level of space heating demand (price elasticity estimate)? 
-   What other determinants affect the level of space heating demand and need to be considered so that their effects are not falsely attributed to energy prices?
-   Are there potential factors for heterogeneity in the sample that may not be reflected in the estimation of an aggregated price elasticity of demand?

The thesis is structured into six Chapters. Chapter \@ref(introduction) provided the introduction. Chapter \@ref(literature) delivers a theoretical background on the price elasticity of demand, presents the prior literature, and develops a conceptual model for this thesis. The data and empirical approach are described in the following Chapter \@ref(methods). Furthermore, Chapter \@ref(methods) also provides descriptive statistics. Chapter \@ref(results) provides the results of the statistical analysis. In Chapter \@ref(discussion), the results are discussed and integrated into the building sector's policy context. Chapter \@ref(conclusion) provides concluding remarks.