```{=html}
<!--
Chapter 2 - Theory and Literature
-->
```
```{r setup2, echo=FALSE}
pacman::p_load(tidyverse, thesisdown, knitr, kableExtra)

# Set how wide the R output will go
options(width = 70)
```

# Theory and Literature {#literature}

The objective of this chapter is to provide a short theoretical background on price elasticities as well as an overview of the previous research on price elasticities of heating demand. The text is organized as follows. The first part of the chapter focuses on the theoretical background and also provides a placement of the effects of varying elasticities in connection to heating demand. The second part of the chapter then moves to summarizing the empirical evidence from the literature.

## Price elasticity of demand {#theory}

Elasticities are one of the key concepts in mirco-economic theory. They are used to express the sensibility of one variable to a change in another. The own price elasticity of demand is one type of elasticity. It represents the magnitude of a change in consumption of a good following from a change in its own price [@pindyck_rubinfeld18]. Formally, the own price elasticity of demand $\epsilon$ can be expressed as

```{=tex}
\begin{equation}
\epsilon = \frac{\frac{\Delta Q}{Q}}{\frac{\Delta P}{P}} = \frac{P \Delta Q}{Q \Delta P}
(\#eq:ep)
\end{equation}
```
where $\Delta P/P$ represents the percentage change in the price $P$ of a good and $\Delta Q/Q$ the corresponding percentage change in the quantity $Q$ demanded of the same good. Due to budget constraints, consumers tend to consume less of a good when its price increases, which implies that under normal conditions the price elasticity of demand is negative. But the responsiveness of the demand reaction may vary across different goods. A common assumption in the econometric literature is that elasticities are constant [@varian10]. Leading to a demand function that can be expressed by

```{=tex}
\begin{equation}
Q = AP^{\varepsilon}
(\#eq:demand)
\end{equation}
```
where $A$ represents an arbitrary constant and $\epsilon$ again is the price elasticity of demand. Taking the logarithms of the demand function removes $\epsilon$ from the exponent and yields

```{=tex}
\begin{equation}
ln(Q) = ln(A) + \epsilon \; ln(P)
(\#eq:demand2)
\end{equation}
```
which can be referred to as the elasticity case and will reappear at a later point, when constructing the regression models to estimate price elasticities from the sample data.

To make the assumption of constant elasticity more intuitive, it seems useful to return to the subject of interest and show the behavior of demand curves for space heating under varying price elasticities of demand. Generally, the literature distinguishes between three different types of elasticities. If the demand response for a good is greater than the change in its price ($\epsilon < -1$), it is called price elastic. If the demand response is exactly equal to the change in price ($\epsilon = -1$), it is said to be unit-elastic. Finally, if the demand response is smaller than the price change ($0 > \epsilon > -1$) -- and therefore indicating a lower responsiveness -- it is called price inelastic [@pindyck_rubinfeld18].[^1]

[^1]: It should be noted that in the literature the minus of the demand elasticity $\epsilon$ is sometimes omitted, as it is assumed that it is generally negative. However, I personally find it more intuitive not to do so -- especially when working empirically where positive estimates are a possibility -- and will therefore continue to use the negative values in this thesis.

Figure \@ref(fig:elasticities-conceptual) visually represents these different price elasticities for the demand of space heating.[^2] To construct the demand curves, a common arbitrary intersection point is chosen for the curves. Specifically, this intersection has a consumption value of 130 kilowatt hours (kWh) per square metre (sqm) and a price of 6 cents per kWh. These values can be considered realistic for an average building in Germany, as we will see later in this thesis when working with the sample data.

[^2]: Please note that in the graph, energy price as the independent variable is plotted on the horizontal axis and energy demand (the level of consumption) as the response variable on the vertical axis. I consider this convention -- which is common in most of science -- to be more intuitive than the traditional Marshallian cross diagram in Economics, where the price is plotted on the vertical axis.

```{r elasticities-conceptual, echo = FALSE, fig.cap = "Demand curves with different price elasticities of demand", out.width = "100%", fig.align = "center"}
knitr::include_graphics(path = "figure/elasticities_plot.pdf")
```

The graph shows a total of four demand curves. In addition to the unit-elastic curve ($\epsilon = -1$), there is also an elastic ($\epsilon = -1.5$) and two inelastic ($\epsilon = -0.25, \epsilon = -0.5$) demand curves. The special feature of the unit elastic demand curve is that it gives the same result for total expenditure on heating costs for all combinations of price times demand. This is not the case with the other elasticities. First, a look to the right of the intersection (higher prices). In the elastic case (yellow curve), total expenditures fall when the price is higher, which means that households can adjust their demand downwards more. In the opposite, inelastic cases (purple and blue curves), on the other hand, households would reduce their demand less and therefore also have higher total expenditure. Conversely, to the left of the intersection point (lower prices), more elastic demand (yellow curve) leads to a higher level of consumption and thus also to higher total expenditure than in the inelastic cases (purple and blue curves), where consumption changes only gradually.

If we now take a step back and consider what kind of good the demand for space heating represents and what price elasticity of demand one might expect, it seems reasonable to assume that the elasticity is rather inelastic. There are two indications leading to this assumption. First, heating energy is to be regarded as an essential good that is associated with a minimal level of demand -- especially in the winter months. Essential goods tend to follow a more inelastic demand pattern than other types of goods [@gwartney76]. Households can only reduce their demand to a certain extent (right side of the intersection), but on the other hand, they do not have a strong rationality for a large increase in heating demand beyond a comfortable level of warmth (left side of the intersection). The second indication for a rather inelastic demand is that due to the dependency on the existing heating system -- especially for the tenant households investigated in this thesis -- there is not really a viable option to substitute space heating from one source with space heating from another.

Another important dimension for distinguishing different price elasticities of demand is time. More precisely, the time that elapses between a price signal and the measurement of the demand response. Here, the literature distinguishes between short-term and long-term price elasticities (also called short- and long-run elasticities) [@pindyck_rubinfeld18]. Short-term elasticities are those that are estimated in temporal proximity to the price change -- in this case, that would be the energy price change within a billing year and the associated change in consumption in the same period (or in the year after the price change). Long-term elasticities, on the other hand, imply that more time has elapsed for consumers to fully adjust to a price change and also to make long-term investment decisions that would change the overall level of consumption [@pindyck_rubinfeld18]. For the subject at hand, such decisions may include changes in the isolation of a building or an exchange of the heating system. Thus, long-term demand elasticities tend to be larger than their short-term counterparts [@schmitz_madlener20]. Therefore, the two types can be seen as mirroring the two different channels through which a pricing policy can influence consumer behavior (see Section \@ref(relevance)).

In the literature, long-term price elasticities are usually estimated using dynamic models, where the elasticity estimated between two periods is only a fraction of the long-term elasticity [e.g., @alberini_etal11]. However, I have chosen follow the example of @schmitz_madlener20 and only estimate short-term elasticities.This is because I believe that the long-term approaches do not apply well to the given context in which I am analyzing a subset of the total residential building segment consisting of rental buildings where households have relatively few opportunities to make long-term adjustments that would significantly affect the energy efficiency or carbon intensity of the energy of their apartments.

## Literature review {#review}

The study of the price elasticity of energy demand is a broad field of academic research which also has a long history dating back to the mid-twentieth century [e.g. @cutler41; @houthakker51]. In the more recent past, as well numerous studies have been published. These studies differ considerably in terms of their focus and design. Several conceptual factors can systematically influence the elasticity estimates. The most important of these factors is the type of energy application analysed: The price responsiveness of household space heating demand has only limited informative value for the price responsiveness of electricity consumption or transport fuel demand. The same is true for varying consumer groups: Households are likely to react differently to energy price changes than industrial consumers. Beyond, other conceptual factors which may as well affect estimates are the statistical methods used (time-series analysis, panel analysis, cross-sectional analysis), varying sources of data (national accounts, aggregate sector-level data, micro-data), varying spatial and temporal focus, measurement heterogeneity, and price trends [@csereklyei20; @miller_alberini16]. While many of the earlier studies either had only a national granularity or, when using micro data, focused predominantly on the US or European countries, a considerable number of studies have also been published recently that cover emerging countries (especially China) as well as other countries.

**Meta-estimates as first indication of elasticity magnitudes**

To organize this comprehensive body of literature with a wide range of energy applications, meta-studies do serve as a good starting point. The recently published meta-analysis by @labandeira_etal17 is particularly suitable for this purpose, as it covers a broad spectrum of energy uses and thus goes beyond previous meta-analyses that primarily focus on the elasticity of petrol demand. I will therefore first delve into the aforementioned meta-analysis before turning to a subset of selected studies that take a closer look at the specific issue of price elasticities for space heating demand in the residential buildings sector.

In their analysis, @labandeira_etal17 include a total of 428 papers, all published between 1990 and 2016. For the demand of natural gas, they find an average short-term (long-term) price elasticity of -0.18 (-0.57) based on 230 individual estimates. And for heating oil they find a comparable average short-term (long-term) elasticity of -0.19 (-0.54), which, however, is based on only 44 individual estimates [@labandeira_etal17]. An average elasticity for space heating -- which is generally less focused on in the literature due to its relatively low prevalence -- is not given.

The above average estimates of the price elasticity of demand for gas and oil, however, come with an important caveat. While they are broken down by energy carrier, they are drawn from a wide range of consumer groups. This means that the underlying studies include not only those that examine demand response patterns for residential consumption, but also others that examine these patterns for the commercial building stock and/or the industrial sector. @labandeira_etal17 also provide average price elasticities stratified by consumer groups, but these are on the other hand not clustered by energy source. When differentiated along the dimension of consumer groups, they report that the average estimates are slightly higher in the residential household sector (short-term: -0.22; long-term: -0.62) than in the industrial sector (short-term: -0.17; long-term: -0.51), which seems reasonable given that households likely have more ability to change their consumption levels than the industrial sector with more fixed production patterns and the ability to forward costs. Overall, however, the average elasticity estimates are all in a similar size range and together convey the message that demand is strongly inelastic in the short-term, confirming the theoretical arguments that energy is an essential good in many applications and also has limited substitutability (see section \@ref(theory)).

The meta-estimates are useful in providing a high-level overview. However, as already mentioned, they include estimates from a wide range of energy applications that not necessarily reflect the demand response for the specific application of space heating in residential buildings. Furthermore, they are average estimates which may have the effect of masking discrepancies and variations between studies. Therefore, the following section presents individual studies that explicitly focus on the space heating demand of the residential sector.

**Selected studies on space heating in the residential buildings sector**

The space heating demand of the residential sector represents a subset of the full literature stream. The selected studies were chosen based on the criteria that they correspond to the focus of this work and are of good quality (peer-reviewed journal articles). A summary of the studies is provided in Table \@ref(tab:estimates-literature).

There are three earlier studies that have a spatial focus on Germany. The first study was conducted by @rehdanz07, who estimates price elasticities for different energy carriers for residential heating demand. She uses social survey data on household-level heating expenditures from the Socio-Economic Panel (SOEP) in the years 1998 and 2003. Methodologically, a cross-sectional statistical analysis with dummy variables is conducted for the two years -- in other words, no panel structure is used. In contrast to the average meta-estimates presented in the previous section, the study finds that demand for heating oil has a high elasticity of -1.68 to -2.03, depending on the model specification. For gas, the results indicate a less elastic demand between -0.44 and -0.63, also depending on the model used. Therefore the study concludes that the type of fuels can have a strong influence on the price sensitivity of households [@rehdanz07].


\begin{table}[]
\centering
\caption{Selected studies on the price elasticity of heating demand}
\label{tab:estimates-literature}
\resizebox{\textwidth}{!}{%

\begin{tabular}{@{}llll@{}}
\toprule
\textbf{Study} & \textbf{Type of data used} & \textbf{\begin{tabular}[c]{@{}l@{}}Energy \\ carrier\end{tabular}} & \textbf{\begin{tabular}[c]{@{}l@{}}Short-term\\ price elasticities\end{tabular}} \\ \midrule
 &  &  &  \\
{\ul \textit{Studies within Germany}} &  &  &  \\
\multirow{2}{*}{Rehdanz (2007)} & \multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Social survey, panel data (SOEP),\\ all types of buildings, 1998 and 2003\end{tabular}} & Gas & -0.44 to -0.63 \\
 &  & Oil & -1.68 to -2.03 \\
Schmitz and Madlener (2020) & \begin{tabular}[c]{@{}l@{}}Social survey, panel data (SOEP), \\ all types of buildings, 1996-2014\end{tabular} & (All) & -0.31 to -0.43 \\
Schulte and Heindl (2017) & \begin{tabular}[c]{@{}l@{}}Social survey, panel data (EVS),\\ all types of buildings, 1993-2008\end{tabular} & (All) & -0.50 \\
 &  &  &  \\
{\ul \textit{Studies outside of Germany}} &  &  &  \\
Alberini et al. (2011) & \begin{tabular}[c]{@{}l@{}}US, household-level social survey, \\ 50 metropolitan areas, panel data, 1995-2007,\\ only single-family homes and duplexes\end{tabular} & Gas & -0.57 to −0.69 \\
Auffhammer and Rubin (2018) & \begin{tabular}[c]{@{}l@{}}US, energy billing data, household-level, \\ only California, panel data, 2003-2014\end{tabular} & Gas & -0.17 to -0.23 \\
\begin{tabular}[c]{@{}l@{}}Leth-Petersen and \\ Togeby (2001)\end{tabular} & \begin{tabular}[c]{@{}l@{}}Denmark, apartment-block level (\textgreater{}1,500 sqm),\\ panel data, 1984-1995\end{tabular} & Oil & -0.08 \\
 &  & District heating & -0.02 \\
\multirow{2}{*}{Meier and Rehdanz (2010)} & \multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}UK, household-level social survey,\\ panel data 1991-2005\end{tabular}} & Gas & -0.34 to -0.56 \\
 &  & Oil & -0.40 to -0.49 \\
Metcalf and Hassett (1999) & \begin{tabular}[c]{@{}l@{}}US, household-level, panel data, \\ 1984, 1987 and 1990\end{tabular} & Gas & -0.48 to -0.71 \\ \bottomrule
\end{tabular}%

}
\end{table}


In a more recent study, also based on the SOEP social survey data but covering a more comprehensive time period between 1996 and 2014, @schmitz_madlener20 find a price elasticity of heating and hot water demand of -0.31 to -0.43, depending on model specifications. They do not differentiate the elasticities by energy carrier. The price elasticities are derived from household expenditure elasticities, as demand is not directly observed. In contrast to @rehdanz07, the study uses a panel structure where observations are clustered by household and year using fixed effects. This produces much lower price elasticity estimates, which are more consistent with the meta-estimates presented in the previous section. Moreover, the study discovers that price elasticity is heterogeneous across different socio-economic groups. Given the household-level information at their disposal, they find that higher-income households are less sensitive to energy price changes than lower-income households. Likewise, homeowners show less sensitivity than renters [@schmitz_madlener20].

Also for Germany, @schulte_heindl17 investigate the price and expenditure elasticities of private energy demand in Germany between 1993 and 2008. For their analysis, they use survey data from the *Einkommens- und Verbrauchsstichprobe (EVS)* and analyse them with a demand system approach (in particular, they estimate a quadratic expenditure system). For space heating, they find an own-price elasticity of -0.50 across all households. They further observe that the price elasticity changes with the level of total household expenditure, with households in higher expenditure strata reacting more strongly to price changes [@schulte_heindl17]. Put differently, this means that following an increase in price level low income households tend to decrease energy consumption to a lesser extent as compared to households with a higher income. Thus, in relation to the income dimension the results must be understood as contradictory to those of @schmitz_madlener20.

All three studies on Germany rely on data from household-level social surveys. Therefore, the studies take a strong perspective on the impact of household income on energy price sensitivity. As the data for this study is aggregated at the building-level and not at the household-level, however, the focus of this thesis is to some extent different. The focus is rather on the overall price response based on actual price and consumption values, which are not available in the social surveys, and on the response in interaction with other characteristics and changes at the building level.

Besides the studies focusing on Germany, Table \@ref(tab:estimates-literature) also reports the evidence from five selected international studies focusing on the price elasticities of space heating demand. @alberini_etal11 examine household demand for gas (and electricity) in single-family homes and duplexes using household-level social survey data in 50 metropolitan areas in the United States (US) over the period 1995-2007. As a modelling approach, they use static and dynamic panel models. For the static models reflecting the approach taken in this thesis, they find an own price elasticity of private gas demand between -0.57 and -0.69, depending on the model specification. These estimates are in their magnitude comparable to the results of an earlier study by @metcalf_hassett99 who use the 1984, 1987 and 1990 waves of the US Residential Energy Consumption Survey (RECS) to examine homeowners' investments in energy efficiency measures and as part of their study find price elasticities of residential gas demand to range between −0.48 and −0.71. In contrast to these estimates, a recent study by @auffhammer_rubin18 finds that the short-term price elasticity of residential gas demand is even more inelastic, ranging from -0.17 to -0.23, not for the whole US but for California in particular. The study by @auffhammer_rubin18 is especially interesting for this thesis because it is the only one that also uses energy billing data. The data analysed covers the period 2003-2014 and consists of monthly energy bills at the household-level. So although the type of data source used to derive the estimates is the most similar to the approach taken in this thesis, there are still two relevant differences: firstly, gas tariffs in the US nay change monthly rather than annually, as is the case in the German dataset used for this analysis, and secondly, the data in @auffhammer_rubin18 relates to individual households rather than the aggregate building-level.

In addition to these three studies focusing on the US, table \@ref(tab:estimates-literature) also contains the results of a Danish study by @leth-petersen_togeby01. They use a panel data approach and analyse data from the period 1984-1995. Interestingly, they also use the building-level as the level of analysis. More specifically, they study apartment buildings with a base area of more than 1,500 square meters (sqm) in Denmark. As an empirical strategy, they rely on fixed effects models. As a result, they find very inelastic demand responses for space heating. For oil, the short-term price elasticity is found to be -0.08 and for district heating as low as -0.02, meaning almost no demand response to a changing price at all. Another study by @meier_rehdanz10 looks at the United Kingdom (UK) and analyses a household-level social survey panel for the 15-year period 1991-2005. For their model they use a log-linear approach. They find that the short-term price elasticity of demand for gas ranges from -0.34 to -0.56, depending on the specification. For residential oil consumption, the results are in the same order of magnitude, but narrower, between -0.40 and -0.49. In addition, they find that the type of occupant is a relevant dimension for a different price responsiveness. While renters adjusted their consumption more, homeowners showed a more inelastic response.

**Synthesis of estimates from the literature**

If one aims to summarize the literature on the price elasticity of residential space heating demand presented previously, commonalities and differences emerge. The most important commonality is that the price elasticity of energy demand in general and household space heating demand in particular is inelastic in almost all empirical estimations. However, when looking at the individual studies, differences in the magnitude of the estimates appear. While the meta-results by @labandeira_etal17 as well as a number of individual studies indicate a strongly inelastic demand response [e.g. @auffhammer_rubin18; @leth-petersen_togeby01], other individual studies still indicate an inelastic but stronger demand response [e.g., @alberini_etal11; @meier_rehdanz10; @metcalf_hassett99; @rehdanz07; @schmitz_madlener20; @schulte_heindl17]. These differences in magnitude may be partly due to the different local and temporal conditions of the studies or due to differences in the data and methodology. However, other aspects such as climatic conditions, the building stock or behavioral aspects related to a specific population may also be important. Therefore, it should be noted that all three studies that focused on Germany found somewhat more sensitive price elasticities ranging from -0.3 to -0.6.[^3]

[^3]: Note that the extreme results of @rehdanz07 for oil as an energy source, which come only from a cross-sectional regression analysis, are omitted here. Since only two isolated periods are considered, these results must be considered more prone to extreme results than is the case with a longer panel.

Another common feature of all the individual studies presented is that they do not focus on one-off extreme price shocks as an identification strategy, but rather on gradual price developments. They are mostly built around a panel data set covering a longer period of time. The period studied in this paper is also a longer period with rather gradual price developments. This should therefore be seen as an important aspect that will facilitate the comparability of the results of this study with the literature estimates presented.


## Conceptual model development {#conceptual_model}

In the previous two subsections, the theoretical basis for a constant own-price elasticity of demand in the context of space heating was presented (see section \@ref(theory)) and previous evidence for empirical estimates (see section \@ref(review)). To complete this foundation, this subsection uses theory, but most importantly the approaches of other studies, to identify relevant variables in the relationship between space heating demand and energy price. The aim is to build a conceptual model of the relationship, which I will empirically examine in the remainder of this thesis.

The starting point for the conceptual model is the relationship between space heating demand and energy price. Here, the basic assumption is that the demand for space heating is influenced by the energy price. If the price increases, the demand decreases. If the price goes down, demand goes up. Transferred to statistical terminology this implies that space heat demand is the response variable of a model and energy price the (main) explanatory variable. To support the development of the conceptual model visually, Figure \@ref(fig:dag) shows a directed acyclic graph (DAG). Since the relationship between space heating demand and energy price is the one of main concern, it is marked with a red arrow.

However, beyond energy price there is a whole set of additional variables which might affect the level of space heat demand. Those are represented in the DAG as additional explanatory variables and as grouping variables. First, focusing on the additional explanatory variables, the local climatic conditions are of relevance. 



he model will likely only reflect the relationships that have an effect on demand for space heating. Yet, for a first overview I have mapped other potential relationships between controls as well. It is important to note that the model shown in the graph should be considered preliminary and variables may be added or removed after exploratory testing.

```{r dag, echo = FALSE, out.width = "100%", fig.align = "center", fig.cap = "Conceptual directed acyclic graph", fig.scap="Conceptual directed acyclic graph"}
knitr::include_graphics(path = "figure/dag_red.pdf")
```




