```{=html}
<!--
Chapter 2 - Theory and Literature
-->
```
```{r setup2, echo=FALSE}
pacman::p_load(tidyverse, thesisdown, knitr, kableExtra)

# Set how wide the R output will go
options(width = 70)
```

# Theory and Literature {#literature}

This Chapter aims to provide the theoretical background on the price elasticity of demand. It then summarises evidence from previous literature on the price elasticity of space heating demand. Subsequently, theory and literature evidence are used to develop a conceptual model with corresponding rationales for the relevance and expected direction of a variable's effect on space heating demand.

## Price Elasticity of Demand {#theory}

In general, elasticities are a key concept in microeconomic theory. They express the sensibility of one variable to a change in another. The own-price elasticity of demand is a specific type of elasticity. It represents the magnitude of a change in demand for a good following a change in its own price [@pindyck_rubinfeld18]. Formally, the own-price elasticity of demand $\epsilon$ can be expressed as:

```{=tex}
\begin{equation}
\epsilon = \frac{\frac{\Delta Q}{Q}}{\frac{\Delta P}{P}} = \frac{P \Delta Q}{Q \Delta P}
(\#eq:ep)
\end{equation}
```

Where $\Delta P/P$ represents the percentage change in the own price $P$ of a good and $\Delta Q/Q$ the corresponding percentage change in the quantity $Q$ demanded of the same good. Due to budget constraints, consumers tend to demand less of a good when its price increases, which implies that under normal conditions, the price elasticity of demand is negative. But the magnitude of the demand reaction may vary across different goods. A common conception in the econometric literature is that elasticities are a constant parameter [@varian10]. Under the assumption of a constant price elasticity of demand, the associated demand function can be expressed as:

```{=tex}
\begin{equation}
Q = AP^{\varepsilon}
(\#eq:demand)
\end{equation}
```

Where $A$ represents an arbitrary constant and $\epsilon$ again is the price elasticity of demand. Taking the logarithms of the demand function removes $\epsilon$ from the exponent and yields:

```{=tex}
\begin{equation}
ln(Q) = ln(A) + \epsilon \; ln(P)
(\#eq:demand2)
\end{equation}
```

The form of the equation can be referred to as the elasticity case. The equation will reappear later in this thesis when constructing the statistical regression models.

**The role of varying price elasticities of demand**

To make the underlying assumption of a constant elasticity more intuitive, it seems helpful to relate it to the subject of interest and show the behaviour of demand curves for space heating under varying price elasticities of demand. Generally, the microeconomics literature distinguishes between three different *types* of elasticities. If the demand response for a good is greater than the change in its own price ($\epsilon < -1$), the good is called price elastic. If the demand response is equal to the price change ($\epsilon = -1$), it is said to be unit-elastic. And if the demand response is smaller than the price change ($0 > \epsilon > -1$) --  indicating lower price responsiveness -- it is said to be price inelastic [@pindyck_rubinfeld18].[^8]

[^8]: It should be noted that in the literature, the minus of the demand elasticity $\epsilon$ is sometimes omitted, as it is assumed that it is generally negative. However, I find it more intuitive not to do so, especially when working empirically where non-negative estimates are possible. I will therefore continue to use the negative values in this thesis.

```{r elasticities-conceptual, echo = FALSE, fig.cap = "Exemplary demand curves associated with varying price elasticities of space heating demand", out.width = "100%", fig.align = "center"}
knitr::include_graphics(path = "figure/elasticities_plot.pdf")
```

Figure \@ref(fig:elasticities-conceptual) graphically represents demand curves for varying price elasticities of demand for space heating.[^9] A common arbitrary intersection point is chosen to construct the demand curves. The intersection point has a demand value of 113 kilowatt hours per square meter and year (kWh/sqm/y) and a price value of 6.49 Cents/kWh. These demand and price values are chosen as realistic values because they reflect the average values in the empirical sample (see Chapter \@ref(descriptives)). The graph includes a total of four demand curves with varying price elasticities: A unit-elastic ($\epsilon = -1$) demand curve, an elastic ($\epsilon = -1.5$) demand curve and two inelastic ($\epsilon = -0.25, \epsilon = -0.5$) demand curves.

[^9]: Please note that in the graph, energy price as the independent variable is plotted on the horizontal axis and energy demand as the response variable on the vertical axis. I consider this convention -- which is common in most science disciplines -- to be more intuitive than the traditional Marshallian cross diagram in Economics, where the price is plotted on the vertical axis.

The unique feature of the unit-elastic demand curve (green) is that it gives the same result for total expenditure on heating costs for all combinations of price and demand. In contrast, this is not the case with the other price elasticities. In the elastic case (yellow curve), total expenditures become lower when the price is higher, which means that households can adjust their demand downwards more. In the opposite, inelastic cases (purple and blue curves), on the other hand, households would reduce their demand less and therefore also face higher total expenditure. Conversely, to the left of the intersection point (lower prices), more elastic demand (yellow curve) leads to a higher level of energy demand and thus also to higher total expenditure than in the inelastic cases (purple and blue curves), where demand changes only gradually.

Different types of goods are structurally associated with different price elasticities of demand [@varian10]. For the household demand for space heating, it is reasonable to assume that demand will be inelastic. Two arguments support this assumption. First, heating energy is an essential good associated with minimal demand in the winter months to ensure residents' well-being. Usually, essential goods follow a more inelastic demand pattern than other goods [@gwartney76]. Households can only reduce their space heating demand to a certain extent (right side of the intersection) and do not have strong rationality for a substantial increase in heating demand beyond a comfortable level of warmth (left side of the intersection). The second indication for a rather inelastic demand is that tenant households, from which the data for this thesis originates, have limited to no viable options to replace space heating from one source with another due to their dependence on the existing heating system. Limited substitute ability typically accompanies the necessity for continued demand [@pindyck_rubinfeld18].

**The difference between short- and long-term price elasticities of demand**

Another relevant dimension that distinguishes different categories of price elasticities of demand is *time*. More precisely, the time that elapses between a price signal and the measurement of the demand response. Here, the literature distinguishes between short-term and long-term price elasticities of demand (also called short- and long-run elasticities) [@pindyck_rubinfeld18]. Short-term elasticities of demand are estimated in temporal proximity to the price change. For space heating, that would be the energy price change within a billing year and the associated change in demand in the same period. Short-term price elasticities are mainly driven by behavioural changes such as lowering the average heating temperature or adjusting the ventilation behaviour. But may also include minor constructional improvements, such as installing door and window sealings. Long-term elasticities, on the other hand, imply that more time has elapsed for consumers to adjust to a price change fully and make long-term investment decisions that would change the overall level of demand [@pindyck_rubinfeld18]. For the space heating demand of residential buildings, such decisions may include significant investments into the thermal insulation of a building or an exchange of the heating system. Due to these structural adjustments, long-term demand elasticities are usually higher (more elastic) than short-term demand elasticities [@schmitz_madlener20]. Overall, the two types of short- and long-term elasticities of demand can be seen as mirroring the two different channels through which a price-related policy instrument such as the BEHG can influence the demand behaviour of residents (see Chapter \@ref(relevance)). Short-term elasticities reflect the immediate demand reaction. And long-term elasticities observe the structural adjustments made based on price developments in the long run.

In the literature, the approaches in individual studies differ in the sense that they estimate either short-term price elasticities or both. For this thesis, I follow the example of @schmitz_madlener20 and only estimate short-term price elasticities of demand. Long-term price elasticities are often estimated using dynamic models, where the elasticity estimated between two periods represents a fraction of the long-term elasticity [e.g., @alberini_etal11]. However, since the methods vary more, I consider it outside of the limited scope of this thesis.

## Literature Review {#review}

In general, the study of the price elasticity of *energy* demand is a broad field of academic research which also has a long history dating back to the mid-twentieth century [e.g., @cutler41; @houthakker51]. Also, in the more recent past, many academic studies have been published. These studies differ considerably in terms of focus and design. Most of the literature focuses on energy demand in general or other types of specific energy applications. But some studies focus specifically on the demand responsiveness of private households for space heating. To organise the vast body of literature, I first focus on meta-studies to provide the first indication of price elasticity for energy demand in general and then concentrate on the subset of relevant studies focusing on household space heating demand.

Several conceptual factors can systematically influence price elasticity estimates. The most important of these factors is the type of energy application analysed. The price responsiveness of private households for space heating demand differs from the price responsiveness for other types of energy applications, such as the demand for transport fuels. Another relevant factor is different groups of consumers. Private households are likely to react differently to energy price changes than industrial or commercial consumers. Beyond those two dimensions, other factors which may also have an effect on elasticity estimates are the statistical methods used (time-series analysis, panel analysis, cross-sectional analysis), varying sources of data (national accounts, aggregate sector-level data, micro-data), varying spatial and temporal focus, measurement heterogeneity, and price trends [@csereklyei20; @miller_alberini16].

**Meta-studies on the price elasticity of energy demand**

To obtain a first indication of price elasticity estimates from previous studies, the meta-analysis by @labandeira_etal17 is particularly suitable. It covers a broad spectrum of energy applications and thus goes beyond previous meta-analyses that primarily focus on the elasticity of gasoline demand [e.g., @brons_etal08; @havranek_etal12]. Moreover, it is the most recent comprehensive meta-analysis and thus also partly covers the period relevant for this thesis.

In their analysis, @labandeira_etal17 included a total of 428 papers, all published between 1990 and 2016. For the demand for natural gas, they find an average short-term (long-term) price elasticity of -0.18 (-0.57) based on 230 individual estimates. And for heating oil, they find a comparable average short-term (long-term) elasticity of -0.19 (-0.54), which, however, is based on only 44 individual estimates [@labandeira_etal17]. An average elasticity for district heating is not given, as it is comparatively less widespread and therefore finds fewer references in the literature. The average estimates of the price elasticity of demand for the energy carrier groups gas and oil include a wide range of consumer groups. This means that the underlying individual studies not only look at the demand reactions of private households but also investigate the commercial building stock and/or the industrial sector. 

@labandeira_etal17 also provide average price elasticities stratified by consumer groups, but these are, on the other hand, not clustered by the type of energy carrier. When differentiated along the dimension of consumer groups, they report that the average estimates are slightly higher in the residential household sector (short-term: -0.22; long-term: -0.62) than in the industrial sector (short-term: -0.17; long-term: -0.51), which seems reasonable given that households likely have more ability to change their level of demand than the industrial sector with more fixed production patterns and the ability to forward costs. Overall, however, the average price elasticity of demand estimates are all in a similar range and convey the message that energy demand is strongly inelastic in the short-term. This evidence is in line with the theoretical arguments that energy is an essential good and involves limited substitutability in many applications (see Chapter \@ref(theory)).

The meta-estimates help provide a high-level overview. However, as already mentioned, they include estimates from a wide range of energy applications that do not necessarily reflect the demand response for the specific application of space heating in residential buildings. Furthermore, they are average estimates which may have the effect of masking discrepancies and variations between studies.

**Subset of studies focusing on the price elasticity of space heating demand in the residential buildings sector**

The studies focusing on the space heating demand of the residential sector represent a subset of the literature stream estimating price elasticities of energy demand. The individual studies presented in the following are selected based on the twin criteria that they specifically focus on space heating in the residential buildings sector and are of good quality (peer-reviewed journal articles or discussion papers of high quality). Table \@ref(tab:estimates-literature) provides an overview of the individual studies.

Three earlier studies have a spatial focus on Germany. The first study was conducted by @rehdanz07, who estimates short-term price elasticities for different energy carrier groups for residential space heating demand. She uses social survey data on household-level heating expenditures from the Socio-Economic Panel (SOEP) in 1998 and 2003. Methodologically, a cross-sectional statistical analysis with dummy variables is conducted for the two observed waves. In contrast to the meta-estimates presented in the last part of the Chapter, the study finds that demand for heating oil is highly elastic, with estimates ranging between -1.68 and -2.03, depending on the model specification. For gas, the results indicate an inelastic elastic demand between -0.44 and -0.63, depending on the model specification, which is nevertheless more elastic than the previously presented meta-estimates. The study concludes that the type of energy carrier can strongly influence private households' price sensitivity [@rehdanz07].


\begin{table}[]
\centering
\caption{Individual studies focusing on the price elasticity of space heating demand}
\label{tab:estimates-literature}
\resizebox{\textwidth}{!}{%

\begin{tabular}{@{}llll@{}}
\toprule
\textbf{Study} & \textbf{Type of data used} & \textbf{\begin{tabular}[c]{@{}l@{}}Energy \\ carrier\end{tabular}} & \textbf{\begin{tabular}[c]{@{}l@{}}Short-term\\ price elasticities\end{tabular}} \\ \midrule
 &  &  &  \\
{\ul \textit{Studies with spatial focus on Germany}} &  &  &  \\
\multirow{2}{*}{Rehdanz (2007)} & \multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}Social survey, cross-sectional data (SOEP),\\ all types of buildings, 1998 and 2003\end{tabular}} & Gas & -0.44 to -0.63 \\
 &  & Oil & -1.68 to -2.03 \\
Schmitz and Madlener (2020) & \begin{tabular}[c]{@{}l@{}}Social survey, panel data (SOEP), \\ all types of buildings, 1996-2014\end{tabular} & (All) & -0.31 to -0.43 \\
Schulte and Heindl (2017) & \begin{tabular}[c]{@{}l@{}}Social survey, panel data (EVS),\\ all types of buildings, 1993-2008\end{tabular} & (All) & -0.50 \\
 &  &  &  \\
{\ul \textit{Studies with focus on other countries and regions}} &  &  &  \\
Alberini et al. (2011) & \begin{tabular}[c]{@{}l@{}}US, household-level social survey, \\ 50 metropolitan areas, panel data, 1995-2007,\\ only single-family homes and duplexes\end{tabular} & Gas & -0.57 to −0.69 \\
Auffhammer and Rubin (2018) & \begin{tabular}[c]{@{}l@{}}US, energy billing data, household-level, \\ only California, panel data, 2003-2014\end{tabular} & Gas & -0.17 to -0.23 \\
\begin{tabular}[c]{@{}l@{}}Leth-Petersen and \\ Togeby (2001)\end{tabular} & \begin{tabular}[c]{@{}l@{}}Denmark, apartment-block level (\textgreater{}1,500 sqm),\\ panel data, 1984-1995\end{tabular} & Oil & -0.08 \\
 &  & District heating & -0.02 \\
\multirow{2}{*}{Meier and Rehdanz (2010)} & \multirow{2}{*}{\begin{tabular}[c]{@{}l@{}}UK, household-level social survey,\\ panel data 1991-2005\end{tabular}} & Gas & -0.34 to -0.56 \\
 &  & Oil & -0.40 to -0.49 \\
Metcalf and Hassett (1999) & \begin{tabular}[c]{@{}l@{}}US, household-level, panel data, \\ 1984, 1987 and 1990\end{tabular} & Gas & -0.48 to -0.71 \\ \bottomrule
\end{tabular}%

}
\end{table}


In a more recent study also based on the SOEP social survey data but covering a more extended period between 1996 and 2014, @schmitz_madlener20 found a price elasticity of space heating and hot water demand between -0.31 and -0.43, depending on the model specifications. They do not differentiate the elasticities by energy carrier group. The price elasticities are derived from household expenditure elasticities, as demand is not directly observed in the SOEP social survey data. In contrast to @rehdanz07, the study uses a panel structure where observations are clustered by households and years using fixed-effects regression models. They produce much lower price elasticity estimates, which are more consistent with the previously presented meta-estimates. Moreover, the study discovers that price elasticity is heterogeneous across different socio-economic groups. They find that higher-income households are less sensitive to energy price changes than lower-income households. Likewise, homeowners show less sensitivity than renter households [@schmitz_madlener20].

Also, with a focus on Germany, @schulte_heindl17 investigate the price and expenditure elasticities of private energy demand in Germany between 1993 and 2008. They use social survey data from the *Einkommens- und Verbrauchsstichprobe (EVS)* and analyse it with a demand system approach. More precisely, they estimate a quadratic expenditure system. For space heating demand, they find an own-price elasticity of -0.50 across all households. They further observe that low-income households reduce their energy demand to a lesser extent after an energy price increase than higher-income households. Thus, their results regarding the effect of household income contradict those of @schmitz_madlener20.

Besides the studies focusing on Germany, Table \@ref(tab:estimates-literature) also reports the findings from five other international studies. @alberini_etal11 examined household demand for gas (and electricity) in single-family homes and duplexes using household-level social survey data in 50 metropolitan areas in the United States (US) from 1995-2007. They use static (short-term elasticities) and dynamic (long-term elasticities) panel models as a modelling approach. For the static models, reflecting the approach also taken in this thesis, they find an own-price elasticity of private gas demand between -0.57 and -0.69, depending on the model specification. These estimates are in their magnitude comparable to the results of an earlier study by @metcalf_hassett99. They use the 1984, 1987 and 1990 waves of the US Residential Energy Consumption Survey (RECS) to examine homeowners' investments in energy efficiency measures and, as part of their study, find price elasticities for residential gas demand range between −0.48 and −0.71.

In contrast to these estimates, a recent study by @auffhammer_rubin18 finds that the short-term price elasticity of residential gas demand is much more inelastic. Instead of covering the whole US, the study focuses on California. The estimates for the price elasticity of gas demand range between -0.17 and -0.23. The study by @auffhammer_rubin18 is especially relevant for this thesis because it is the only previous study that relies on energy billing data. The data analysed covers the period between 2003 and 2014 and consists of monthly energy bills at the household level. Although the data is similar in principle, two structural differences exist compared to the energy billing data used in this thesis. First, gas tariffs in the US can change monthly rather than annually. Thus, the dataset includes more frequent observations. Furthermore, another difference is that the data from @auffhammer_rubin18 corresponds to individual households and not to the aggregate building level.

In addition, another study by @leth-petersen_togeby01 provides evidence from Denmark. They rely on a panel data approach and analyse data from 1984-1995 on apartment buildings with a more than 1,500 square meters of floor area. Similar to this thesis, they also use the building level as the level of analysis. As an empirical strategy, they rely on fixed-effects models and find highly inelastic demand responses for space heating. For oil, the short-term price elasticity is found to be -0.08, and for district heating to be -0.02, meaning almost no demand response to a changing price.

The last individual study by @meier_rehdanz10 focuses on the United Kingdom (UK) and analyses a household-level social survey panel from 1991-2005. For their model, they use a log-linear approach. They find that the short-term price elasticity of demand for gas ranges from -0.34 to -0.56, depending on the specification. For residential oil demand, the results are in the same order of magnitude, but narrower, between -0.40 and -0.49. In addition, they find that the type of occupant is a relevant dimension for varying demand responses. While renters showed higher responsiveness to energy price changes, homeowners showed a more inelastic demand response.

**Synthesis of price elasticity estimates from the literature**

When summarising the literature on the price elasticity of residential space heating demand presented previously, commonalities and differences emerge. The most important commonality is that the price elasticity of energy demand in general and of space heating demand in private households in particular is inelastic in almost all empirical estimates. While the meta-results by @labandeira_etal17 and some individual studies indicate a strongly inelastic demand response [e.g. @auffhammer_rubin18; @leth-petersen_togeby01], other individual studies still indicate an inelastic but more comprehensive demand response [e.g., @alberini_etal11; @meier_rehdanz10; @metcalf_hassett99; @rehdanz07; @schmitz_madlener20; @schulte_heindl17]. These differences in magnitude may be partly due to the individual studies' different local and temporal conditions or due to differences in the data and methodology.

```{r literature-estimates-plot, echo = FALSE, fig.cap = "Price elasticity estimates for space heat demand from individual studies", out.width = "100%", fig.align = "center"}
knitr::include_graphics(path = "figure/plot_literature_estimates.pdf")
```

Figure \@ref(fig:literature-estimates-plot) graphically summarises the short-term elasticity estimates from the individual studies. The studies which focused on Germany found somewhat more responsive price elasticities ranging between -0.3 and -0.6.[^10] But the majority of other studies' estimates can also be found in that range.

[^10]: Note that the extreme results of @rehdanz07 for oil as the carrier type are omitted here. They rely on a cross-sectional regression analysis with only two isolated periods considered. Therefore, the estimates must be regarded as more prone to extreme results than more extended data panels.

Another common feature that links all the individual studies presented is that they do not focus on one-off extreme price shocks as an identification strategy but mainly observe gradual price developments. Furthermore, they are mostly built around a panel dataset covering a more extended period. The period examined in this study also covers more than a decade characterised by relatively gradual price developments. This correspondence between the conditions underlying the estimates in the literature and this study is an essential condition that will later facilitate the comparability of the results.

## Conceptual Model {#conceptual-model}

The last two parts of the Chapter presented the theoretical foundations of the price elasticity of demand (see Chapter \@ref(theory)) and the previous evidence from the literature on household demand response for space heating to energy price changes (see chapter \@ref(review)). To synthesise this basis, this final part of the Chapter is dedicated to developing a conceptual model relying on both theory and approaches from previous studies. The aim is to identify relevant determinants for the level of space heating demand of private households and to justify their relevance. The conceptual model will then be transferred into a statistical model and empirically examined in the further course of this thesis.

**Price elasticity of demand**

The starting point for the development of the conceptual model is the relationship between energy price and space heating demand, the assumption from the theory of price and demand being that the demand for space heating is influenced by the changes in the energy price. If the energy price increases, energy demand moves down along the demand curve and vice versa (see Figure \@ref(fig:elasticities-conceptual)). Expressed in statistical terminology, this implies that space heat demand is the response variable of a model and energy price is the main explanatory variable. To support the development of the conceptual model visually, Figure \@ref(fig:dag) depicts it as a directed acyclic graph (DAG). At the centre of the DAG, the response variable space heating demand (yellow dot) is shown, the variance of which is to be explained. The effect of energy price (purple dot) on space heating demand represents the *price elasticity of demand* and is therefore highlighted by the red arrow. The temporal dimension is also relevant to the rationale of how the energy price influences energy demand. More precisely, the question of whether the energy price in the current billing period determines demand levels or the price of the previous billing period, which is communicated to households through the delivery of the previous year's energy bill. In this thesis, I follow the approach used in most previous studies and employ the energy price in the same period as the standard approach. However, I will also present results for the case where the energy price of the previous billing period is used as the basis for demand decisions to obtain a more thorough picture.

```{r dag, echo = FALSE, out.width = "100%", fig.align = "center", fig.cap = "Conceptual model as directed acyclic graph", fig.scap="Conceptual model as directed acyclic graph"}
knitr::include_graphics(path = "figure/dag_red.pdf")
```

In addition to energy price, various other determinants may influence the level of space heating demand. These are shown in the DAG as additional explanatory variables (blue dots) and as grouping variables (green dots). A structured approach was used to determine the set of additional relevant variables. First, the recent study by @schmitz_madlener20 was considered a suitable starting point for creating a list of variables. Second, the other individual studies presented previously (see Chapter \@ref(review)) were then used to cross-validate the initial list of variables and to screen for further ones. This approach led to the variables presented and reasoned for in the following.

**Climatic conditions**

First, considering the local climatic conditions in a billing period is important, as they can substantially affect space heating demand [@hesse20]. Due to the high relevance of climatic conditions, all individual studies presented in Chapter \@ref(review), except @schulte_heindl17, consider and control for them. For practical implementation, climatic conditions are usually approximated by the outdoor temperature over a defined period and aggregated to periodical (heating) degree days. The basic intuition for the influence of varying climatic conditions is that with lower outdoor temperatures, transmission heat losses increase and thus the demand for space heating increases to compensate for indoor heat losses. In line with @vdi13, degree days in this study are defined as the temperature difference between a room temperature of 20°C and the daily average outdoor temperature, provided it is below a heating limit of 15°C.[^12] Lower mean outdoor temperatures are associated with a higher aggregated number of degree days in a period. Consequently, as a rationale for the direction of the effect of degree days, one would expect a higher space heating demand when degree days are higher in an annual heating period and vice versa. Varying degree days occur on a temporal (one period is warmer/cooler than another) and on a spatial scale (places at higher altitudes or in the continent's interior have structurally lower temperatures).

[^12]: It should be noted that the definition of degree days differs from the internationally frequently used definition of heating degree days (HDD), which is calculated by the difference between the average outdoor temperature below a heating threshold over time regardless of an additionally defined room temperature. In this thesis, the VDI-based definition of *degree days* was chosen, as this is the standard methodology in Germany. It is, for example, also used to create comparable scoring for the official building energy performance certificates (see @halbig_namyslo14).

**Building-level characteristics**

In addition, two building-level characteristics are included in the model as additional explanatory variables. First, the size of a building can influence the heat demand, as in larger buildings the ratio between the external surface of the building and the heating surface decreases. Thus, the relative share of transmission heat losses per unit of heating surface is lower correspondingly. Following this rationale, one must assume that in buildings with a larger heating surface -- which approximates the size of a building -- the space heating demand per unit is lower if all other factors remain constant.

Second, the type of energy carrier used for the heating system is another building-level characteristic considered in almost all previous studies (see Chapter \@ref(review)). The intuition for including the energy carrier as a categorical control variable (oil, gas, district heating) is that different types of energy carriers are associated with varying heating technologies, which may lead to structural differences in energy use and efficiency. This effect may also be further amplified by the fact that certain energy carriers have been more commonly introduced in specific periods in the past. So, for example, an average oil heating system may be older and thus less energy efficient than an average gas heating system due to technological progress.

**District socio-economic variables**

Besides building-related characteristics, socio-economic variables may also affect the level of space heating demand. In the literature that draws its data from social surveys, household-level socio-economic information such as income, number of inhabitants, age, education level, or employment status is usually part of the core survey data and therefore considered as control variables in the analysis [e.g., @meier_rehdanz10; @rehdanz07; @schmitz_madlener20]. This study, in contrast, relies on energy billing data at the building level, which can provide a more accurate picture of actual energy demand and prices than social surveys, but in return, does not provide socio-economic information, neither at the household nor the aggregated building-level. However, since previous studies provide sound arguments that the inclusion of socio-economic variables is relevant, an attempt is made to account for their relevance by using the most accurate data available at the district or postal code level. The approach here is that even if they cannot capture household-level effects, they might capture superordinate socio-economic effects at a higher level of aggregation.

The first of the three socio-economic variables included is the per capita disposable household income at the district level. In line with most of the literature, one would expect lower income to lead to lower heating demand as households have fewer funds at their disposal. However, it could also be that income at the district level is more of an approximation of a district's general affluence; effects such as more economic resources available for investment in building efficiency might outweigh the budget perspective, possibly leading to an opposite effect of district income. Secondly, retirement share is included as an additional socio-economic variable at the district level. Here, the intuition is that with high retirement rates in a district, more people spend more of their time at home, which may lead to a rise in space heat demand. Lastly, the population density within a postal code area is considered to observe if there may be structural differences in heating demand between densely populated metropolitan areas on the one hand and sparsely populated rural areas on the other.
