<!--
Chapter 5 - Results
-->

```{r setup5, echo=FALSE}
pacman::p_load(tidyverse, thesisdown, knitr, kableExtra, rio, gtsummary, gt)

# Set how wide the R output will go
options(width = 70)
```

# Results {#results}

## Descriptive statistics {#descriptives}

\begin{table}[]
\centering
\caption{Summary statistics}
\label{tab:summarytable}
\begin{tabular}{@{}llc@{}}
\toprule
\textbf{Variable}             & \textbf{Unit}            & \textbf{Median (IQR)}   \\ \midrule
Space heat consumption, eff.  & {[}kWh/sqm.{]}           & 113 (86, 145)           \\
Energy price, real            & {[}Cents/kWh{]}          & 6.49 (5.77, 7.69)       \\
                              &                          &                         \\
Degree days                   &                          & 3,446 (3,214, 3,733)    \\
Building heating surface      & {[}sqm{]}                & 404 (260, 707)          \\
Building housing units        &                          & 6 (3, 10)               \\
District household income     & {[}€/a{]}                & 20,695 (18,786, 22,568) \\
District retirement share     & {[}\%{]}                 & 0.207 (0.193, 0.220)    \\
District population density   & {[}inhabitants/sq. km{]} & 572 (217, 1,960)        \\
                              &                          &                         \\
\textit{Energy carrier group} & {\ul \textit{}}          &                         \\
Gas                           &                          & 1,647,563 (61\%)        \\
Oil                           &                          & 802,451 (30\%)          \\
District heating              &                          & 268,232 (9.9\%)         \\ \midrule
\textit{(N = 2,718,246; No missings)}  &                          & \multicolumn{1}{l}{}   
\end{tabular}
\end{table}



## Full sample analysis {#full_results}



Notes (copied from other papers):

-Accounting for unobserved heterogeneity reduces the elasticities by 15% to 32% relative to a model that does not include any effects at all, where the own price elasticity of electricity demand is −1.).


## Subsample analysis {#full_results}



```{r tabletest, echo=FALSE, results="asis"}
d <- import("data/flights.csv")

table_destination <- d %>% 
  group_by(dest_name) %>% 
  summarise(n = n())

table_destination %>% 
  kbl(col.names = c("Destination", "Number of connections"), 
      caption = "Number of flight connections per destination", 
      longtable = T, 
      booktabs = T,
      digits = 2) %>% 
  kable_styling(latex_options = "repeat_header",
                font_size = 10) %>%
  footnote(number = c("This table was created based on the flights dataset.", "Source: R Packages."))

# if too wide: 
# kable_styling(latex_options = "scale_down")




```


