<!--
Chapter 5 - Results
-->

```{r setup5, echo=FALSE}
pacman::p_load(tidyverse, thesisdown, knitr, kableExtra, rio, gtsummary, gt)

# Set how wide the R output will go
options(width = 70)
```

# Results {#results}

## Descriptive statistics {#descriptives}


```{r occurance-buildings, echo = FALSE, fig.cap = "", out.width = "65%", fig.align = "center"}
knitr::include_graphics(path = "figure/occurance_buildings.pdf")
```


Spatial and temporal distribution:

In Appendix \@ref(fig:buildings-distribution) the spatial and temporal distribution of observations in the processed full sample is shown on the district-level. On the temporal dimension, structurally fewer observations of around 40,000 are available in 2007 and 2008, as price data were first recorded in these years. In the decade between 2009 and 2019, a minimum of 201,856 and an average of 239,183 buildings are observed annually, with good spatial coverage. There are only a few districts where no buildings are observed within a year, and for most district-year combinations more than 10 and up to more than 10,000 buildings are observed.



\begin{table}[]
\centering
\caption{Summary statistics}
\label{tab:summarytable}
\begin{tabular}{@{}llc@{}}
\toprule
\textbf{Variable}             & \textbf{Unit}            & \textbf{Median (IQR)}   \\ \midrule
Space heat consumption, eff.  & {[}kWh/sqm.{]}           & 113 (86, 145)           \\
Energy price, real            & {[}Cents/kWh{]}          & 6.49 (5.77, 7.69)       \\
                              &                          &                         \\
Degree days                   &                          & 3,446 (3,214, 3,733)    \\
Building heating surface      & {[}sqm{]}                & 404 (260, 707)          \\
Building housing units        &                          & 6 (3, 10)               \\
District household income     & {[}€/a{]}                & 20,695 (18,786, 22,568) \\
District retirement share     & {[}\%{]}                 & 0.207 (0.193, 0.220)    \\
District population density   & {[}inhabitants/sq. km{]} & 572 (217, 1,960)        \\
                              &                          &                         \\
\textit{Energy carrier group} & {\ul \textit{}}          &                         \\
Gas                           &                          & 1,647,563 (61\%)        \\
Oil                           &                          & 802,451 (30\%)          \\
District heating              &                          & 268,232 (9.9\%)         \\ \midrule
\textit{(N = 2,718,246; No missings)}  &                          & \multicolumn{1}{l}{}   
\end{tabular}
\end{table}


```{r demand-descriptive-graph, echo = FALSE, fig.cap = "Distribution of energy demand", out.width = "100%", fig.align = "center"}
knitr::include_graphics(path = "figure/demand_descriptive.pdf")
```


```{r price-descriptive-graph, echo = FALSE, fig.cap = "Distribution of nominal and real energy prices", out.width = "100%", fig.align = "center"}
knitr::include_graphics(path = "figure/prices_descriptive.pdf")
```


## Full sample analysis {#full_results}


\begin{table}[]
\centering
\caption{Regression table for full sample analysis}
\label{tab:reg-table-full-sample}
\resizebox{\textwidth}{!}{%
\begin{tabular}{llrrrrrr}
\cline{3-8}
 &  & \multicolumn{6}{c}{Response variable in all model specifications: Ln of space heat demand} \\
 &  & \multicolumn{2}{c}{OLS regression} & \multicolumn{1}{c}{} & \multicolumn{3}{c}{Fixed Effects (FE) regression} \\
\multicolumn{1}{c}{} & \multicolumn{1}{c}{} & \multicolumn{1}{c}{(1)} & \multicolumn{1}{c}{(2)} & \multicolumn{1}{c}{} & \multicolumn{1}{c}{(3)} & \multicolumn{1}{c}{(4)} & \multicolumn{1}{c}{(5)} \\ \cline{1-1} \cline{3-4} \cline{6-8} 
(Intercept) &  & 5.399 *** & 3.593 *** &  &  &  &  \\
 &  & (0.002) & (0.028) &  &  &  &  \\
\begin{tabular}[c]{@{}l@{}}Ln of\\ energy price\end{tabular} &  & -0.365 *** & -0.365 *** &  & -0.251 *** & -0.243 *** & -0.243 *** \\
 &  & (0.001) & (0.001) &  & (0.001) & (0.001) & (0.001) \\
\begin{tabular}[c]{@{}l@{}}Ln of\\ degree days\end{tabular} &  &  & 0.626 *** &  &  & 0.779 *** & 0.779 *** \\
 &  &  & (0.002) &  &  & (0.003) & (0.003) \\
\begin{tabular}[c]{@{}l@{}}Ln of\\ heating surface\end{tabular} &  &  & -0.133 *** &  &  & -0.405 *** & -0.405 *** \\
 &  &  & (0.000) &  &  & (0.003) & (0.003) \\
\begin{tabular}[c]{@{}l@{}}Energy carrier:\\ Oil\end{tabular} &  &  & 0.031 *** &  &  & 0.097 *** & 0.097 *** \\
 &  &  & (0.001) &  &  & (0.002) & (0.002) \\
\begin{tabular}[c]{@{}l@{}}Energy carrier:\\ District heating\end{tabular} &  &  & -0.061 *** &  &  & -0.020 *** & -0.020 *** \\
 &  &  & (0.001) &  &  & (0.002) & (0.002) \\
\begin{tabular}[c]{@{}l@{}}Ln of\\ district income\end{tabular} &  &  & -0.277 *** &  &  & 0.022 ** & 0.022 ** \\
 &  &  & (0.002) &  &  & (0.008) & (0.008) \\
\begin{tabular}[c]{@{}l@{}}Ln of district\\ population density\end{tabular} &  &  & 0.043 *** &  &  & 0,000 &  \\
 &  &  & (0.000) &  &  & (0.005) &  \\
\begin{tabular}[c]{@{}l@{}}Ln of\\ retirement share\end{tabular} &  &  & -0.084 *** &  &  & 0.435 *** & 0.435 *** \\
 &  &  & (0.010) &  &  & (0.030) & (0.030) \\ \cline{1-1} \cline{3-4} \cline{6-8} 
N &  & 2718246 & 2718246 &  & 2718246 & 2718246 & 2718246 \\
R2 &  & 0,049 & 0,156 &  & 0,822 & 0,829 & 0,829 \\
logLik &  & -1322263,852 & -1159565,133 &  &  &  &  \\
AIC &  & 2644533,703 & 2319150,267 &  &  &  &  \\ \cline{1-1} \cline{3-4} \cline{6-8} 
\multicolumn{8}{l}{*** p \textless 0.001;  ** p \textless 0.01;  * p \textless 0.05.}
\end{tabular}%
}
\end{table}





Notes (copied from other papers):

-Accounting for unobserved heterogeneity reduces the elasticities by 15% to 32% relative to a model that does not include any effects at all, where the own price elasticity of electricity demand is −1.).


## Subsample analysis {#full_results}

```{r elasticity-predictions-subsample-graph, echo = FALSE, fig.cap = "Predicted price elasticities of demand by energy carrier", out.width = "100%", fig.align = "center"}
knitr::include_graphics(path = "figure/elasticity_predictions_subsample.pdf")
```




