<!--
Chapter 5 - Results
-->

```{r setup5, echo=FALSE}
pacman::p_load(tidyverse, thesisdown, knitr, kableExtra, rio, gtsummary, gt)

# Set how wide the R output will go
options(width = 70)
```

# Results {#results}

The results Chapter is structured in two parts. First, the results from the analysis of the full sample are presented (see Figure \@ref(fig:workflow1)). Subsequently, the results of the deepened analysis based on the stratified random subsample are reported (see Figure \@ref(fig:workflow2)).

## Full Sample Results {#full-sample-results}

**Price elasticity estimates**

The results of the full sample analysis are shown in Table \@ref(tab:reg-table). The overarching finding regarding the price elasticity of space heating demand is that the elasticities in all model specifications are *negative* and *relatively inelastic*. On the one hand, the price elasticities being negative confirms that the data reflect the basic intuition of declining demand with rising prices. On the other hand, the relative inelastic magnitude of the demand response validates the assumption derived from the literature that space heating in private households -- especially in the short-term -- is an inelastic good. 

Specifications (m1.ols) and (m2.ols) are the cross-sectional models based on the OLS estimator. Interestingly, the energy price coefficient remains stable when moving from Specification (m1.ols) with price as the only predictor of space heating demand to Specification (m2.ols), where the set of additional control variables is added to the model. The price elasticity coefficient in Specification (m2.ols) is to be interpreted that a 1% increase in energy price leads to a -0.37% reduction in space heating demand.


\begin{table}[]
\centering
\caption{Regression table for full sample analysis}
\label{tab:reg-table}
\resizebox{\textwidth}{!}{%
\begin{tabular}{llrrrrrrr}
\cline{3-9}
 &  & \multicolumn{7}{c}{\begin{tabular}[c]{@{}c@{}}Response variable in all model specifications:\\ Ln of space heat demand\end{tabular}} \\
 &  & \multicolumn{2}{c}{OLS} & \multicolumn{1}{c}{} & \multicolumn{4}{c}{Fixed Effects (FE)} \\
 &  & \multicolumn{1}{c}{(m1.ols)} & \multicolumn{1}{c}{(m2.ols)} & \multicolumn{1}{c}{} & \multicolumn{1}{c}{(m1.fe)} & \multicolumn{1}{c}{(m2.fe)} & \multicolumn{1}{c}{(m3.fe)} & \multicolumn{1}{c}{(m3.fe, lagged)} \\ \cline{1-1} \cline{3-4} \cline{6-9} 
(Intercept) &  & 5.399 *** & 3.593 *** &  &  &  &  &  \\
 &  & (0.002) & (0.028) &  &  &  &  &  \\
\begin{tabular}[c]{@{}l@{}}Ln of\\ energy price\end{tabular} &  & -0.365 *** & -0.365 *** &  & -0.251 *** & -0.243 *** & -0.243 *** &  \\
 &  & (0.001) & (0.001) &  & (0.001) & (0.001) & (0.001) &  \\
\begin{tabular}[c]{@{}l@{}}Ln of lagged\\ energy price (t-1)\end{tabular} &  &  &  &  &  &  &  & -0.071 *** \\
 &  &  &  &  &  &  &  & (0.001) \\
\begin{tabular}[c]{@{}l@{}}Ln of\\ degree days\end{tabular} &  &  & 0.626 *** &  &  & 0.779 *** & 0.779 *** & 0.850 *** \\
 &  &  & (0.002) &  &  & (0.003) & (0.003) & (0.004) \\
\begin{tabular}[c]{@{}l@{}}Ln of \\ heating surface\end{tabular} &  &  & -0.133 *** &  &  & -0.405 *** & -0.405 *** & -0.408 *** \\
 &  &  & (0.000) &  &  & (0.003) & (0.003) & (0.004) \\
\begin{tabular}[c]{@{}l@{}}Energy carrier:\\ Oil\end{tabular} &  &  & 0.031 *** &  &  & 0.097 *** & 0.097 *** & 0.074 *** \\
 &  &  & (0.001) &  &  & (0.002) & (0.002) & (0.002) \\
\begin{tabular}[c]{@{}l@{}}Energy carrier:\\ District heating\end{tabular} &  &  & -0.061 *** &  &  & -0.020 *** & -0.020 *** & -0.067 *** \\
 &  &  & (0.001) &  &  & (0.002) & (0.002) & (0.003) \\
\begin{tabular}[c]{@{}l@{}}Ln of district\\ per capita household income\end{tabular} &  &  & -0.277 *** &  &  & 0.022 ** & 0.022 ** & 0,015 \\
 &  &  & (0.002) &  &  & (0.008) & (0.008) & (0.010) \\
District retirement share &  &  & -0.084 *** &  &  & 0.435 *** & 0.435 *** & 0.418 *** \\
 &  &  & (0.010) &  &  & (0.030) & (0.030) & (0.037) \\
\begin{tabular}[c]{@{}l@{}}Postal code\\ population density\end{tabular} &  &  & 0.043 *** &  &  & 0,000 &  &  \\
 &  &  & (0.000) &  &  & (0.005) &  &  \\ \cline{1-1} \cline{3-4} \cline{6-9} 
N &  & 2718246 & 2718246 &  & 2718246 & 2718246 & 2718246 & 2020628 \\
R2 &  & 0,049 & 0,156 &  & 0,822 & 0,829 & 0,829 & 0,847 \\
logLik &  & -1322263,852 & -1159565,133 &  &  &  &  &  \\
AIC &  & 2644533,703 & 2319150,267 & \multicolumn{1}{l}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l}{} \\ \cline{1-1} \cline{3-4} \cline{6-9} 
*** p \textless 0.001;  ** p \textless 0.01;  * p \textless 0.05. &  & \multicolumn{1}{l}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l}{}
\end{tabular}%
}
\end{table}


When moving from the OLS estimator to the fixed-effects (FE) estimator in Specifications (m1.fe) - (m3.fe), the estimated price elasticities of demand become more inelastic. This effect is in line with what, for example, also @miller_alberini16 found when moving from a cross-sectional model to using unit-level fixed effects. This is probably due to unobserved factors at the building level that are not fully reflected in the building characteristics controlled for and temporal effects. The large increase in r-squared, as a metric of the variance in the response parameter explained by the model, also supports this assumption. While r-squared is at about 15.6% in Specification (m2.ols), it increases to 82,2% when moving to Specification (m1.fe). The price coefficient becomes slightly more inelastic when moving from Specification (m1.fe) (price is the sole predictor of space heating demand) to Specification (m2.fe) (includes additional explanatory variables). But the change remains small. Specification (m3.fe) drops the population density variable from the equation since in Specification (m2.fe), it is found to have no effect. Specification (m3.fe) reflects the final model for the full sample analysis. The price elasticity coefficient in Specification (m3.fe) is interpreted as a 1% increase in energy price leads to a -0.24% reduction in space heating demand.

Specification (m3.fe, lagged), also shown in the table, shows the results for the case that not the energy price in the same period but the price in the previous period (communicated through the energy bill of the previous year) influences the demand decision of households. When the lagged energy price is used as a predictor, the magnitude of the demand response is reduced to one third of the effect size: from -0.24% (price in $t$) to -0.07% (price in $t-1$). So while the demand response is still negative, this represents a major reduction in the magnitude of the effect. In the literature, there are varying rationales on whether the price in the same or previous periods informs demand decisions for space heating. In this study, I follow the approach of the majority of the literature and assume that prices in the same period drive demand levels (see Chapter \@ref(methods)). However, the results when using the price from the previous billing period are reported to obtain a more comprehensive picture of the price elasticity estimates. Importantly, it should be noted that part of the difference between the estimates may also be because in Specification (m3.fe, lagged) about 0.7 million fewer observations are included in the estimation. That is because a building must occur in the sample in two consecutive periods (energy price in $t-1$ and energy demand in $t$) for an observation to be included. This exclusion criterion may lead to selection bias, where buildings with specific characteristics continuously appearing in the sample are more heavily represented.

**Other estimates and effects**

Besides the interpretation of the price elasticities, the signs of other model predictors also point in the expected direction (see rationales established in Chapter \@ref(conceptual-model)). As most of the predictors are also included in the model in the ln-form, one can as well interpret them in terms of percentage point change. Referring again to the main final Specification (m3.fe), this means that, for example, a rise in degree days (lower outside temperatures) of 1% would be associated with a higher energy demand of 0.779%. And a larger building heating surface of 1% would be associated with a lower energy demand of -0.405%. Furthermore, the results indicate that an oil heating system is associated with a higher heating energy demand than gas as the reference category and a district heating system with lower demand. While the results in the cross-sectional OLS model indicate that an increasing district per capita household income would be associated with decreasing energy demand, this effect disappears and becomes close to no effect when moving to the fixed-effects models. So one can assume that district per capita household income does not have a strong influence in explaining the variance in energy demand at all. For district retirement share, the sign of the coefficient as well switches when moving from the OLS to the fixed-effects model. The results in the final Specification (m3.fe) indicate that a higher retirement share is associated with higher energy demand.

In general, almost all model predictors are found to be statistically significant, with a p-value < 0.1%. However, this is not surprising considering that more than 2.7 million observations are included in the model and that the number of observations is inversely proportional to the standard error of a model [@gelman_stern06; @johnson99]. Therefore one should focus more on the magnitudes of the coefficients rather than on statistical significance levels.

## Subsample Results {#subsample-results}

**Price elasticity estimates**

Figure \@ref(fig:posterior-distributions) summarises the posterior distributions for the price elasticities of space heating demand in the Bayesian model Specifications (see Figure \@ref(fig:workflow2) to compare which model structure and predictors are used in the respective Specification). The complete model summaries -- besides the energy price coefficients -- are reported in Appendix \@ref(tab:brms-full-model-results).

Also, for the Bayesian models, the estimates for the price elasticity of space heating demand are in the negative and relatively inelastic range. In contrast to the results of the total sample, however, it is noticeable that the price elasticities found in the subsample are more elastic. This could be partly due to the alternative modelling approach. But probably, it is mainly the different composition of the sample that causes the difference in the estimates. The overall trend of the results is similar in that the estimated price elasticities become more inelastic when the grouping terms for buildings and years are included.

Specifications (b.1) and (b.2) represent the simpler model design without the multilevel structure. In Specification (b.1), with energy price as the sole predictor of energy demand, the price elasticity is estimated to be -0.56 [-0.59; -0.53] (Mean estimate [95% CI]). When moving to Specification (b.2), where the additional explanatory variables are added, the estimate moves to -0.41 [-0.45; -0.38] and thus becomes more inelastic and wider. Again, using the results from Specification (b.2) as an example, the price elasticity coefficients are o be interpreted that a 1% increase in energy price leads to an average -0.41% reduction in space heating demand.

```{r posterior-distributions, echo = FALSE, fig.cap = "Posterior distributions for the price elasticities of space heating demand based on the subsample", out.width = "100%", fig.align = "center"}
knitr::include_graphics(path = "figure/posterior-distributions.pdf")
```

In the more sophisticated Specifications (bm.x) involving multilevel (partial pooling) with varying intercepts for buildings and years, the price elasticities become lower, similar to the fixed-effects models in the full sample analysis. In Specification (bm.1), energy price is the sole predictor, with the group-level terms included. In Specification (bm.2), the additional explanatory variables are added to the model. The inclusion of the heating surface and population density variables resulted in divergent transitions in the modelling process. Therefore, these two variables were excluded from the model to ensure model convergence and reliability of results. In Specification (bm.2), the estimate for the price elasticity of space heating demand is -0.30 [-0.33; -0.27], meaning that a 1% increase in energy price would lead to an average -0.30% reduction in space heating demand. Again, similar to the full sample analysis, the by far lowest estimate is found when using the lagged energy price variable ($t-1$) instead of the energy price in the same period ($t$) to explain energy demand levels. For Specification (bm.2 Lagged) the price elasticity estimate is -0.16 [-0.20; -0.12].

Generally, one should note that a direct comparison of the resulting estimates is not sensible due to the varying representation of energy carriers between the full- and the subsample. In the full sample, there is a higher representation of observations relying on a gas heating system and fewer observations with a district heating system. The stratified sampling approach removes these differences.

**Re-scaling elasticity results to the demand scale and integrating prediction intervals to convey uncertainty**

Figures \@ref(fig:elasticity-predictions-b1) and \@ref(fig:elasticity-predictions-bm2) illustrate what the differences found in the price elasticity estimates imply when being converted back to the scale of actual demand. Figure \@ref(fig:elasticity-predictions-b1) shows predictions for the basic Specification (b.1), where energy price acts as the sole predictor of space heating demand and group-level differences between buildings and years are not yet considered. Panel A on the left shows the observations in the subsample with model predictions on the ln-scale. Panel B on the right shows the re-scaled values on the scale of actual demand.[^1192]

[^1192]: Values are re-scaled using the $exp()$ link.

```{r elasticity-predictions-b1, echo = FALSE, fig.cap = "Predicted price elasticities of space heating demand based on simple model (Specification b.1)", out.width = "100%", fig.align = "center"}
knitr::include_graphics(path = "figure/b1_prediction.pdf")
```

Methodologically, the line in the centre represents the average energy demand at a respective price. The narrower, darker shaded interval around the mean represents the 95% intervals for the posterior distribution of the expected value of the demand parameter $\mu$ of the price elasticity (see Chapter \@ref(workflow-empirical-approach)). These posterior predictive distributions have a lower variance because only the uncertainty of the mean is included in the draws, but the residual error is ignored. On the other hand, the wider, lighter shaded interval represents the posterior predictive distributions which include the residual error. The lighter shaded interval is thus to be interpreted as the interval in which the model expects to find 95% of the actual energy demands in the population at each price.[^1193]

[^1193]: Cf. @mcelreath20, Ch. 4 and the brms-documentation on the $posterior.epred$ and $posterior_predict$ functions to interpret the posterior predictive intervals shown in the graph.

The two prediction intervals convey uncertainty in the parameter value (darker shaded interval) and uncertainty in the sampling process (lighter shaded interval). The distribution of simulated outcomes (lighter shaded interval) includes sampling variation and therefore indicates what the future data would look like [@mcelreath20]. In other words, it tells us what demand levels can be expected at different prices in a population of German multi-apartment buildings with equal shares of the three energy carriers gas, oil and district heating.[^1198]

[^1198]: Again, here, one should note that due to the stratified sampling procedure, the interpretations are only generalisable to a limited extent, as the relative shares of the energy carriers do not reflect the shares in the actual stock of rental buildings. The representations are more focused on representing the uncertainty in the models.

For Specification (b.1) in Figure \@ref(fig:elasticity-predictions-b1), the average space heating demand expected by the model drops from about 150 kWh/sqm/y at 4 cents/kWh to about 60 kWh/sqm/y at 20 cents/kWh.[^8282] In addition to the mean, the intervals show the uncertainty associated with the model predictions well. For the low end of the price range at 4 cents/kWh, the model expects 95% of the actual energy demand of the population to be between 70 and 315 kWh/sqm/y, which is an extensive range. At the upper end of the price range, the predicted interval narrows. At 20 cents/kWh, the model expects 95% of the actual energy demand to be between 35 and 130 kWh/sqm/y. Overall, the predictions of the simple model thus show the negative relationship between price and demand but are still subject to significant uncertainty due to the sizeable residual error of the model when only considering price as a sole predictor.

[^8282]: Note that the price range shown in the graph is arbitrarily chosen. It includes the lowest reasonable price taking into account historical market conditions at the lower end, and reflects about 2-3 times the observed prices during the period under study at the upper end.

```{r elasticity-predictions-bm2, echo = FALSE, fig.cap = "Predicted price elasticities of space heating demand based on multilevel partial pooling model (Specification bm.2)", out.width = "104%", fig.align = "center"}
knitr::include_graphics(path = "figure/bm2_prediction.pdf")
```

Figure \@ref(fig:elasticity-predictions-bm2) presents the same figure type but relies on the multilevel Specification (bm.2). The additional blue line in Panel A represents the simple linear relationship between price and demand. The different trajectory between the linear model (blue line) and the prediction for the model Specification (bm.2) (green line) shows the inclusion of the additional explanatory variables, and the group-level partial pooling with varying intercepts for buildings and years has a substantial impact on the prediction. The predicted elasticity is less elastic, and the intervals are considerably narrower. On the demand scale, again shown in Panel B, the average energy demand expected by the model drops from about 125 kWh/sqm/y at 4 cents/kWh to about 80 kWh/sqm/y at 20 cents/kWh. Moving to the simulated outcomes (lighter shaded intervals), the model expects 95% of the actual energy demand in the population to be between 85 and 175 kWh/sqm/y for the lower end of the price range shown at 4 cents/kWh. And at the higher end, at 20 cents/kWh, the interval narrows again to 55 to 110 kWh/sqm/y. Thus, the intervals have become much narrower, indicating that the prediction of Specification (bm.2) provides a more comprehensive picture of the interplay between energy price and energy demand than the prediction from the simple model in Specification (b.1).

Overall, the figures indicate what the reported price elasticities of space heating demand would mean in practice, given a range of realistic price levels. Furthermore, they also communicate the uncertainty inherent to the model estimates.

**Investigation of potential heterogeneity in price responsiveness**

While the results presented thus far for the subsample largely mirror the procedure used for the full sample analysis, but based on Bayesian inference and using a different sample composition, the final part of the subsample analysis aims to go further by examining potential sources of heterogeneity in the price responsiveness. This further analysis involves two steps. First, the variables used in the analysis are examined for potential heterogeneity in price responsiveness, relying on graphical representations. Second, the heterogeneity found for the different energy carrier groups is then formalised in regression analysis employing an interaction term between energy price and energy carrier groups.

For the first part, the visual examination of potential heterogeneity in price responsiveness, the following variables are considered: year, federal state, energy carrier group, degree days, heating surface, district per capita household income, and district retirement share.[^8283] The graphs for the categorical variables years and federal states are shown in Appendix \@ref(fig:heterogeneity-year-state-plot).[^7272] Between years, one can identify no significant differences. The lines all follow a similar slope. The same is true for the federal state variable. While some differences exist between the individual states, there is no pattern of strongly diverging price elasticities of space heating demand.

[^7272]: The observations in the subsample are grouped by years or federal states and presented in a scatter plot between energy price and demand. The lines represent simple linear models for each grouped category (year or federal state). 

[^8283]: Note that postal code population density was not considered because it was neither included in the final model for the analysis of the full sample (Specifications (m2.fe)) nor the main model for the analysis of the subsample (Specifications (bm.2)).

Similarly, Appendix \@ref(fig:other-variables-heterogeneity-plot) presents the visual examinations for the variables degree days, heating surface, district per capita household income, and district retirement share. Because all four variables are continuous, they are each divided into three equally sized groups to be able to investigate possible heterogeneity in the price response visually (e.g., one-third of observations with the highest number of degree days, one-third with the medium number of degree days, one-third of observations with the lowest number of degree days). For the two variables, degree days and district retirement share, the three grouped lines are almost entirely parallel, suggesting no differences in the price response between the groups. For district per capita household income, minor differences can be observed. Districts with a lower per capita household income exhibit a more elastic demand response (steeper slope). For the building heating surface, more pronounced differences are found, with larger buildings being associated with a more elastic demand response than smaller buildings.

```{r heterogeneity-energy-carrier-plot, echo = FALSE, fig.cap = "Scatter plot for the investigation of heterogeneity between energy source groups", out.width = "80%", fig.align = "center"}
knitr::include_graphics(path = "figure/carrier_heterogeneity_plot.pdf")
```

In addition, the three energy carrier groups  (oil, gas, and district heating) were investigated for potential heterogeneity. The results are shown in Figure \@ref(fig:heterogeneity-energy-carrier-plot). Among all the variables investigated, the differences in the price responsiveness between the energy carrier groups are most pronounced. For buildings with oil heating, demand is almost constant, which indicates that they do not adjust their energy demand to a large extent to a change in the oil price. Buildings with gas heating show a moderate demand response to price changes. And buildings with district heating have the most elastic demand response when relying on the simple linear relationship shown in the graph.

Based on the visual examination of the variables, the findings were used to construct a better fitting model which integrates heterogeneity in price responsiveness by adding an interaction term. Since the visual analysis revealed that there was no heterogeneity detected between varying years, federal states, degree days and district retirement shares, these variables were not examined further. The heterogeneity found between the energy carrier groups, the district per capita household income, and building heating surface variables was, in contrast, assessed further. Exploratory testing of models showed that interacting energy price with district per capita household income showed no relevant difference. Furthermore, interacting energy price with building heating surface led to non-convergence of model results and was therefore not reported. However, interacting energy price with the energy carrier group led to relevant differences in the price elasticity pared with a converging model.

The posterior distributions for the price elasticities of space heating demand for the interaction model Specification (bm.3) are shown in Figure \@ref(fig:posterior-distribution-interaction). The varying posterior distributions confirm the results of the visual examination. The most inelastic demand response is found for buildings with oil heating -0.16 [-0.23; -0.10] (Mean estimate [95%CI]). The estimates for buildings with gas heating are in a medium range of -0.35 [-0.40; -0.29]. Demand is most elastic for buildings with district heating, with estimates of -0.53 [-0.60; -0.46]. Too, the width of the distributions provides some insights. For gas and district heating, the posterior distribution is wider than for oil. This is likely due to the observations for gas and district heating to be more widely dispersed (see also the previous Figure \@ref(fig:heterogeneity-energy-carrier-plot)). In line with the earlier estimates of price elasticity, the coefficients are to be interpreted as a corresponding decrease in energy demand given a 1% increase in energy prices.

```{r posterior-distribution-interaction, echo = FALSE, fig.cap = "Posterior distributions for the price elasticities of space heating demand by energy carrier group based on the subsample", out.width = "100%", fig.align = "center"}
knitr::include_graphics(path = "figure/posterior-distribution-interaction.pdf")
```

As with the previous models, the results of the model Specification (bm.3) are shown with prediction intervals in Figure \@ref(fig:elasticity-predictions-energy-carrier). Again, the inner darker shaded intervals represent the uncertainty in the parameter value, and the lighter shaded intervals represent the uncertainty in the sampling procedure and thus include the residual error of the model [@mcelreath20]. In Panel A, the models are shown on the ln-scale together with the underlying energy price and demand observations. In Panel B, the values from Panel A are re-scaled to the demand scale.

```{r elasticity-predictions-energy-carrier, echo = FALSE, fig.cap = "Predicted price elasticities of space heating demand by energy carrier group (Specification bm.3)", out.width = "104%", fig.align = "center"}
knitr::include_graphics(path = "figure/elasticity_predictions_subsample.pdf")
```

Mirroring the varying price elasticities of demand for the three energy carriers, the demand curve for oil is least responsive to changes in the energy price, and the demand curve for district heating is most responsive. Interestingly, due to the exponential relationship, the prediction intervals for gas and district heating become more narrow with rising energy prices. In contrast, the oil prediction interval remains relatively wide for higher energy prices. For gas, one can interpret the results in the way that for the lower end of the shown price range at 4 Cents/kWh, the model expects to find 95% of the actual energy demand in the building stock using gas heating in the demand range between about 90 and 180 kWh/sqm/y. For the upper end of the price range at 20 Cents/kWh, the expected demand values are between about 50 and 105 kWh/sqm/y. For oil, the model predicts 95% of the actual energy demand to be between about 85 and 175 kWh/sqm/y for the lower end and between about 70 and 140 kWh/sqm/y for the upper end of the same price range. And lastly, for district heating, the model predicts 95% of the actual energy demand to be between about 95 and 190 kWh/sqm/y for the lower end and between about 40 and 85 kWh/sqm/y for the upper end of the price range.

**Model comparison**

In the previous parts of the Chapter, various model Specifications for the Bayesian subsample analysis are presented, all producing varying results for the price elasticity of space heating demand. Thus, it is crucial to investigate which model has the more likely fit to isolate the most reliable elasticity estimate. For this purpose, Bayesian regression analysis resorts to the principle of model selection, which assumes that many potential models could be generating our data and that, based on tools, one should investigate which model is doing this more likely than others [@vehtari_etal17]. To practically implement a model comparison, I rely on the loo-package, which uses efficient approximate leave-one-out cross-validation (LOO) for Bayesian models where the approximation uses Pareto smoothed importance sampling (PSIS) [@vehtari2018loo].


\begin{table}[]
\centering
\caption{Model comparison based on PSIS-LOO}
\label{tab:model-comparison}
\begin{tabular}{@{}lrr@{}}
\toprule
\textbf{Model Specification} & \multicolumn{1}{l}{\textbf{Difference in ELPD}} & \multicolumn{1}{l}{\textbf{SE of the Difference}} \\ \midrule
bm.3 & 0,00 & 0,00 \\
bm.2 & -59.1 & 14.2 \\
bm.1 & -155.6 & 21.5 \\
b.2 & -5102.6 & 110.7 \\
b.1 & -5763.1 & 116.1 \\ \bottomrule
\end{tabular}
\end{table}

\par

The results of the model comparison are shown in Table \@ref(tab:model-comparison). In addition, a visual comparison of model fit is provided in Appendix \@ref(fig:plot-model-comparison).[^2626] In Table \@ref(tab:model-comparison), the models are compared in terms of their difference in expected log pointwise predictive density (ELDP) using LOO as a method. The results reveal that the Specification (bm.3) has the highest ELDP and thus the best model fit. The fit of the other models is represented relative to the best fitting model. Since the standard error associated with the ELDP estimate for Specification (bm.3) (14.2) is much smaller than the ELDP difference compared to Specification (bm.2) (59.1) as the model with the second best fit, there is no ambiguity regarding Specification (bm.3) being the model with the best fit. Furthermore, the results in Table \@ref(tab:model-comparison) show that the model fit strongly increases when moving from the simple model Specifications (b.1 & b.2) to the multilevel models with varying intercepts (bm.x). Hence, the model results using the multilevel approach must be considered more reliable than the results from the simple Specifications. Furthermore, Specification (bm.3) having the best fit indicates that including the interaction term between energy price and energy carrier group increases the reliability of the model. And that the price elasticity estimate becomes more precise when differentiated by the energy carrier group.

[^2626]: Note that the model Specification (bm.2, lagged), which uses the price from the previous period and thus does not have the same number of observations, cannot be compared directly with the others. Therefore, it is not included in Table \@ref(tab:model-comparison). However, the visual representation of model fit in Appendix \@ref(fig:plot-model-comparison) relies on simulated density distributions that can be produced for each model individually. Specification (bm.2, lagged) is included there. The pattern of simulated density distributions indicates a similar model fit compared to the other Specifications with a multilevel approach (bm.x).
