<!--
Chapter 5 - Results
-->

```{r setup5, echo=FALSE}
pacman::p_load(tidyverse, thesisdown, knitr, kableExtra, rio, gtsummary, gt)

# Set how wide the R output will go
options(width = 70)
```

# Results {#results}


## Full sample analysis {#full_results}


\begin{table}[]
\centering
\caption{Regression table for full sample analysis}
\label{tab:reg-table-full-sample}
\resizebox{\textwidth}{!}{%
\begin{tabular}{llrrrrrr}
\cline{3-8}
 &  & \multicolumn{6}{c}{Response variable in all model specifications: Ln of space heat demand} \\
 &  & \multicolumn{2}{c}{OLS regression} & \multicolumn{1}{c}{} & \multicolumn{3}{c}{Fixed Effects (FE) regression} \\
\multicolumn{1}{c}{} & \multicolumn{1}{c}{} & \multicolumn{1}{c}{(1)} & \multicolumn{1}{c}{(2)} & \multicolumn{1}{c}{} & \multicolumn{1}{c}{(3)} & \multicolumn{1}{c}{(4)} & \multicolumn{1}{c}{(5)} \\ \cline{1-1} \cline{3-4} \cline{6-8} 
(Intercept) &  & 5.399 *** & 3.593 *** &  &  &  &  \\
 &  & (0.002) & (0.028) &  &  &  &  \\
\begin{tabular}[c]{@{}l@{}}Ln of\\ energy price\end{tabular} &  & -0.365 *** & -0.365 *** &  & -0.251 *** & -0.243 *** & -0.243 *** \\
 &  & (0.001) & (0.001) &  & (0.001) & (0.001) & (0.001) \\
\begin{tabular}[c]{@{}l@{}}Ln of\\ degree days\end{tabular} &  &  & 0.626 *** &  &  & 0.779 *** & 0.779 *** \\
 &  &  & (0.002) &  &  & (0.003) & (0.003) \\
\begin{tabular}[c]{@{}l@{}}Ln of\\ heating surface\end{tabular} &  &  & -0.133 *** &  &  & -0.405 *** & -0.405 *** \\
 &  &  & (0.000) &  &  & (0.003) & (0.003) \\
\begin{tabular}[c]{@{}l@{}}Energy carrier:\\ Oil\end{tabular} &  &  & 0.031 *** &  &  & 0.097 *** & 0.097 *** \\
 &  &  & (0.001) &  &  & (0.002) & (0.002) \\
\begin{tabular}[c]{@{}l@{}}Energy carrier:\\ District heating\end{tabular} &  &  & -0.061 *** &  &  & -0.020 *** & -0.020 *** \\
 &  &  & (0.001) &  &  & (0.002) & (0.002) \\
\begin{tabular}[c]{@{}l@{}}Ln of\\ district income\end{tabular} &  &  & -0.277 *** &  &  & 0.022 ** & 0.022 ** \\
 &  &  & (0.002) &  &  & (0.008) & (0.008) \\
\begin{tabular}[c]{@{}l@{}}Ln of district\\ population density\end{tabular} &  &  & 0.043 *** &  &  & 0,000 &  \\
 &  &  & (0.000) &  &  & (0.005) &  \\
\begin{tabular}[c]{@{}l@{}}Ln of\\ retirement share\end{tabular} &  &  & -0.084 *** &  &  & 0.435 *** & 0.435 *** \\
 &  &  & (0.010) &  &  & (0.030) & (0.030) \\ \cline{1-1} \cline{3-4} \cline{6-8} 
N &  & 2718246 & 2718246 &  & 2718246 & 2718246 & 2718246 \\
R2 &  & 0,049 & 0,156 &  & 0,822 & 0,829 & 0,829 \\
logLik &  & -1322263,852 & -1159565,133 &  &  &  &  \\
AIC &  & 2644533,703 & 2319150,267 &  &  &  &  \\ \cline{1-1} \cline{3-4} \cline{6-8} 
\multicolumn{8}{l}{*** p \textless 0.001;  ** p \textless 0.01;  * p \textless 0.05.}
\end{tabular}%
}
\end{table}



\begin{table}[]
\caption{Summary of coefficients for energy price in Bayesian models}
\label{tab:price-coefs-brms}
\resizebox{\textwidth}{!}{%
\begin{tabular}{@{}lllllllll@{}}
\toprule
Coef.                               & Model             & Estimate & Est.Error & l-95\% CI & u-95\% CI & Rhat & Bulk ESS & Tail ESS \\ \midrule
\multirow{5}{*}{Ln of energy price} & b.1               & -0.56    & 0.02      & -0.59     & -0.52     & 1    & 3165     & 2380     \\
                                    & b.2               & -0.50    & 0.02      & -0.54     & -0.46     & 1    & 3502     & 2521     \\
                                    & bm.1              & -0.30    & 0.02      & -0.34     & -0.25     & 1    & 1286     & 2114     \\
                                    & bm.2              & -0.25    & 0.02      & -0.29     & -0.20     & 1    & 2240     & 2742     \\
                                    & bm.2 - Renovation & -0.25    & 0.02      & -0.29     & -0.20     & 1    & 2184     & 2920     \\
Ln of energy price in t-1           & bm.2 - Price t-1  & -0.14    & 0.03      & -0.19     & -0.09     & 1    & 2326     & 2920     \\ \bottomrule
\end{tabular}%
}
\end{table}


\begin{table}[]
\centering
\caption{Model comparison based on PSIS-LOO}
\label{tab:model-comparison}
\begin{tabular}{@{}lrr@{}}
\toprule
     & \multicolumn{1}{l}{elpd\_diff} & \multicolumn{1}{l}{se\_diff} \\ \midrule
bm.3 & 0.0                            & 0.0                          \\
bm.2 & -14.7                          & 7.5                          \\
bm.1 & -90.3                          & 14.0                         \\
b.2  & -2856.7                        & 80.4                         \\
b.1  & -3100.5                        & 81.6                         \\ \bottomrule
\end{tabular}
\end{table}


Notes (copied from other papers):

-Accounting for unobserved heterogeneity reduces the elasticities by 15% to 32% relative to a model that does not include any effects at all, where the own price elasticity of electricity demand is −1.).


## Subsample analysis {#full_results}

```{r elasticity-predictions-subsample-graph, echo = FALSE, fig.cap = "Predicted price elasticities of demand by energy carrier", out.width = "100%", fig.align = "center"}
knitr::include_graphics(path = "figure/elasticity_predictions_subsample.pdf")
```




