<!--
Chapter 5 - Results
-->

```{r setup5, echo=FALSE}
pacman::p_load(tidyverse, thesisdown, knitr, kableExtra, rio, gtsummary, gt)

# Set how wide the R output will go
options(width = 70)
```

# Results {#results}

The results Chapter is structured in two subsections. First, the results from the analysis of the full sample are presented (reflects first part of workflow, see Figure \@ref(fig:workflow1)). And subsequently the results of the deepened analysis based on the stratified subsample are reported (reflects the second part of the workflow, see Figure \@ref(fig:workflow2)).

## Full sample results {#full_sample_results}

**Price elasticity estimates**

The results of the full sample analysis are shown in Table \@ref(tab:reg-table). The overarching finding regarding the price elasticity of space heating demand is that the elasticities in all model specifications are *negative* and *relatively inelastic*. On the one hand, the price elasticities being negative confirms that the basic intuition of falling demand with rising prices is reflected in the data. The relative inelastic magnitude of the demand response, on the other hand, validates the assumption derived from the literature that space heating in private households -- especially in the short-term -- is an inelastic good. 

Specifications (m1.ols) and (m2.ols) are the cross-sectional models based on the OLS estimator. Interestingly, the energy price coefficient remains stable when moving from Specification (m1.ols) with price as the only predictor of space heating demand to Specification (m2.ols) where the set of additional control variables is added to the model. The price elasticity coefficient in Specification (m2.ols) is to be interpreted that a 1% increase in energy price leads to a -0.365% reduction in space heating demand.


\begin{table}[]
\centering
\caption{Regression table for full sample analysis}
\label{tab:reg-table}
\resizebox{\textwidth}{!}{%
\begin{tabular}{llrrrrrrr}
\cline{3-9}
 &  & \multicolumn{7}{c}{\begin{tabular}[c]{@{}c@{}}Response variable in all model specifications:\\ Ln of space heat demand\end{tabular}} \\
 &  & \multicolumn{2}{c}{OLS} & \multicolumn{1}{c}{} & \multicolumn{4}{c}{Fixed Effects (FE)} \\
 &  & \multicolumn{1}{c}{(m1.ols)} & \multicolumn{1}{c}{(m2.ols)} & \multicolumn{1}{c}{} & \multicolumn{1}{c}{(m1.fe)} & \multicolumn{1}{c}{(m2.fe)} & \multicolumn{1}{c}{(m3.fe)} & \multicolumn{1}{c}{(m3.fe, lagged)} \\ \cline{1-1} \cline{3-4} \cline{6-9} 
(Intercept) &  & 5.399 *** & 3.593 *** &  &  &  &  &  \\
 &  & (0.002) & (0.028) &  &  &  &  &  \\
\begin{tabular}[c]{@{}l@{}}Ln of\\ energy price\end{tabular} &  & -0.365 *** & -0.365 *** &  & -0.251 *** & -0.243 *** & -0.243 *** &  \\
 &  & (0.001) & (0.001) &  & (0.001) & (0.001) & (0.001) &  \\
\begin{tabular}[c]{@{}l@{}}Ln of lagged\\ energy price (t-1)\end{tabular} &  &  &  &  &  &  &  & -0.071 *** \\
 &  &  &  &  &  &  &  & (0.001) \\
\begin{tabular}[c]{@{}l@{}}Ln of\\ degree days\end{tabular} &  &  & 0.626 *** &  &  & 0.779 *** & 0.779 *** & 0.850 *** \\
 &  &  & (0.002) &  &  & (0.003) & (0.003) & (0.004) \\
\begin{tabular}[c]{@{}l@{}}Ln of \\ heating surface\end{tabular} &  &  & -0.133 *** &  &  & -0.405 *** & -0.405 *** & -0.408 *** \\
 &  &  & (0.000) &  &  & (0.003) & (0.003) & (0.004) \\
\begin{tabular}[c]{@{}l@{}}Energy carrier:\\ Oil\end{tabular} &  &  & 0.031 *** &  &  & 0.097 *** & 0.097 *** & 0.074 *** \\
 &  &  & (0.001) &  &  & (0.002) & (0.002) & (0.002) \\
\begin{tabular}[c]{@{}l@{}}Energy carrier:\\ District heating\end{tabular} &  &  & -0.061 *** &  &  & -0.020 *** & -0.020 *** & -0.067 *** \\
 &  &  & (0.001) &  &  & (0.002) & (0.002) & (0.003) \\
\begin{tabular}[c]{@{}l@{}}Ln of district\\ per capita household income\end{tabular} &  &  & -0.277 *** &  &  & 0.022 ** & 0.022 ** & 0,015 \\
 &  &  & (0.002) &  &  & (0.008) & (0.008) & (0.010) \\
District retirement share &  &  & -0.084 *** &  &  & 0.435 *** & 0.435 *** & 0.418 *** \\
 &  &  & (0.010) &  &  & (0.030) & (0.030) & (0.037) \\
\begin{tabular}[c]{@{}l@{}}Postal code\\ population density\end{tabular} &  &  & 0.043 *** &  &  & 0,000 &  &  \\
 &  &  & (0.000) &  &  & (0.005) &  &  \\ \cline{1-1} \cline{3-4} \cline{6-9} 
N &  & 2718246 & 2718246 &  & 2718246 & 2718246 & 2718246 & 2020628 \\
R2 &  & 0,049 & 0,156 &  & 0,822 & 0,829 & 0,829 & 0,847 \\
logLik &  & -1322263,852 & -1159565,133 &  &  &  &  &  \\
AIC &  & 2644533,703 & 2319150,267 & \multicolumn{1}{l}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l}{} \\ \cline{1-1} \cline{3-4} \cline{6-9} 
*** p \textless 0.001;  ** p \textless 0.01;  * p \textless 0.05. &  & \multicolumn{1}{l}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l}{} & \multicolumn{1}{l}{}
\end{tabular}%
}
\end{table}


When moving from the OLS estimator to the fixed-effects (FE) estimator in Specifications (m1.fe) - (m3.fe), the estimated price elasticities of demand become more inelastic. This effect is in line with what, for example, also @miller_alberini16 find when moving from a cross-sectional model to the use of unit-level fixed effects. The reason for this is probably unobserved factors at the building level that are not fully reflected in the building characteristics controlled for, as well as temporal effects. The assumption is also supported by the change in r-squared as a metric of the variance in the response parameter explained by the model. While r-squared is at about 15.6% in Specification (m2.ols), it increases strongly to 82,2% when moving to Specification (m1.fe). For the price coefficient, the change from Specification (m1.fe), where price is the only predictor of space heating demand, to Specification (m2.fe), which also includes the other predictor variables, becomes slightly more inelastic, but the effect remains small. Specification (m3.fe) drops the population density from the equation since it is found to have no effect in Specification (m2.fe). Thus, Specification (m3.fe) reflects the final model for the full sample analysis. The price elasticity coefficient in Specification (m3.fe) is to be interpreted that a 1% increase in energy price leads to a -0.243% reduction in space heating demand. The results for the price elasticity of demand are of a comparable magnitude to previous evidence in the literature (see Figure \@ref(fig:literature-estimates-plot)).

Specification (m3.fe, lagged), also shown in the table, shows the results for the case that not the energy price in the same period, but the price in the previous period (communicated through the energy bill of the previous year) influences the demand decision of households. When the lagged energy price is used as a predictor, the magnitude of the demand response is reduced to one third of the effect size: from -0.243% when using the price in $t$ to -0.071% for the price in $t-1$ (contingent on a 1% increase in energy prices). While the demand response is still negative, this represents a major reduction in magnitude of the effect. In the literature, there are varying rationales on whether the price in the same or in previous periods informs demand decisions for space heat. In this study I follow the approach of @schmitz_madlener20 and assume that prices in the same period drive demand levels. However, the results for when using the price in previous billing period are reported to obtain a more comprehensive picture on the price elasticity estimates. Importantly, it should be noted that part of the difference between the estimates may also be due to the fact that in Specification (m3.fe, lagged) about 0.7 million fewer observations are included in the estimation, since a building must occur in the sample in two consecutive periods (energy price in $t-1$ and energy demand in $t$) for an observation to be included. This exclusion criterion may lead to some type of selection bias, where buildings with certain characteristics that continuously appear in the sample are more heavily represented.

**Other estimates and effects**

In general, almost all model predictors are found to be statistically significant with a p-value < 0.1%. However, this is not surprising given the very large number of more than 2.7 million observations [@gelman_stern06; @johnson99]. Besides the interpretation of the price elasticities, the signs of the other relevant model predictors also point in the expected direction (see rationales established in Chapter \@ref(conceptual-model)). As most of the predictors are also included in the model in the ln-form, they can as well be interpreted in terms of percentage point change.

Referring again to the main model Specification (m3.fe), this means that, for example, a rise in degree days (lower outside temperatures) of 1% would be associated with to a higher energy demand of 0.779% and a larger building heating surface of 1% would be associated with a lower energy demand of -0.405%. Furthermore, the results indicate that an oil heating system is associated with a higher heating energy demand than gas as the reference category and a district heating system with a lower demand. While the results in the cross-sectional OLS model indicate that an increasing district per capita household income would be associated with a decreasing energy demand, this effect disappears in the FE model and is even reversed. At the same time, the magnitude of the district per capita household income effect in Specification (m3.fe) is so small that one must assume that it does not have a strong influence in explaining the variance in energy demand at all. For district retirement share the sign of the coefficient as well switches when moving from the OLS to the FE model. The results in Specification (m3.fe) indicate that a higher district level retirement share is associated with higher energy demand.

## Subsample results {#subsample_results}

**Price elasticity estimates**

Figure \@ref(fig:posterior-distributions) summarizes the posterior distributions for the price elasticities of space heating demand in the Bayesian models (see Figure \@ref(fig:workflow2) to compare which model structure and predictor is used in the respective Specification). The complete model summaries -- going beyond the price coefficients -- are reported in Appendix \@ref(tab:brms-full-model-results). 

Also for the Bayesian models, the estimates for the price elasticity of space heating demand are in the negative and relatively inelastic range. In contrast to the results of the total sample, however, it is noticeable that the price elasticities found in the subsample are more elastic. This could be partly due to the alternative modelling approach. But probably it is mainly the different composition of the sample that causes the difference in the estimates. That the overall findings of the full and subsample analysis agree well is also evident from the fact that the trend towards lower estimates for the price elasticity is the same when additional predictors are included and when controlling for the group-level effects of buildings and time.

Specifications (b.1) and (b.2) represent the simpler model design without varying intercepts. In Specification (b.1), with energy price as the sole predictor of energy demand, the price elasticity is estimated to be -0.56 [-0.59; -0.53] (Mean estimate [95% CI]). When moving to Specification (b.2) where the additional explanatory variables are added, the estimate moves to -0.41 [-0.45; -0.38] and thus becomes wider and less price responsive (inelastic). Again, using the results from Specification (b.2) as an example, the price elasticity coefficients are o be interpreted that a 1% increase in energy price leads to a -0.41% [-0.45; -0.38] reduction in space heating demand.

```{r posterior-distributions, echo = FALSE, fig.cap = "Posterior distributions for the price elasticities of space heating demand based on the subsample", out.width = "100%", fig.align = "center"}
knitr::include_graphics(path = "figure/posterior-distributions.pdf")
```

In the more sophisticated specifications involving multilevel partial pooling for buildings and years (bm.x models), the price elasticities become lower, similar to the FE models in the full sample analysis. In Specification (bm.1) energy price is used as a sole predictor and group-level terms are included. In Specification (bm.2) the additional explanatory variables are included. The inclusion of the additional variables resulted in divergent transitions caused by the heating area and population density variables. Therefore, the two were excluded from the model to ensure convergence and reliability of results. In Specification (bm.2) the estimate for the price elasticity of space heating demand is -0.30 [-0.33; -0.27], meaning that a 1% increase in energy price would lead to a -0.30% [-0.33; -0.27] reduction in space heating demand. Also similarly to the full sample analysis, the by far lowest estimate is found when using the lagged energy price variable ($t-1$) instead of the energy price in the same period ($t$) to explain the variance in space heat demand. For Specification (bm.2 Lagged) the estimate is -0.16 [-0.20; -0.12].

Generally, it should be noted that due to the varying representation of energy carriers between the full- and the subsample, a direct comparison of the resulting estimates is not sensible. In the full sample there is a larger representation of observations relying on a gas heating system and fewer observations with a district heating system. These differences are removed by the stratified sampling approach.

**Re-scaling elasticity results to the demand scale and integrating prediction intervals to convey uncertainty**

Figures \@ref(fig:elasticity-predictions-b1) and \@ref(fig:elasticity-predictions-bm2) illustrate what the differences found in the price elasticity estimates imply when being converted back to the actual scale of demand. Figure \@ref(fig:elasticity-predictions-b1) shows predictions for the basic Specification (b.1) where energy price acts as the sole predictor of space heating demand and group-level differences between buildings and years are not yet considered. Panel A on the left shows the observations in the subsample with model predictions on the ln-scale. Panel B on the right shows the re-scaled values on the scale of actual demand.[^1192]

[^1192]: Values are re-scaled using $exp()$.

```{r elasticity-predictions-b1, echo = FALSE, fig.cap = "Predicted price elasticities of space heating demand based on simple model (b.1)", out.width = "100%", fig.align = "center"}
knitr::include_graphics(path = "figure/b1_prediction.pdf")
```

Methodologically, the line in the center represents the average energy demand at a respective price. The narrower, darker shaded interval around the mean represents the 95% intervals for the posterior distribution of the expected value of the demand parameter $\mu$ (see Chapter \@ref(empirical-model)). These posterior predictive distributions have a lower variance because only the uncertainty of the mean is included in the draws, but the residual error is ignored. The wider, lighter shaded interval, on the other hand, represents the posterior predictive distributions including the residual error. It is thus to be interpreted as the interval in which the model expects to find 95% of the actual energy demands in the population at each price.[^1193]

[^1193]: Cf. @mcelreath20, Chapter 4 and the brms documentation on the $posterior.epred$ and $posterior_predict$ functions in R for the interpretation of the posterior predictive intervals shown in the graph.

The two types of prediction intervals convey two types of uncertainty: uncertainty in the parameter value (darker shaded interval) and uncertainty in the sampling process (lighter shaded interval). The distribution of simulated outcomes (lighter shaded interval) includes sampling variation and therefore indicates what the future data would look like [@mcelreath20]. In other words, it tells us what demand levels can be expected at different prices in a population of German multi-apartment buildings with equal shares of the three energy carriers gas, oil and district heating. The lighter shaded intervals are therefore the most important intervals for interpretation. It should be noted, however, that due to the stratified sampling procedure used to create the subsample, the interpretations are only generalisable to a limited extent, as the relative shares of the energy carriers do not reflect the shares in the actual building stock.

For the simple Specification (b.1) in Figure \@ref(fig:elasticity-predictions-b1), the mean space heating demand expected by the model drops from about 150 kWh/sqm/year at a price of 4 cents/kWh to about 60 kWh/sqm/year at a price of 20 cents/kWh.[^8282] In addition to the mean, the intervals show well the uncertainty associated with the model predictions. For the low end of the chosen price range at 4 cents/kWh, the model expects 95% of the actual energy demand of the population to be between 70 and 315 kWh/sqm/year, which is obviously a very wide range. At the upper end of the price range, the predicted interval narrows. At 20 cents/kWh, the model expects 95% of the actual energy demand to be in the range between 35 and 130 kWh/sqm/year. Overall, the predictions of the simple model thus show the negative relationship between price and demand, but are still subject to large uncertainty due to the residual error of the model.

[^8282]: Note that the price range shown in the graph were chosen arbitrarily to some extent. They were chosen to represent the lowest reasonable price at the low end and to reflect about 2-3 times the historical prices observed in the sample at the high end.

```{r elasticity-predictions-bm2, echo = FALSE, fig.cap = "Predicted price elasticities of space heating demand based on multilevel partial pooling model (bm.2)", out.width = "104%", fig.align = "center"}
knitr::include_graphics(path = "figure/bm2_prediction.pdf")
```

Figure \@ref(fig:elasticity-predictions-bm2) presents the same type of figure, but relies on the multilevel partial pooling model from Specification (bm.2). The additional blue line in Panel A represents the simple linear model between price and demand. The different trajectory between the simple linear model (blue line) and the prediction for the model Specification (bm.2) (green line) shows that the inclusion of the additional explanatory variables and the group-level pooling for buildings and years has a strong impact on the prediction. The predicted elasticity is less elastic and the intervals are much narrower. On the demand scale, again shown in Panel B, the average energy demand predicted by the model drops from about 125 kWh/sqm/year at a price of 4 cents/kWh to about 80 kWh/sqm/year at a price of 20 cents/kWh. Moving to the simulated results (lighter shaded intervals), the model expects 95% of the actual energy demand in the population to be between 85 and 175 kWh/sqm/year for the lower end of the price range shown at 4 cents/kWh. At the higher end, at a price of 20 cents/kWh, the interval narrows again to 55 to 110 kWh/sqm/year. Thus, the intervals have become much narrower, indicating that the prediction of Specification (bm.2) provides a more realistic picture on the interplay between energy price and demand as compared to the prediction from the simple model in Specification (b.1).

Overall, the figures provide a good indication of what the reported price elasticities of space heating demand would mean in practice, given a range of realistic price levels.

**Investigation of potential heterogeneity in price responsiveness**

While the results presented thus far for the subsample largely mirror the procedure used for the full sample analysis, but based on Bayesian inference and using a different sample composition, the final part of the subsample analysis aims to go further by examining potential sources of heterogeneity in the price responsiveness. This is done in two steps. First, the variables used in the analysis are examined for potential heterogeneity relying on graphical representations. Second, the heterogeneity found visually for the different energy carrier groups is then being formalized in a regression analysis by means of extending Specification (bm.2) by a interaction term between energy price and energy carrier groups.

For the first part, the visual examination of potential heterogeneity in price responsiveness, the following variables are considered: year, federal state, energy carrier group, degree days, heating surface, district per capita household income, and district retirement share.[^8283] For the categorical variables years and federal states the graphs are shown in Appendix \@ref(fig:heterogeneity-year-state-plot). The observations in the subsample are grouped by year or federal state and presented in a scatter plot between energy price and demand. The lines represent simple linear models for each grouped category of year or federal state. For the years variable, the graph well indicates that the energy demand in the sample is lower overall in the later periods. With regard to potential heterogeneity in price elasticity between the years, however, no significant differences can be identified. The lines for all years follow more or less the same slope. The same is true for the federal state variable. While there are some differences between the individual states, there is no pattern of strongly diverging price elasticities of space heating demand.

[^8283]: Note that postal code population density was not considered because it was neither included in the final model for the analysis of the full sample (Specifications (m2.fe)) nor in the final model for the analysis of the subsample (Specifications (bm.2)).

Similarly, Appendix \@ref(fig:other-variables-heterogeneity-plot) presents the visual examinations for the variables degree days, heating surface, district per capita household income, and district retirement share. Since all four variables are continuous, they are each divided into three equally sized groups to be able to investigate possible heterogeneity in the price response visually (e.g., one third of observations with highest number of degree days, one third with medium number of degree days, one third of observations with lowest number of degree days). For the two variables degree days and district retirement share, the three grouped lines are almost completely parallel, suggesting that there are no differences in the price response between the groups. For district per capita household income minor differences can be observed. Districts with a lower per capita household income exhibit a more elastic demand response (steeper slope). For the building heating surface as well some differences can be found with larger buildings being associated with a more elastic demand response than smaller buildings.

```{r heterogeneity-energy-carrier-plot, echo = FALSE, fig.cap = "Investigation of heterogeneity between energy carrier groups", out.width = "80%", fig.align = "center"}
knitr::include_graphics(path = "figure/carrier_heterogeneity_plot.pdf")
```

In addition, the energy carrier group was investigated for potential differences between the three carrier types (oil, gas, district heating). The results are shown in Figure \@ref(fig:heterogeneity-energy-carrier-plot). The differences in the price responsiveness between the energy carrier groups are most pronounced. For buildings with oil heating, demand is almost constant, which indicates that they do not adjust their energy demand to a large extent to a change in the oil price. Buildings with gas heating show a moderate demand response to price changes. And buildings with district heating have by far the most elastic demand response when relying on the simple linear relationship shown in the graph.

Based on the visual examination of the variables, the findings were used to construct a better fitting model which integrates heterogeneity in price responsiveness by adding an interaction term. Since the visual analysis revealed that there were no heterogeneity detected between varying years, federal states, degree days and district retirement shares, these variables were not probed further. The heterogeneity found between the energy carrier groups but also among the district per capita household income and building heating surface variables were assessed further. Exploratory testing of models showed that interacting energy price with district per capita household income showed no relevant difference. Furthermore, interacting energy prices with building heating surface lead to non-convergence of model results and was therefore not further pursued. Interacting energy price with the energy carrier group, however, lead to relevant differences for the price elasticity pared with a converging model.

The posterior distributions for the price elasticities of space heating demand for the additional estimated interaction model Specification (bm.3) are shown in Figure \@ref(fig:posterior-distribution-interaction). The different posterior distributions reflect the results of the visual examination of the simple linear relationship. The most inelastic demand response is found for buildings with oil heating -0.16 [-0.23; -0.10] (Mean estimate [95%CI]). For buildings with gas as the energy carrier type, the estimates are in a medium range of -0.35 [-0.40; -0.29]. For buildings with district heating, demand is most elastic with estimates of -0.53 [-0.60; -0.46]. Also the width of the distributions provides some insights. For gas and district heating as energy carriers the posterior distribution is wider than for oil. This is likely due to the observations for gas and district heating to be more widely spread (see also the previous Figure \@ref(fig:heterogeneity-energy-carrier-plot)).  In line with the previous estimates of price elasticity, the coefficients are to be interpreted as a corresponding decrease in energy demand  given a 1% increase in energy prices.

```{r posterior-distribution-interaction, echo = FALSE, fig.cap = "Posterior distributions for the price elasticities of space heating demand by energy carrier group based on the subsample", out.width = "100%", fig.align = "center"}
knitr::include_graphics(path = "figure/posterior-distribution-interaction.pdf")
```

As with the previous models, the results of the model Specification (bm.3), where energy price and energy carrier group are interacted, are shown with prediction intervals in Figure \@ref(fig:elasticity-predictions-energy-carrier). Again, the inner darker shaded intervals represent the uncertainty in the parameter value and the lighter shaded intervals represent the uncertainty in the sampling procedure and thus include the residual error of the model [@mcelreath20].

```{r elasticity-predictions-energy-carrier, echo = FALSE, fig.cap = "Predicted price elasticities of demand by energy carrier group (bm.3)", out.width = "104%", fig.align = "center"}
knitr::include_graphics(path = "figure/elasticity_predictions_subsample.pdf")
```

In Panel A the models are shown on the ln-scale together with the underlying observations for energy price and demand. In Panel B the values from Panel A are re-scaled to the demand scale. Mirroring the varying price elasticities of demand for the three energy carriers, the demand curve for district heating is most responsive to rising energy prices and the demand curve for oil the least responsive. Interestingly, due to the exponential relationship, the prediction intervals for gas and district heating become more narrow with rising energy prices, while the prediction interval for oil remains relatively wide, also for higher energy prices. For gas, the results can be interpreted in the way that for the lower end of the shown price range at 4 cents/kWh, the model expects to find 95% of the actual energy demand in the building stock using gas heating in the demand range between about 90 and 180 kWh/sqm/year. For the upper end of the price range shown at 20 cents/kWh, the expected demand values are between about 50 and 105 kWh/sqm/year. For oil, the model expects 95% of the actual energy demand to be between about 85 and 175 kWh/sqm/year for the lower end and between about 70 and 140 kWh/sqm/year for the upper end of the indicated price range of 4 to 20 cents/kWh. Ans lastly, for district heating, the model expects 95% of the actual energy demand to be between about 95 and 190 kWh/sqm/year for the lower end and between about 40 and 85 kWh/sqm/year for the upper end of the indicated price range of 4 to 20 cents/kWh.

**Model comparison**

The set of model Specifications for the subsample presented in the previous parts all produce different results for the price elasticity of space heating demand. Thus, it is important to investigate model fit to (...)


\begin{table}[]
\centering
\caption{Model comparison based on PSIS-LOO}
\label{tab:model-comparison}
\begin{tabular}{@{}lrr@{}}
\toprule
     & \multicolumn{1}{l}{elpd\_diff} & \multicolumn{1}{l}{se\_diff} \\ \midrule
bm.3 & 0.0                            & 0.0                          \\
bm.2 & -14.7                          & 7.5                          \\
bm.1 & -90.3                          & 14.0                         \\
b.2  & -2856.7                        & 80.4                         \\
b.1  & -3100.5                        & 81.6                         \\ \bottomrule
\end{tabular}
\end{table}

