<!--
Chapter 6 - Discussion
-->

```{r setup6, echo=FALSE}
pacman::p_load(tidyverse, thesisdown, knitr, kableExtra)

# Set how wide the R output will go
options(width = 70)
```

# Discussion {#discussion}

## Discussion of Empirical Results {#discussion-results}

**Findings on the price elasticity of space heating demand**

The first research question which this thesis aimed to answer is: How does a change in the price of energy affect the level of space heating demand? Put differently, the fist research question aims at determining an estimate of the price elasticity of space heating demand.

The results presented previously confirm the assumption that the demand for space heating is a highly inelastic. In the final model Specification in the analysis based on the full sample, the price elasticity was estimated such that the demand for space heating would decrease by -0.243% if the energy price increased by 1% (see Chapter \@ref(full-sample-results)). This means that the overall price elasticity estimated from the full set of data is lower than estimates from previous studies focusing on Germany, most of which estimate an elasticity between -0.3 and -0.6 (see Chapter \@ref(review)). Interestingly, however, the estimated overall price elasticity mirrors the results by @auffhammer_rubin18, the only other previous study that also relies on energy billing data. This might suggest that the previous studies on Germany which are based on social survey data, and therefore involve no observation-specific information on energy prices, overstate actual demand responses. Furthermore, the fact that the price elasticity estimates observed in the different model Specifications moved downwards when the grouping variables for buildings and years were included either as fixed-effects in the full sample or with a partial pooling approach for the Bayesian regression analysis of the subsample is consistent with findings also made by others in the literature [e.g., @miller_alberini16]. This refers in particular to the results of the study by @rehdanz07, which finds very high elasticities compared to this thesis and also compared to other studies, but does not control for unobserved inter-individual differences between buildings and temporal dynamics over time.

Due to the very large sample size of more than 2.7 million observations in the full sample analysis and its inversely proportional relationship to the standard error of a model, the standard error of the price elasticity estimates is very small and does not necessarily reflect the true uncertainty and dispersion about the different combinations of price and demand observed in the data. The uncertainty and dispersion becomes more apparent when looking at the results for the stratified random subsample (see Chapter \@ref(subsample-results)). Even though the estimates for the price elasticity of demand are not directly comparable due to the different sample composition, the final model Specification (bm.2), which does not yet include the interaction term, clearly shows that the prediction for price elasticity as a single parameter is rather narrow, but that the prediction interval becomes significantly wider when the residual error term of a model is included (see Figure \@ref(fig:elasticity-predictions-bm2)).

In addition, it is also necessary to discuss the estimated elasticity that results when the price of the previous period, rather than the price of the same period, is used to determine the price elasticity of space heating demand. The results of the analysis of the full sample show that the estimated elasticity becomes even far more inelastic, dropping from -0.243 when using the same period price to -0.071 when using the previous period price. This means that if one assumes that the price from the previous year informs the demand decision of inhabitants, because they were made aware of the prices through the last energy bill, the demand response is considerably more limited. However, it remains questionable whether a direct comparison of the two elasticity estimates is appropriate. Due to the need to have two consecutive billing observations (energy price in $t-1$ and other variables in $t$), about 0.7 million observations are removed from the sample in the model with the price lag. In contrast to the dropped observations, buildings that appear consecutively in the sample may have special characteristics (e.g., only larger buildings). This may introduce a bias compared to the full sample used in the other model Specifications. Moreover, it remains questionable whether the information obtained from a previous year's bill is actually more informative for residents about energy prices than their monthly prepayments in the current billing period together with the reporting of energy prices in the media or via other information channels. Nevertheless, to incorporate the estimate, one could use it, for example, as a lower bound of the price responsiveness when doing sensitivity testing with more than one estimate.

In summary, the inelastic results found for the price elasticity of space heating demand indicate that households do adjust their demand to the incremental price changes that occurred during the period under investigation, but only to a rather limited extent. Transferring this finding to the effects of price-related policy instruments, such as the BEHG recently introduced in Germany for the building sector to price carbon externalities, means that the direct demand effect to be expected is relatively limited in the short-term. Importantly, for the case of renter households that are observed here, the short-term elasticity may also reflect the long-term elasticity under the current design of the BEHG. As the carbon price is currently only borne by renters, who generally have limited options to make structural changes to the thermal conditions of their housing and change the heating technology, no additional long-term effect can be expected. However, this aspect is discussed further later in this section (see Chapter \@ref(policy-context)).

In addition, it should be pointed out once again that the historical period examined in this thesis is characterized by a gradual change in energy prices. Accordingly, the estimates for the price elasticity of space heating demand can only be transferred to the current situation of energy and especially natural gas scarcity in Europe to a limited extent. At the same time, the effects of the current strong price increases for residential heating energy on the price elasticity are anything but clear. On the one hand, it can be assumed that the strong public attention to energy prices probably leads to a promotion of energy-efficient behavior, which in turn would result in a more elastic demand. At the same time, the strong price increases have a particularly strong impact on the budget of low- and middle-income households, which is an additional factor likely to contribute to a more elastic demand. On the other hand, the price increase is also so strong that households may not be able to reduce their demand further in order to maintain a minimum level of comfort, which in turn would have the opposite effect on the elasticity. Overall, the price elasticities of space heating demand presented in this thesis can provide an empirical starting point also for an analysis of the current crisis situation. However, the underlying structural difference in price development and the limited transferability that this may entail should be made clear.

**Heterogeneity between the energy carrier groups**

Furthermore, this thesis also intended to provide evidence on the additional research question whether there any potential factors for heterogeneity in the sample that may not be reflected in the estimation of an overall mean price elasticity of demand?

Through the analysis of the subsample multiple variables were investigated for potential heterogeneity in price responsiveness. It was found that the type of energy carrier used within a building represented a relevant dimension for varying price elasticities of demand. The interaction term model used to quantify this difference showed that buildings with an oil heating system are associated with a price elasticity of -0.16 [-0.10; -0.23] (mean estimate [95%CI]) and thus have the lowest price responsiveness. Buildings with a gas heating system exhibit a price elasticity of -0.35 [-0.29; -0.40], implying a medium price responsiveness, but higher than the elasticity found for the full sample. And buildings with a district heating system show a price elasticity of -0.53 [-0.46; -0.60] and thus represent the group with the highest price responsiveness. Since the model comparison showed that the model interacting energy price and energy carrier group had the best model fit, the elasticities should be considered as more precise than the overall elasticity, when information on the type of energy carrier is available.

The reasons for the different price elasticities of demand between the three energy source groups are not completely evident. However, there are some lines of argumentation that may serve as partial explanations for the observed differences. One possible reason for the very inelastic demand response for oil could be that market prices for oil were much more volatile than for the other two energy carrier groups during the study period. This volatility may have led people to view price volatility as more of a normal and familiar state of events and no longer adjusted their demand based on price movements. A second explanation could be that the oil prices observed in the sample are probably less correlated with the current market prices for oil, since oil, unlike gas and district heating, which are on-demand fuels, is usually stored in the building and the price for new supplies is usually mixed with the price for the remaining heating oil stock. These stock effects might obscure the demand reactions within buildings with an oil heating.

Another reason for the difference in price elasticities could be that buildings with a gas or district heating system usually have a fixed cost component (flat-rate basic charge) as part of their energy contract. Due to the aggregated form of the data on energy costs, which are not broken down into fixed cost component (flat-rate basic charge) and variable cost component (demand-based price per kWh), it could happen that buildings with a low energy demand also have a high energy price per kWh because a larger share of the fixed cost component is attributed to a unit of demand. As a result of this trend, the low demand levels could be incorrectly attributed to the role of energy price in a model, potentially distorting the picture. Although this effect may have some impact, it should be noted that the inclusion of the building ID grouping variable helps to mitigate the effect by determining whether low energy demands are common for a particular building.

If one relates the heterogeneity in the price responsiveness found for the energy carriers groups back to the literature, the results can at first glance be seen as contradictory to the findings of the meta-study by @labandeira_etal17, which was presented in detail in chapter \@ref(review). The meta-results indicate that the average short-term elasticities for gas and oil (district heating was not observed) are similar to each other and in the order of magnitude of about -0.2. Thus, the results of this thesis indicate for gas that the price response for the specific purpose of residential space heating demand is higher than for the total pool of gas energy uses observed by @labandeira_etal17, and vice versa for oil. However, since the scope of the underlying individual studies in the meta-study is naturally much larger, it can be assumed that the results of this thesis are more accurate for the specific purpose of space heating demand in the residential buildings sector, although less generalisable beyond this scope.

Besides the heterogeneity found for the energy carrier groups, it was established trough the analysis that there was little to no heterogeneity in the sample for the other variables considered (e.g., households in colder regions in Germany do not systematically deviate in their energy demand response to changing energy prices from households in warmer regions). 

**Effects of other determinants on space heating demand**

The third research question this thesis aimed was answer was: What other determinants do affect the level space heating demand and need to be considered so that their effects are not falsely attributed to energy prices? 

The empirical analysis considered a set of additional variables, most of which showed a relevant association with the demand for space heating. The most relevant effect on the model results was clearly related to the inclusion of unobserved inter-individual differences between the units (buildings) and also an intra-individual dynamic over time (years) either trough fixed effects in the full sample analysis or trough the partial pooling approach in the subsample analysis. The model comparison conducted for the subsample analysis showed how significantly the model fit increased when considering the two grouping variables.

Even though the additional consideration of the other variables in most specifications has only a minor impact on the coefficient of the energy price when already controlling for the building and the year (see e.g., Figure \@ref(fig:posterior-distributions), the results for some of the other relevant variables point in the expected directions. The results show that at lower outside temperatures (higher degree days) space heating demand increases to a relevant extent. It can also be seen that the space heating demand in larger buildings is systematically lower than in smaller buildings. The effects of the additionally considered socio-economic variables at district or postal code level, on the other hand, are not clear and lead to different results with different model specifications.

## Integration into the Policy Context {#policy-context}

In addition to the discussion of the empirical results, it is also of relevance to embed the price elasticity of space heating demand estimate into the current policy context. The focus is on the medium- to long-term challenges of decarbonising the residential buildings sector in order to achieve climate targets, and less on the current situation of energy and gas scarcity, which has already been briefly discussed above. The highly inelastic demand response to energy price changes observed in this work illustrates that carbon pricing in the building sector is a relevant tool, but one that needs to be embedded in a broader policy mix to make the transition to a low-carbon buildings sector socially desirable and politically viable [@braungardt_etal21]. The estimated low elasticity means that households are price takers when it comes to price-related policy instruments for space heating. Since the demand response is much more inelastic than a unit-elastic response, the budget that households have to allocate to meet their desired demand for space heating becomes larger when energy prices rise.

This is particularly problematic for rental households, as they are the sole bearers of the carbon price signal under the current design of the Fuel Emissions Trading Act (BEHG) (see Chapter \@ref(relevance) for more details on implementation). Owners of rental buildings, on the contrary, are not burdened by the carbon price in the current design of the instrument and therefore have no financial incentive to improve a building's energy quality [@braungardt_etal21]. Since the opportunities for renters to improve the energetic quality of their apartments is rather limited, it is to be expected that the long-term price elasticity does not deviate strongly from the short-term elasticity. The elasticity estimate presented here therefore implies that the budgets of rental households are not only more burdened by additional heating expenditures in the short-term, but also in the long-term, and that this effect is amplified by the rising carbon price trajectory. This particularly affects the household budgets of low- and middle-income households, as @kroger_etal22 show in their analysis of the impact of the current sharp increase in household heating fuel prices.

In order to make the impact of increasing household expenditure from space heating more understandable, it is useful to return to the representation of average demand curves shown earlier in this thesis (see Chapter \@ref(theory)). The following Figure \@ref(fig:budget-effect) shows the impact of the BEHG-induced carbon price of 55 euros per tonne of CO~2~e in 2025 on gas. The baseline chosen is the average gas demand in the full sample (116 kWh/sqm/y) and the price is the average gas price in the five-year period between 2016 and 2020 (6.73 Cents/kWh). Although current market prices for gas are evidently much higher, the graph is intended to serve as an exemplary illustration of the structural effects of the carbon price on household expenditures. For the price elasticity represented by the demand curve, the average estimate of -0.35 for gas from the interaction model, Specification (bm.3), is used.

```{r budget-effect, echo = FALSE, fig.cap = "Exemplary expenditure effect of inelastic gas demand through the BEHG carbon price in 2025", out.width = "100%", fig.align = "center"}
knitr::include_graphics(path = "figure/budget_effect_gas_example.pdf")
```

Under the baseline scenario households would incur expenditures of 7.81 Euros/sqm/y (represented by the rectangle whose corner point lies in the red point in the demand curve) and under the scenario of the BEHG induced price effect in addition to the baseline price households would pay 8.61 Euros/sqm/y (represented by the second rectangle going further to the right). 

It is clear that this effect on household expenditures needs to be addressed by the wider policy mix for the decabonisation of the buildings sector in order to make the transition to a low-carbon buildings sector socially desirable and politically viable. Firstly, it is of overarching importance that the carbon price signal is passed on, at least in part, to the owners of rental buildings in the near future. Doing so would ensure that the long-term price elasticities in the rental segment of the buildings sector would increase. A gradual sharing of the total carbon price between renters and owners depending on the energetic condition of a building, as proposed by the Federal Ministry for Housing, Urban Development and Building earlier this year, is an option where the carbon price signal is partially passed on to building owners [cf. @bmwsb22]. An alternative approach might be a warm-rent model, as it is implemented in Sweden, for example. In a warm rent model, a surcharge for heat supply is added to the cold rent based on the energy condition of a rental building. Under this system, the price signal of a carbon price lie entirely with rental property owners.

In addition to passing on the carbon price signal, other policy measures are also important. A recent study for @uba22 also concludes that an ambitious mix of instruments is needed to achieve the 2030 emission reduction targets based on a range of possible scenarios. From a regulatory perspective, additional instruments include, above all, the introduction of minimum energy performance standards (MEPS), as proposed by the EU Commission in November 2021 as part of the revision of the Energy Performance of Buildings Directive [cf. @ep22]. Furthermore, clear phase-out rules for fossil heating systems, especially gas, are an important lever for emissions in the building sector. Both bans on the installation of fossil heating systems and the MEPS could be implemented into national law in Germany by amending the Building Energy Act (GEG) [cf. @bmwk/bmwsb22]. In addition to regulatory measures, according to the scenarios, fiscal instruments to support the energy-efficient renovation of the building stock, as well as programs to promote renewable heating technologies, are also regarded as highly relevant [@uba22]. As well as the expansion of central heat generation via district heating including network planning and the supply-side decabonisation of electricity and district heat generation [@uba22]. The carbon price, in turn, strengthens the effects of some instruments in the policy mix such as the use of funding programs. In addition, it can be assumed that the carbon price signal will increase the level of ambition, for example in the case of energy-efficient renovations [@uba22].
