<!--
Chapter 6 - Discussion
-->

```{r setup6, echo=FALSE}
pacman::p_load(tidyverse, thesisdown, knitr, kableExtra)

# Set how wide the R output will go
options(width = 70)
```

# Discussion {#discussion}

## Discussion of Empirical Results {#discussion-results}

**Findings on the price elasticity of space heating demand**

The first research question that this thesis aims to answer is: How does a change in the price of energy affect the level of space heating demand? The previously presented results confirm the assumption that the demand for space heating is highly inelastic. In the final model Specification in the full sample analysis, the price elasticity was estimated such that the demand for space heating would decrease by -0.24% if the energy price increased by 1% (see Chapter \@ref(full-sample-results)). This implies that the overall price elasticity estimated from the full set of data is more inelastic than estimates from previous studies focusing on Germany, most of which estimate an elasticity between -0.3 and -0.6 (see Chapter \@ref(review)). Interestingly, however, the estimated overall price elasticity mirrors the results by @auffhammer_rubin18, the only previous study that also relies on energy billing data. This might suggest that the earlier studies with a spatial focus on Germany, which are all based on social survey data -- and therefore involve no observation-specific information on energy demand and prices -- overstate actual demand responses.

In addition, the fact that the price elasticity estimates observed in the different model specifications became more inelastic when the grouping variables for buildings and years were included either as fixed effects or with a multilevel approach to Bayesian regression analysis is consistent with findings from the literature [e.g. @miller_alberini16]. This consistency with the literature links back to the study by @rehdanz07, which also focuses on Germany and finds very high elasticity estimates compared to the other previous studies on Germany and also to this thesis. Since @rehdanz07 does not control for unobserved interindividual differences between buildings and temporal dynamics over time in her methodological approach, it stands to reason that the results from her study are overestimated. And that the true elasticities are more likely to be reflected by the results of the other studies with reference to Germany or also this thesis.

Due to the huge sample size of more than 2.7 million observations in the full sample analysis and its inversely proportional relationship to the standard error of a model, the standard error of the price elasticity estimates is small and does not necessarily reflect the true uncertainty and dispersion about the different combinations of price and demand observed in the data. The uncertainty and dispersion become more apparent when looking at the results for the stratified random subsample (see Chapter \@ref(subsample-results)). Even though the estimates for the price elasticity of demand are not directly comparable due to the different sample composition, the model Specification (bm.2), which does not yet include the interaction term, clearly shows that the prediction for price elasticity as a single parameter is rather narrow, but that the prediction interval becomes significantly wider when the residual error term of a model is included (see Figure \@ref(fig:elasticity-predictions-bm2)).

In addition, it is also necessary to discuss the estimated elasticity that results when the price of the previous period -- rather than the price of the same period -- is used to determine the price elasticity of space heating demand. The analysis of the full sample shows that the estimated elasticity becomes even far more inelastic, dropping from -0.243 when using the same period price to -0.071 when using the previous period price. This means that if one assumes that the previous year's price informs the inhabitants' demand decision because they were made aware of the prices through the last energy bill, the demand response is considerably more limited. However, it remains questionable whether a direct comparison of the two elasticity estimates is appropriate. Due to the need to have two consecutive billing observations (energy price in $t-1$ and other variables in $t$), about 0.7 million observations are removed from the sample in the model with the price lag. Buildings that appear consecutively in the sample may have special characteristics (e.g., larger buildings). This may introduce a bias compared to the sample used in the other model Specifications. Moreover, it remains questionable whether the information obtained from a previous year's bill is more informative for residents about energy prices than their monthly prepayments in the current billing period, together with the reporting of energy prices in the media or via other information channels. Nevertheless, if one wants to incorporate the estimate into an analysis, it may be used as a lower bound of the price responsiveness when doing sensitivity testing.

In summary, the inelastic results found for the price elasticity of space heating demand indicate that households adjust their demand to the incremental price changes that occurred during the period under investigation, but only to a limited extent. Transferring this finding to the effects of price-related policy instruments, such as the BEHG recently introduced in Germany for the building sector to price carbon externalities, means that the direct demand effect to be expected is relatively limited in the short-term. Importantly, for the case of renter households that are observed in this thesis, the short-term elasticity may also reflect the long-term elasticity under the current design of the BEHG. As the carbon price is currently only borne by renters, who have limited options to make structural changes to the thermal conditions of their housing or change the heating technology, only minimal additional long-term effects are to be expected. However, this aspect is discussed in more detail later in this section (see Chapter \@ref(policy-context)).

In addition, it should be pointed out once again that a gradual change in energy prices characterises the historical period examined in this thesis. Accordingly, one should be cautious when aiming to transfer the estimates for the price elasticity of space heating demand to the current situation of energy and natural gas scarcity in Europe. However, the estimates may still be of use because the effects of the recent substantial price increases for residential heating energy on the price elasticity estimate are anything but straightforward. On the one hand, one can assume that public attention to energy prices probably promotes energy-efficient behaviour, resulting in a more elastic demand. Furthermore, the price increases have a particularly high impact on the budget of low- and middle-income households, which is an additional factor likely to contribute to a more elastic demand. On the other hand, the price increase is so substantial that households may be unable to reduce their demand further to maintain a minimum level of comfort, which in turn would have the opposite effect on the elasticity. Due to these countervailing effects, it is unclear how the elasticity estimate in the current situation may change. Thus, overall, the price elasticities of space heating demand presented in this thesis may also provide an empirical starting point for an analysis of the current crisis situation. However, the underlying structural difference in price development should be made clear.

**Heterogeneity between the energy carrier groups**

Furthermore, this thesis also intends to provide evidence on the additional research question of whether there are any potential factors for heterogeneity in the sample that may not be reflected in the estimation of an overall mean price elasticity of demand?

The subsample analysis investigated multiple variables for potential heterogeneity in price responsiveness. It was found that the type of energy carrier used within a building represented a relevant dimension for varying price elasticities of demand. The interaction term model used to quantify this difference showed that buildings with an oil heating system are associated with a price elasticity of -0.16 [-0.10; -0.23] (mean estimate [95%CI]) and thus have the most inelastic demand. Buildings with a gas heating system exhibit a price elasticity of -0.35 [-0.29; -0.40], so about two times as elastic as the gas response and also higher than the elasticity found for the full sample. And buildings with a district heating system show a price elasticity of -0.53 [-0.46; -0.60] and thus show the most elastic demand. Since the model comparison showed that the model interacting energy price and energy carrier group had the best model fit, one should consider the elasticities as more precise than the overall elasticity when information on the type of energy carrier is available.

The reasons for the different price elasticities of demand between the three energy source groups are not entirely evident. However, some lines of argumentation may serve as partial explanations for the observed differences. One possible reason for the very inelastic demand response for oil could be that market prices for oil were much more volatile than for the other two energy carrier groups during the period under study. This volatility may have led people to view price changes to represent a familiar state of events leading them to no longer adjust their demand levels. A second explanation could be that the oil prices observed in the sample are probably less correlated with the current market prices for oil. This is because oil is stored in the building, unlike gas and district heating, which are demand-based fuels, and the price of new supplies is usually blended with the price of the remaining heating oil. These stock effects might obscure the demand reactions within buildings with oil heating.

Another reason for the difference in price elasticities could be that buildings with a gas or district heating system usually have a fixed cost component (flat-rate basic charge) as part of their energy contract. Due to the aggregated form of the data on energy costs, which are not broken down into fixed cost component (flat-rate basic charge) and variable cost component (demand-based price per kWh), it could happen that buildings with a low energy demand also have a high energy price per kWh because a larger share of the fixed cost component is attributed to a unit of demand. As a result, low demand levels for these buildings might be incorrectly attributed to the role of energy price in a model, potentially distorting the picture. Although this effect may have some impact, it should be noted that the inclusion of the building ID grouping variable helps to mitigate the effect by determining whether low energy demands are the baseline for a particular building or occurred in coincidence with variable changes -- such as rising energy prices.

If one relates the heterogeneity in the price responsiveness found for the energy carriers groups back to the literature, the results can at first glance be seen as contradictory to the findings of the meta-study by @labandeira_etal17, which was presented in detail in Chapter \@ref(review). The meta-results indicate that the average short-term elasticities for gas and oil (district heating was not observed) are similar to each other and in the order of magnitude of about -0.2. Thus, the results of this thesis indicate for oil that the demand response for the specific purpose of residential space heating demand is less elastic than for the total pool of oil energy applications observed by @labandeira_etal17, and vice versa for gas. However, since the scope of the underlying individual studies in the meta-study is naturally much wider, it may be assumed that the results of this thesis are more accurate for the specific purpose of space heating demand in the residential buildings sector, although therefore also not generalisable beyond this scope.

Besides the heterogeneity found for the energy carrier groups, it was established through the analysis that there was little to no heterogeneity in the sample for the other variables considered (e.g., translated to the example of the degree days variable this means that households in colder regions in Germany do not systematically deviate in their energy demand response to changing energy prices from households in warmer regions). 

**Effects of other determinants on space heating demand**

The third research question this thesis aims to answer is: What other determinants affect the level of space heating demand and need to be considered so that their effects are not falsely attributed to energy prices? 

The empirical analysis considered a set of additional variables, most of which showed a relevant association with the demand for space heating. The most relevant effect on the model results was related to the inclusion of unobserved inter-individual differences between the units (buildings) and also an intra-individual dynamic over time (years) either through fixed effects in the full sample analysis or through the multilevel approach in the subsample analysis. The model comparison for the subsample analysis showed how significantly the model fit increased when considering these two grouping variables.

Even though the additional consideration of the other variables in most specifications has only a minor impact on the coefficient of the energy price when already controlling for building and year (see e.g., Figure \@ref(fig:posterior-distributions), the results for some of the other relevant variables point in the expected directions. The results show that at lower outside temperatures (higher degree days), space heating demand increases to a relevant extent. One can also see that the space heating demand in larger buildings is systematically lower than in smaller buildings. On the other hand, the effects of all three additionally considered socio-economic variables at the district or postal code level are unclear and lead to dissimilar results with different model specifications.

## Integration into the Policy Context {#policy-context}

In addition to the discussion of the empirical results, it is also of relevance to embed the price elasticity of space heating demand estimate into the current policy context. Here, the focus is on the medium- to long-term challenges of decarbonising the residential buildings sector to achieve climate targets and less on the current situation of energy and gas scarcity, which has already been briefly discussed above. This part of the discussion thus connects to the reasoning for the relevance of this thesis (see Chapter \@ref(relevance)). 

The highly inelastic demand response to energy price changes observed in this work illustrates that carbon pricing in the building sector is a relevant tool but needs to be embedded in a broader policy mix to make the transition to a low-carbon buildings sector socially desirable and politically viable [@braungardt_etal21]. The estimated low elasticity means that households are price takers regarding price-related policy instruments for space heating. Since the demand response is much more inelastic than a unit-elastic response, the budget that households have to allocate to meet their desired demand for space heating becomes larger when energy prices rise.

This is particularly problematic for rental households, as they are the sole bearers of the carbon price signal under the current design of the Fuel Emissions Trading Act (BEHG) (see Chapter \@ref(relevance) for more details on the implementation). Owners of rental buildings, on the contrary, are not burdened by the carbon price in the current design of the instrument and therefore have no financial incentive to improve a building's energy quality [@braungardt_etal21]. Since the opportunities for renters to improve the energetic quality of their apartments are limited, it is to be expected that the long-term price elasticity does not deviate strongly from the short-term elasticity. This implies that the budgets of rental households would not only be more burdened by additional heating expenditures in the short-term, but also in the long-term. And that the rising carbon price trajectory amplifies this effect. Furthermore, rising prices for space hating disproportionally affect low- and middle-income households, as @kroger_etal22 show in their analysis of the impact of the current sharp increase in household heating fuel prices.

In order to make the impact of increasing household expenditure -- especially for renting households who have limited ability to adapt -- for space heating more understandable, it is useful to return to the representation of average demand curves shown earlier in this thesis (see Chapter \@ref(theory)). Gas as the predominant energy carrier type in the sample is used to exemplify the effect. The graph shows the impact of the BEHG-induced carbon price of 55 euros per tonne of CO~2~e in 2025. The baseline energy demand is the average gas demand in the full sample (116 kWh/sqm/y). And the baseline energy price is the average gas price in the five years between 2016 and 2020 (6.73 Cents/kWh). The demand curve's price elasticity is the average estimate of -0.35 for gas from the interaction model (Specification (bm.3)). Evidently, current market prices for gas are much higher. But the graph is mainly intended to serve as an exemplary illustration of the structural price effect of the BEHG on household expenditures.

```{r budget-effect, echo = FALSE, fig.cap = "Exemplary expenditure effect of inelastic gas demand through the BEHG carbon price in 2025", out.width = "100%", fig.align = "center"}
knitr::include_graphics(path = "figure/budget_effect_gas_example.pdf")
```

In the baseline scenario, households would incur expenditures of 7.81 Euros/sqm/y (represented by the rectangle whose corner point lies in the red dot in the demand curve). In the BEHG-induced price effect scenario in 2025 (added to the baseline price), households would pay 8.61 Euros/sqm/y (represented by the second rectangle going further to the right). This significant impact on household expenditures, already at the relatively low price of 55 euros per tonne of CO~2~e in 2025, indicates that the broader policy mix must moderate the carbon price effect to make the transition to a low-carbon buildings sector socially desirable and politically viable.

Firstly, it is important that the carbon price signal is passed on, at least in part, to the owners of rental buildings in the near future. Doing so would ensure that the long-term price elasticities in the rental segment of the buildings sector would increase. A gradual sharing of the total carbon price between renters and owners, depending on the energetic condition of a building, as proposed by the Federal Ministry for Housing, Urban Development and Building earlier this year, is an option where the carbon price signal is partially passed on to building owners [cf. @bmwsb22]. An alternative approach might be a warm-rent model, as it is, for example, implemented in Sweden. In a warm rent model, a surcharge for heat supply is added to the cold rent based on the energy condition of a rental building. Under this system, the price signal of a carbon price lies entirely with rental property owners [@agoraenergiewende20].

In addition to passing on the carbon price signal, other policy measures are also important. A recent study for @uba22 concludes that an ambitious mix of instruments is needed to achieve the 2030 emission reduction targets for the buildings sector, based on an analysis of a range of possible scenarios. As regulatory policies, supplementary instruments include, for example, the introduction of minimum energy performance standards (MEPS), as proposed by the EU Commission in November 2021 as part of the revision of the Energy Performance of Buildings Directive [cf. @ep22]. The introduction of MEPS would ensure that buildings with a very high energy demand would have to be renovated. As the effect on household expenditure in energy inefficient buildings is much higher than in efficient ones, one can assume that the introduction of MEPS could also make the introduction of an increasing carbon price path possible, as the most extreme effects on household expenditure would no longer occur. Furthermore, clear phase-out rules for fossil heating systems are an important lever for emissions in the building sector. The introduction of phase-out rules mainly prevents lock-in effects for fossil heating technologies. Assuming a rising carbon price for the buildings sector, fossil heating technologies installed at a later date would be exposed to higher carbon price effects for longer.

Besides regulatory measures, fiscal instruments to support the energy-efficient renovation of the building stock and programs to promote renewable heating technologies are highly relevant in the @uba22 scenarios. Fiscal instruments would mainly lead to energy-efficient investments becoming viable for building owners at lower marginal costs because subsidies support them. This might induce more renovation activities under lower carbon price signals.

In summary, the observed inelastic demand response for space heating has an impact on the political feasibility of increasing carbon prices in the buildings sector. This is particularly true for the rental segment of the sector, as renters are limited in their ability to adjust the energy conditions of their homes and may therefore suffer not only short-term but also long-term expenditure effects on the budget they have to spend on space heating. To mitigate these impacts, it is essential that the carbon price signal is passed on to rental building owners to encourage them to make long-term investments in energy efficiency, and that the transition is supported with a broader mix of policy instruments.
