<!--
Chapter 5 - Results
-->

```{r setup5, echo=FALSE}
pacman::p_load(tidyverse, thesisdown, knitr, kableExtra, rio, gtsummary, gt)

# Set how wide the R output will go
options(width = 70)
```

# Results {#results}

## Full sample analysis {#full_results}

### Descriptive statistics {#full_results_descriptive}

### Regression results {#full_results_regression}




##  Subsample analysis {#full_results}

### Descriptive statistics {#full_results_descriptive}

### Regression results {#full_results_regression}



```{r tabletest, echo=FALSE, results="asis"}
d <- import("data/flights.csv")

table_destination <- d %>% 
  group_by(dest_name) %>% 
  summarise(n = n())

table_destination %>% 
  kbl(col.names = c("Destination", "Number of connections"), 
      caption = "Number of flight connections per destination", 
      longtable = T, 
      booktabs = T,
      digits = 2) %>% 
  kable_styling(latex_options = "repeat_header",
                font_size = 10) %>%
  footnote(number = c("This table was created based on the flights dataset.", "Source: R Packages."))

# if too wide: 
# kable_styling(latex_options = "scale_down")




```

\captionsetup[table]{labelformat=empty,skip=1pt}
\begin{longtable}{lc}
\toprule
\textbf{Variable} & \textbf{N = 2,993,841} \\ 
\midrule
Space heat consumption, eff. [kWh/sqm.] &  \\ 
Mean (SD) & 119 (47) \\ 
Median [Range] & 113 [21, 521] \\ 
Energy price, real [Cents/kWh] &  \\ 
Mean (SD) & 7.01 (2.06) \\ 
Median [Range] & 6.56 [2.00, 99.50] \\ 
Degree days &  \\ 
Mean (SD) & 3,496 (386) \\ 
Median [Range] & 3,445 [2,284, 7,003] \\ 
(Missing) & 3,209 \\ 
Energy carrier category &  \\ 
District heating & 290,010 (9.7\%) \\ 
Gas & 1,806,199 (60\%) \\ 
Oil & 875,945 (29\%) \\ 
Others & 21,687 (0.7\%) \\ 
Building heating surface [sqm.] &  \\ 
Mean (SD) & 694 (1,050) \\ 
Median [Range] & 402 [31, 73,437] \\ 
Building housing units &  \\ 
Mean (SD) & 10 (16) \\ 
Median [Range] & 6 [2, 1,694] \\ 
Regional household income [€] &  \\ 
Mean (SD) & 20,732 (2,925) \\ 
Median [Range] & 20,568 [13,821, 42,555] \\ 
(Missing) & 3,270 \\ 
Regional share retirement [>65 yrs] &  \\ 
Mean (SD) & 0.209 (0.024) \\ 
Median [Range] & 0.207 [0.145, 0.327] \\ 
(Missing) & 6,777 \\ 
 \bottomrule
\end{longtable}
\begin{minipage}{\linewidth}
\emph{This data is simulated}\\ 
\end{minipage}
