<!--
Appendix
-->

`r if(knitr:::is_latex_output()) '\\appendix'`

`r if(!knitr:::is_latex_output()) '# (APPENDIX) Appendix {-}'` 

```{r setup8, echo=FALSE}
pacman::p_load(tidyverse, thesisdown, knitr, kableExtra)

# Set how wide the R output will go
options(width = 70)
```

# Supplementary Materials

\singlespacing
\newpage

```{r degree-days-distribution, echo = FALSE, fig.cap = "Spatial and temporal variation in degree days", out.width = "77%", fig.align = "center"}
knitr::include_graphics(path = "figure/distribution_degree_days.pdf")
```

\noindent
Figure \@ref(fig:degree-days-distribution) shows the spatial and temporal variation of degree days as an operationalisation of outdoor temperature conditions. Degree days data was taken from @iwu21 and are used as an additional determinant of space heat demand in the regression analysis. Data is shown on the postal code level. Higher annual degree days are associated with lower outdoor temperatures and vice versa. Lower outdoor temperatures are in turn likely to be associated with increased space heating demand. In terms of spatial variation, the maps show colder outdoor temperatures (higher degree-days) mainly in the higher altitude regions, but also in eastern Germany, which is further inland in the European continent. In terms of temporal variation, considerable differences can be observed between the years. Especially 2010, but also 2012 and 2013 were cooler than the average of the years observed. The years 2014, 2018 and 2019, on the other hand, were significantly warmer.

\newpage

```{r correlation-plot, echo = FALSE, fig.cap = "Matrix with Pearson’s correlation coefficients", out.width = "90%", fig.align = "center"}
knitr::include_graphics(path = "figure/correlation_matrix.pdf")
```
\noindent
The values in Figure \@ref(fig:correlation-plot) represent Pearson's correlation coefficients. Values closer to 1 indicate a stronger positive relationship, values closer to -1 indicate a stronger negative relationship. Values close to 0 indicate no relationship. None of the correlations observed between the variables exceed moderate values, ruling out possible multicollinearity issues. Energy price is negatively correlated with energy demand, indicating that the expected negative effect of price on demand appears to be reflected in the data. Note that all variables except the district retirement share, which is already a ratio, are ln-transformed and reflect the data transformation used in the regression analysis.

\newpage


\begin{table}[]
\centering
\caption{Descriptive statistics for age of buildings and heating systems}
\label{tab:age-building-heating-system}
\resizebox{\textwidth}{!}{%
\begin{tabular}{@{}lllll@{}}
\toprule
Variable &
  \begin{tabular}[c]{@{}l@{}}Overall\\ N = 2,718,246\end{tabular} &
  \begin{tabular}[c]{@{}l@{}}Gas\\ N = 1,647,563\end{tabular} &
  \begin{tabular}[c]{@{}l@{}}Oil\\ N = 802,451\end{tabular} &
  \begin{tabular}[c]{@{}l@{}}District heating\\ N = 268,232\end{tabular} \\ \midrule
                                      &                    &                 &               &               \\
{\ul Building construction year}      &                    &                 &               &               \\
Until 1918                            & 34,536 (9.5\%)     & 22,004 (10.0\%) & 6,622 (6.1\%) & 5,910 (17\%)  \\
1919-1948                             & 22,652 (6.2\%)     & 14,494 (6.6\%)  & 4,742 (4.3\%) & 3,416 (9.7\%) \\
1949-1978                             & 121,949 (33\%)     & 68,253 (31\%)   & 42,973 (39\%) & 10,723 (30\%) \\
1979-1995                             & 126,299 (35\%)     & 76,869 (35\%)   & 40,072 (37\%) & 9,358 (26\%)  \\
1996-2009                             & 58,582 (16\%)      & 38,409 (17\%)   & 14,411 (13\%) & 5,762 (16\%)  \\
After 2010                            & 1,172 (0.3\%)      & 738 (0.3\%)     & 288 (0.3\%)   & 146 (0.4\%)   \\
(Missing)                             & 2,353,056          & 1,426,796       & 693,343       & 232,917       \\
                                      &                    &                 &               &               \\
\multicolumn{2}{l}{{\ul Heating system installation year}} &                 &               &               \\
Until 1978                            & 11,486 (3.5\%)     & 6,360 (3.2\%)   & 4,296 (4.5\%) & 830 (2.6\%)   \\
1979-1995                             & 118,401 (36\%)     & 71,848 (36\%)   & 36,359 (38\%) & 10,194 (32\%) \\
1996-2009                             & 151,172 (46\%)     & 93,734 (47\%)   & 41,161 (43\%) & 16,277 (51\%) \\
After 2010                            & 46,556 (14\%)      & 27,564 (14\%)   & 14,532 (15\%) & 4,460 (14\%)  \\
(Missing)                             & 2,390,631          & 1,448,057       & 706,103       & 236,471       \\
                                      &                    &                 &               &               \\ \midrule
\textit{Note: n (\%)}                 &                    &                 &               &               \\ \bottomrule
\end{tabular}%
}
\end{table}

\noindent
Table \@ref(tab:age-building-heating-system) provides descriptive statistics for the age of buildings and heating systems grouped by the type of energy carrier. The information is taken from the Energy Performance Certificates (EPC) data which is otherwise not used for this thesis, as it is only available for a relatively small share of observations. For those buildings where the information is available, it can be observed from the table above that a relatively larger share of district heating systems are installed in older buildings but at a more recent date. Overall, however, the differences are moderate and likely only a minor determinant for the difference in energy demand between buildings with gas and oil heating on the one side (relatively higher demand) and district heating on the other side (relatively lower demand). Rather, it can be assumed that two other reasons explain most of the observed difference. Firstly, there is no additional loss of efficiency in local combustion in the boiler when using district heating, as is the case with oil and gas. And secondly, the observed buildings with district heating are on average significantly larger than the observed buildings with gas or oil heating (cf. Table \@ref(tab:summary-statistics). At the same time, it should be noted that the validity of the age classification of buildings and heating system installation year presented here is limited, as the information is available only for 15.5% and 13.7% of the total number of observations, respectively. This is also the reason why the two variables were not included as variables in the regression analysis.

\newpage


```{r density-distribution-comparison-samples, echo = FALSE, fig.cap = "Comparison of density distribution for energy demand and energy price between full sample and subsample", out.width = "100%", fig.align = "center"}
knitr::include_graphics(path = "figure/density-distribution-comparison-samples.pdf")
```

\noindent
Figure \@ref(fig:density-distribution-comparison-samples) compares the density distribution for the two key variables energy demand and energy price between the full sample (transparent) and the subsample (purple area) with 400 randomly selected buildings per energy carrier group (gas, oil and district heating). The visual comparison shows that there are only relatively small deviations in energy demand (Panel A), with the distribution in the subsample being slightly shifted to the left (lower demand). The comparison of distribution in energy prices (Panel B), on the other hand, show more divergence, with the subsample distribution showing a lower peak around $Ln(price) = 2$ and also more observations in the higher price range. However, the deviations in energy demand and energy prices can be well explained by the stratified sampling approach for the subsample. The stratified sample leads primarily to a relative reduction of gas observations and to a relative increase in district heating observations. For the demand variable, more district heating observations lead to more observations with a lower demand. For the price variable, it leads to a less pronounced peak in gas observations at the center (around $Ln(price) = 2$) and more district heating observations with higher prices at the upper end of the distribution.

\newpage

\tiny

\begin{longtable}[c]{lrrrrrrr}
\caption{Complete model summaries for Specifications in subsample analysis}
\label{tab:brms-full-model-results}\\
\cline{2-8}
 & \multicolumn{1}{c}{\textbf{Estimate}} & \multicolumn{1}{c}{\textbf{Est.Error}} & \multicolumn{1}{c}{\textbf{l-95\% CI}} & \multicolumn{1}{c}{\textbf{u-95\% CI}} & \multicolumn{1}{c}{\textbf{Rhat}} & \multicolumn{1}{c}{\textbf{Bulk ESS}} & \multicolumn{1}{c}{\textbf{Tail ESS}} \\ \cline{2-8} 
\endfirsthead
%
\endhead
%
\hline
\endfoot
%
\endlastfoot
%
 &  &  &  &  &  &  &  \\ \hline
\textbf{B.1} &  &  &  &  &  &  &  \\
{\ul \textit{Population-Level-Effects:}} &  &  &  &  &  &  &  \\
Intercept & 5.75 & 0.03 & 5.69 & 5.82 & 1.00 & 3240 & 2312 \\
ln(Price) & -0.56 & 0.01 & -0.59 & -0.53 & 1.00 & 3240 & 2432 \\ \hline
 &  &  &  &  &  &  &  \\ \hline
\textbf{B.2} &  &  &  &  &  &  &  \\
{\ul \textit{Population-Level-Effects:}} &  &  &  &  &  &  &  \\
Intercept & 4.93 & 0.49 & 3.94 & 5.89 & 1.00 & 3423 & 2799 \\
ln(Price) & -0.41 & 0.02 & -0.45 & -0.38 & 1.00 & 2740 & 2644 \\
ln(Degree.days) & 0.57 & 0.04 & 0.50 & 0.65 & 1.00 & 4404 & 2986 \\
ln(Heating.surface) & -0.15 & 0.01 & -0.16 & -0.14 & 1.00 & 3662 & 3067 \\
Carrier.group.oil & 0.04 & 0.01 & 0.02 & 0.06 & 1.00 & 3633 & 2650 \\
Carrier.group.district.heating & -0.06 & 0.01 & -0.09 & -0.03 & 1.00 & 2588 & 2890 \\
ln(District.income) & -0.35 & 0.03 & -0.41 & -0.29 & 1.00 & 4126 & 2746 \\
ln(Pop.density) & 0.05 & 0.00 & 0.04 & 0.05 & 1.00 & 3955 & 3125 \\
District.retire & -0.05 & 0.17 & -0.37 & 0.28 & 1.00 & 4222 & 3258 \\ \hline
 &  &  &  &  &  &  &  \\ \hline
\textbf{BM.1} &  &  &  &  &  &  &  \\
{\ul \textit{Population-Level-Effects:}} &  &  &  &  &  &  &  \\
Intercept & 5.31 & 0.04 & 5.23 & 5.39 & 1.00 & 1141 & 1800 \\
ln(Price) & -0.33 & 0.02 & -0.36 & -0.30 & 1.00 & 1565 & 1970 \\
 &  &  &  &  &  &  &  \\
{\ul \textit{Group-Level-Effects:}} &  &  &  &  &  &  &  \\
$\sim$id (Number of levels: 1200) &  &  &  &  &  &  &  \\
sd(Intercept) & 0.36 & 0.01 & 0.34 & 0.37 & 1.00 & 482 & 990 \\
 &  &  &  &  &  &  &  \\
$\sim$year (Number of levels: 13) &  &  &  &  &  &  &  \\
sd(Intercept) & 0.09 & 0.02 & 0.06 & 0.14 & 1.01 & 745 & 1844 \\ \hline
 &  &  &  &  &  &  &  \\ \hline
\textbf{BM.2} &  &  &  &  &  &  &  \\
{\ul \textit{Population-Level-Effects:}} &  &  &  &  &  &  &  \\
Intercept & 1.65 & 0.77 & 0.12 & 3.14 & 1.01 & 471 & 1247 \\
ln(Price) & -0.30 & 0.02 & -0.33 & -0.27 & 1.00 & 2318 & 2796 \\
ln(Degree.days) & 0.69 & 0.05 & 0.60 & 0.79 & 1.00 & 1378 & 1915 \\
Carrier.group.oil & 0.06 & 0.02 & 0.02 & 0.10 & 1.01 & 443 & 1067 \\
Carrier.group.district.heating & -0.12 & 0.02 & -0.16 & -0.08 & 1.02 & 401 & 930 \\
ln(District.income) & -0.20 & 0.06 & -0.32 & -0.08 & 1.01 & 364 & 934 \\
District.retire & -0.18 & 0.25 & -0.67 & 0.30 & 1.01 & 660 & 1176 \\
 &  &  &  &  &  &  &  \\
{\ul \textit{Group-Level-Effects:}} &  &  &  &  &  &  &  \\
$\sim$id (Number of levels: 1200) &  &  &  &  &  &  &  \\
sd(Intercept) & 0.35 & 0.01 & 0.34 & 0.37 & 1.01 & 461 & 855 \\
 &  &  &  &  &  &  &  \\
$\sim$year (Number of levels: 13) &  &  &  &  &  &  &  \\
sd(Intercept) & 0.03 & 0.01 & 0.02 & 0.05 & 1.00 & 1013 & 1516 \\ \hline
 &  &  &  &  &  &  &  \\ \hline
\textbf{BM.2 Lagged} &  &  &  &  &  &  &  \\
{\ul \textit{Population-Level-Effects:}} &  &  &  &  &  &  &  \\
Intercept & 0.88 & 0.92 & -0.97 & 2.70 & 1.01 & 532 & 975 \\
ln(Price, lagged t-1) & -0.16 & 0.02 & -0.20 & -0.12 & 1.00 & 2254 & 2452 \\
ln(Degree.days) & 0.72 & 0.06 & 0.60 & 0.83 & 1.01 & 1251 & 1975 \\
Carrier.group.oil & 0.04 & 0.02 & 0.00 & 0.09 & 1.01 & 364 & 617 \\
Carrier.group.district.heating & -0.20 & 0.02 & -0.25 & -0.16 & 1.00 & 460 & 903 \\
ln(District.income) & -0.17 & 0.07 & -0.31 & -0.03 & 1.01 & 455 & 963 \\
District.retire & -0.33 & 0.28 & -0.86 & 0.22 & 1.00 & 586 & 1219 \\
 &  &  &  &  &  &  &  \\
{\ul \textit{Group-Level-Effects:}} &  &  &  &  &  &  &  \\
$\sim$id (Number of levels: 1164) &  &  &  &  &  &  &  \\
sd(Intercept) & 0.36 & 0.01 & 0.34 & 0.37 & 1.00 & 506 & 866 \\
 &  &  &  &  &  &  &  \\
$\sim$year (Number of levels: 11) &  &  &  &  &  &  &  \\
sd(Intercept) & 0.03 & 0.01 & 0.02 & 0.05 & 1.00 & 1116 & 1753 \\ \hline
 &  &  &  &  &  &  &  \\ \hline
\textbf{BM.3} &  &  &  &  &  &  &  \\
{\ul \textit{Population-Level-Effects:}} &  &  &  &  &  &  &  \\
Intercept & 2.13 & 0.77 & 0.65 & 3.71 & 1.01 & 557 & 1248 \\
ln(Price) & -0.35 & 0.03 & -0.40 & -0.29 & 1.00 & 1309 & 2241 \\
ln(Degree.days) & 0.68 & 0.05 & 0.58 & 0.78 & 1.01 & 1106 & 1872 \\
Carrier.group.oil & -0.32 & 0.06 & -0.44 & -0.19 & 1.01 & 1175 & 2149 \\
Carrier.group.district.heating & 0.31 & 0.08 & 0.15 & 0.46 & 1.00 & 1139 & 1866 \\
ln(District.income) & -0.23 & 0.06 & -0.35 & -0.12 & 1.02 & 426 & 651 \\
District.retire & -0.18 & 0.27 & -0.71 & 0.34 & 1.01 & 370 & 924 \\
ln(Price):Carrier.group.oil & 0.19 & 0.03 & 0.13 & 0.26 & 1.00 & 1441 & 2276 \\
ln(Price):Carrier.group.district.heating & -0.18 & 0.04 & -0.25 & -0.11 & 1.00 & 1202 & 2009 \\
 &  &  &  &  &  &  &  \\
{\ul \textit{Group-Level-Effects:}} &  &  &  &  &  &  &  \\
$\sim$id (Number of levels: 1200) &  &  &  &  &  &  &  \\
sd(Intercept) & 0.35 & 0.01 & 0.33 & 0.36 & 1.01 & 386 & 687 \\
 &  &  &  &  &  &  &  \\
$\sim$year (Number of levels: 13) &  &  &  &  &  &  &  \\
sd(Intercept) & 0.03 & 0.01 & 0.01 & 0.04 & 1.00 & 912 & 1752 \\ \hline
\end{longtable}

\normalsize

\noindent
Table \@ref(tab:brms-full-model-results) presents the complete model summaries for the brms models in the subsample analysis. The parameters are summarized using the mean (estimate) and the estimated error of the posterior distribution with the 95% confidence intervals reported. Rhat is a measure of model convergence. Values below 1.05 indicate that chains mixed well and the model has converged. Bulk ESS and Tail ESS are diagnostics of the sampling efficiency.

\newpage

```{r heterogeneity-year-state-plot, echo = FALSE, fig.cap = "Scatter plot for the investigation of heterogeneity between years and federal states", out.width = "100%", fig.align = "center"}
knitr::include_graphics(path = "figure/year_state_heterogeneity_plot.pdf")
```

\noindent
Figure \@ref(fig:heterogeneity-year-state-plot) presents a visual analysis designed to detect potential heterogeneity for the price elasticity of space heating demand between years (Panel A) and federal states (Panel B). To construct the graphs, the observations in the subsample are grouped by year or federal state and presented in a scatter plot between energy price and demand. The lines in the diagrams reflect simple linear models for the grouped years or federal states. Between the years observed in the sample, all lines are almost parallel, indicating that there is no relevant difference in price elasticity. In the case of the federal states, the lines diverge a little. But no relevant pattern can be detected (e.g., differences between formerly eastern and western federal states or other regional clusters). Thus, no relevant heterogeneity in price responsiveness of detected for the two variables.

\newpage

```{r other-variables-heterogeneity-plot, echo = FALSE, fig.cap = "Scatter plot for the investigation of heterogeneity for the variables degree days, heating surface, district income, and district retirement share", out.width = "95%", fig.align = "center"}
knitr::include_graphics(path = "figure/other_variables_heterogeneity_plot.pdf")
```

\noindent
Figure \@ref(fig:other-variables-heterogeneity-plot) presents a visual analysis designed to detect potential heterogeneity for the price elasticity of space heating demand for the four variables degree days (Panel A), building heating surface (Panel B), average district income (Panel C), and district retirement share (Panel D). Since all four variables are continuous, the observations in the subsample are grouped into three equally sized groups[^99] and presented in a scatter plot between energy price and demand. The lines in the graphs reflect simple linear models for the respective three groups of equal size. For degree days and district retirement share, the lines are almost parallel and therefore indicate no difference in the price elasticity of demand. For district per capita household income minor differences can be observed. For building heating surface a more pronounced difference is observed with larger buildings being associated with a more elastic demand response than smaller buildings. The possible heterogeneity of the price responsiveness for district per capita income and heating surface is further investigated (see Chapter \@ref(subsample_results)), but ultimately discarded.

[^99]: Taking the example of degree days, this means that the third of the observations with the lowest degree days (between 7.8 and 8.1), the third of the observations with the medium degree days (between 8.1 and 8.2) and the third of the observations with the highest degree days (between 8.2 and 8.58) are combined as groups.

\newpage

```{r plot-model-comparison, echo = FALSE, fig.cap = "Graphical model comparison for Specifications in subsample analysis", out.width = "100%", fig.align = "center"}
knitr::include_graphics(path = "figure/plot-model-comparison.pdf")
```

\noindent
Figure \@ref(fig:plot-model-comparison) presents a visual examination of the model fit for all six models estimated for the subsample analysis. In contrast to the model comparison table presented in Chapter \@ref(subsample_results) here also model Specification (bm.2, Lagged) could be included. One can observe in the graphs, that the predictions (yellow lines) align better with the reference (purple line) once moving to the partial pooling models (bm.x). Among the partial polling models, the differences reported in Table \@ref(tab:model-comparison) in Chapter \@ref(subsample_results) are difficult to see visually. 

\newpage


