<!--
Appendix
-->

`r if(knitr:::is_latex_output()) '\\appendix'`

`r if(!knitr:::is_latex_output()) '# (APPENDIX) Appendix {-}'` 

```{r setup8, echo=FALSE}
pacman::p_load(tidyverse, thesisdown, knitr, kableExtra)

# Set how wide the R output will go
options(width = 70)
```

# Supplementary materials

\singlespacing
\newpage

```{r degree-days-distribution, echo = FALSE, fig.cap = "Spatial and temporal variation in degree days", out.width = "77%", fig.align = "center"}
knitr::include_graphics(path = "figure/distribution_degree_days.pdf")
```


\noindent
Figure \@ref(fig:degree-days-distribution) show the spatial and temporal variation of outside temperature measured in degree days. Degree days data was taken from @iwu21 and are used as an additional determinant of space heat demand in the regression analysis. Data is shown on the postal code level. Higher annual degree days are associated with lower temperatures and vice versa. In terms of spatial variation, the maps show cooler temperatures (higher degree days) mainly in the higher altitude regions, but also in eastern Germany, which is further inland from the European continent. In terms of temporal variation, 2010 in particular, but also 2012 and 2013 were cooler years than the average year in the sample. 2014, 2018 and 2019, on the other hand, were significantly warmer years.

\newpage

```{r correlation-plot, echo = FALSE, fig.cap = "Pearson’s correlation matrix", out.width = "90%", fig.align = "center"}
knitr::include_graphics(path = "figure/correlation_matrix.pdf")
```
\noindent
The values in Figure \@ref(fig:correlation-plot) represent Pearson's correlation coefficients. Values closer to 1 indicate a stronger positive relationship, values closer to -1 indicate a stronger negative relationship. Values close to zero indicate no relationship. None of the correlations observed between the variables exceed moderate values, ruling out possible multicollinearity issues. Energy price is negatively correlated with energy demand, indicating that the expected negative effect of price on demand indeed exists. Note that tall variables are ln-transformed.

\newpage


\begin{table}[]
\centering
\caption{Descriptive statistics for age of buildings and heating systems}
\label{tab:age-building-heating-system}
\resizebox{\textwidth}{!}{%
\begin{tabular}{@{}lllll@{}}
\toprule
Variable &
  \begin{tabular}[c]{@{}l@{}}Overall\\ N = 2,718,246\end{tabular} &
  \begin{tabular}[c]{@{}l@{}}Gas\\ N = 1,647,563\end{tabular} &
  \begin{tabular}[c]{@{}l@{}}Oil\\ N = 802,451\end{tabular} &
  \begin{tabular}[c]{@{}l@{}}District heating\\ N = 268,232\end{tabular} \\ \midrule
                                      &                    &                 &               &               \\
{\ul Building construction year}      &                    &                 &               &               \\
Until 1918                            & 34,536 (9.5\%)     & 22,004 (10.0\%) & 6,622 (6.1\%) & 5,910 (17\%)  \\
1919-1948                             & 22,652 (6.2\%)     & 14,494 (6.6\%)  & 4,742 (4.3\%) & 3,416 (9.7\%) \\
1949-1978                             & 121,949 (33\%)     & 68,253 (31\%)   & 42,973 (39\%) & 10,723 (30\%) \\
1979-1995                             & 126,299 (35\%)     & 76,869 (35\%)   & 40,072 (37\%) & 9,358 (26\%)  \\
1996-2009                             & 58,582 (16\%)      & 38,409 (17\%)   & 14,411 (13\%) & 5,762 (16\%)  \\
After 2010                            & 1,172 (0.3\%)      & 738 (0.3\%)     & 288 (0.3\%)   & 146 (0.4\%)   \\
(Missing)                             & 2,353,056          & 1,426,796       & 693,343       & 232,917       \\
                                      &                    &                 &               &               \\
\multicolumn{2}{l}{{\ul Heating system installation year}} &                 &               &               \\
Until 1978                            & 11,486 (3.5\%)     & 6,360 (3.2\%)   & 4,296 (4.5\%) & 830 (2.6\%)   \\
1979-1995                             & 118,401 (36\%)     & 71,848 (36\%)   & 36,359 (38\%) & 10,194 (32\%) \\
1996-2009                             & 151,172 (46\%)     & 93,734 (47\%)   & 41,161 (43\%) & 16,277 (51\%) \\
After 2010                            & 46,556 (14\%)      & 27,564 (14\%)   & 14,532 (15\%) & 4,460 (14\%)  \\
(Missing)                             & 2,390,631          & 1,448,057       & 706,103       & 236,471       \\
                                      &                    &                 &               &               \\ \midrule
\textit{Note: n (\%)}                 &                    &                 &               &               \\ \bottomrule
\end{tabular}%
}
\end{table}

\noindent
Table \@ref(tab:age-building-heating-system) provides descriptive statistics for the age of buildings and heating systems grouped by the type of energy carrier. For those buildings, where information is available, a relatively larger share of district heating systems are installed in older buildings but at a more recent date. Overall, however, the differences are moderate and cannot explain the difference in energy demand between buildings with gas and oil heating on the one side (relatively higher consumption) and district heating on the other side (relatively lower consumption). Instead, it is presumably the size of the buildings that induces to the lower demand levels for buildings with district heating (cf. Table \@ref(tab:summary-statistics). At the same time, it should be noted that the validity of the age classification of buildings and heating system installation year presented here is limited, as the information is available only for 15.5% and 13.7% of the total number of observations, respectively.

\newpage


\begin{table}[]
\centering
\caption{Comparison between full sample and sub-sample}
\label{tab:sample-comparison}
\resizebox{\textwidth}{!}{%
\begin{tabular}{@{}lllll@{}}
\toprule
Variable                                                                                       & Energy carrier group & \begin{tabular}[c]{@{}l@{}}Full sample\\ (N = 2.719.270)\end{tabular} & \begin{tabular}[c]{@{}l@{}}Sub-sample\\ (N = 4.410)\end{tabular} & Difference {[}\%{]}       \\ \midrule
\multirow{3}{*}{\begin{tabular}[c]{@{}l@{}}Energy demand\\ {[}kWh/m2{]}\end{tabular}}          & Gas                  & 122.5 {[}56.4; 209.33{]}                                              & 121.39 {[}55.4; 211.32{]}                                        & -0.91 {[}-1.81; 0.94{]}   \\
                                                                                               & Oil                  & 122.36 {[}58.67; 204{]}                                               & 116.97 {[}53.57; 198.17{]}                                       & -4.61 {[}-9.52; -2.94{]}  \\
                                                                                               & District heating     & 89.38 {[}43.23; 156.88{]}                                             & 83.98 {[}41.92; 141.01{]}                                        & -6.43 {[}-3.12; -11.25{]} \\
\multicolumn{5}{l}{} \\                                                                                                 
\multirow{3}{*}{\begin{tabular}[c]{@{}l@{}}Energy price, real \\ {[}Cents/kWh{]}\end{tabular}}        & Gas                  & 6.28 {[}4.56; 8.14{]}                                                 & 6.22 {[}4.59; 8.04{]}                                            & -0.96 {[}0.65; -1.24{]}   \\
                                                                                               & Oil                  & 7.14 {[}5.1; 9.18{]}                                                  & 7.21 {[}5.2; 9.23{]}                                             & 0.97 {[}1.92; 0.54{]}     \\
                                                                                               & District heating     & 10.45 {[}6.86; 15.07{]}                                               & 10.76 {[}7.17; 16.15{]}                                          & 2.88 {[}4.32; 6.69{]}     \\
\multicolumn{5}{l}{} \\   
\multirow{3}{*}{Degree days}                                                                   & Gas                  & 3467.6 {[}2921.14; 4180.42{]}                                         & 3461.14 {[}2914.56; 4180.4{]}                                    & -0.19 {[}-0.23; 0{]}      \\
                                                                                               & Oil                  & 3568.13 {[}2971.7; 4269.19{]}                                         & 3582.48 {[}2946.19; 4318.42{]}                                   & 0.4 {[}-0.87; 1.14{]}     \\
                                                                                               & District heating     & 3441.1 {[}2924.64; 4137.61{]}                                         & 3447.97 {[}2948.92; 4180.11{]}                                   & 0.2 {[}0.82; 1.02{]}      \\
\multicolumn{5}{l}{} \\   
\multirow{3}{*}{\begin{tabular}[c]{@{}l@{}}Building heating \\ surface {[}sqm{]}\end{tabular}} & Gas                  & 648.84 {[}163; 1849.4{]}                                              & 727.47 {[}168.04; 1817{]}                                        & 10.81 {[}3; -1.78{]}      \\
                                                                                               & Oil                  & 452.82 {[}160; 1195.6{]}                                              & 489.55 {[}167.1; 1362.8{]}                                       & 7.5 {[}4.25; 12.27{]}     \\
                                                                                               & District heating     & 1714.09 {[}255; 5030.18{]}                                            & 2034.55 {[}297.52; 6968.98{]}                                    & 15.75 {[}14.29; 27.82{]}  \\ \midrule
\multicolumn{5}{l}{\textit{*) Mean {[}90\% CI{]}}}                                                                                                                                                                                                                                          
\end{tabular}%
}
\end{table}

\noindent
Table \@ref(tab:sample-comparison) compares the total sample and the sub-sample with respect to key variables. It contains the mean values for the two samples and the lower and upper limits of a 90% interval divided by the energy carrier. It also gives the relative difference between the total sample and the sub-sample in percent. The statistics show that in the sub-sample with the available epc information, the buildings are on average 7.5 % to 15.8 % larger, depending on the energy carrier. This deviation is presumably due to the fact that larger buildings are more often managed by larger property managers who decide to systematically exchange their information with ista and also obtain their energy performance certificates through this standardised channel. The larger building heating area is likely also the reason for the deviations between the total sample and the sub-sample in energy demand for oil (-4.6%) and district heating (-6.4%). However, the deviations are within reasonable limits. Energy price and degree days from both samples are in good agreement.

\newpage

```{r heterogeneity-year-state-plot, echo = FALSE, fig.cap = "Investigation of heterogeneity for years and federal states", out.width = "100%", fig.align = "center"}
knitr::include_graphics(path = "figure/year_state_heterogeneity_plot.pdf")
```
\noindent
Figure \@ref(fig:heterogeneity-year-state-plot) presents a visual examination of the heterogeneity of price elasticity between years (Panel A) and federal states (Panel B). For this purpose, the observations in the sub-sample are grouped by year/federal state and presented in a scatter plot. The lines in the diagrams reflect simple linear models for the years/federal states as groups. Between years, all lines are almost parallel, indicating that there is no relevant difference in price elasticity between years. For the federal states, the lines scatter a little. At the same time, however, no strong pattern can be discerned through clusters such as differences between formerly eastern and western federal states or through other regional clusters.

\newpage

```{r other-variables-heterogeneity-plot, echo = FALSE, fig.cap = "Investigation of heterogeneity for degree days, heating surface, district income, and district retirement share", out.width = "95%", fig.align = "center"}
knitr::include_graphics(path = "figure/other_variables_heterogeneity_plot.pdf")
```
\noindent
Figure \@ref(fig:other-variables-heterogeneity-plot) presents a visual examination of the heterogeneity of price elasticity for the four variables degree days (Panel A), building heating surface (Panel B), average district income (Panel C), and district retirement share (Panel D). Since all four variables represent continuous variables, observations in the sub-sample are ln-transformed, grouped into three equally sized groups,[^99] and presented in a scatter plot. The lines in the diagrams reflect simple linear models for the respective three groups of equal size. For degree days, the lines are almost parallel and show no difference in price elasticity. For the other three variables, a slight dispersion of the grouped lines can be observed. But again, there are no pronounced differences that would prompt a more detailed investigation.

[^99]: E.g., one third of observations with lowest number of degree days (between 7.85 and 8.1), one third of observations medium number of degree days (between 8.1 and 8.2), and one third of observations highest number of degree days (between 8.2 and 8.63).

\newpage

```{r elasticity-predictions-b1, echo = FALSE, fig.cap = "Predicted price elasticities of demand based on price as only predictor", out.width = "100%", fig.align = "center"}
knitr::include_graphics(path = "figure/plot_b1.pdf")
```
\noindent
Figure \@ref(fig:elasticity-predictions-b1) presents (...)

\newpage

```{r plot-model-comparison, echo = FALSE, fig.cap = "Graphical model comparison for subsample analysis", out.width = "100%", fig.align = "center"}
knitr::include_graphics(path = "figure/plot-model-comparison.pdf")
```
\noindent
Figure \@ref(fig:plot-model-comparison) presents a visual examination of the model fit for all models used on the subsample.

\newpage


